---
title: "The two-stage estimator"
author: "Dharmesh D. Bhuva"
package: spiDE
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{The two-stage estimator}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
set.seed(1)
```

# Overview

`fitSpiDE()` *corrects* a cell-level fit: the design has one row per cell,
and random effects and reference degrees of freedom repair the standard errors
that pseudo-replication breaks (`vignette("spiDE-model")`).
`twoStageSpiDE()` takes the other route. It never tests a cell-level
coefficient. It first reduces each patient to a small set of **niche
slopes** — one per (index cell type, niche cell type) pair — and then runs
the condition contrast on those patient-level numbers, where the sample size
*is* the number of patients and no correction is needed.

This vignette documents the estimator end to end: what each stage computes,
the equations with every term defined, why the design choices were made, and
when to prefer it. Its measured operating characteristics are in
`vignette("spiDE-calibration")`.

```{r load, message = FALSE}
library(spiDE)
```

# Why a second estimator

`random = "none"`, `"intercept"` and `"slope"` all fit the *same* cell-level
design and differ in which columns are ridge-penalised. The two-stage route
is not a fourth `random` mode — it never fits the niche design at all. It
operationalises the observation that drives the model vignette: `condition`
is assigned per patient, so patients are the experimental units. If the
quantity compared between conditions is "how strongly this gene tracks this
niche in this patient", that quantity can be estimated per patient *first*
and compared afterwards.

The division of labour between the stages is the estimator's key property.

- **Stage 1 needs to be reasonable, not perfect.** Each per-patient slope is
  an estimate with a variance attached, and stage 2 tests the condition
  contrast against the *empirical* spread of the patient slopes. A
  misspecified stage 1 inflates that spread and costs power; it does not
  invalidate the test. Validity comes from contrasting patients.
- **Stage 2 is an ordinary small-sample regression** — tens of numbers per
  triplet, one per patient — which is the setting `limma`'s moderated $t$
  was built for.

Two things follow, and both matter for when to use it. Because every
(patient, index type) subset is fit on its own, the estimator borrows no
strength across patients or cell types: a thin subset — a dozen niche columns
on a few dozen cells — gives a poorly conditioned fit whose Fisher variance
understates its uncertainty, so calibration is governed by cells per subset.
And because it is within-sample by construction, it is immune to the
between-patient composition association that the model vignette's nested
intercept exists to remove.

# The pipeline at a glance

$$
\underbrace{\text{SPE with niches}}_{\text{input}}
\;\longrightarrow\;
\underbrace{\hat\beta^{(s,t)}_{g},\ v^{(s,t)}_{g}}_{\substack{\text{stage 1: joint WLS}\\ \text{per (sample, index type)}}}
\;\longrightarrow\;
\underbrace{\hat\beta_{g,p},\ v_{g,p}}_{\substack{\text{pool cores}\\ \text{within patient}}}
\;\longrightarrow\;
\underbrace{\hat\tau^2}_{\substack{\text{DerSimonian–Laird,}\\ \text{median over genes}}}
\;\longrightarrow\;
\underbrace{t_g,\ p_g}_{\substack{\text{stage 2: weighted}\\ \text{moderated limma}}}
\;\longrightarrow\;
\text{BH over triplets}
$$

Throughout, $g$ indexes genes, $c$ cells, $s$ samples (tissue cores), $p$
patients (a patient may contribute several cores; with no `patient` column
each sample is its own patient), $t$ the **index** cell type whose expression
is modelled, and $t'$ the **niche** cell type whose local density is the
predictor. A *triplet* is one $(g, t, t')$ hypothesis. The niche covariates
$N_{c,t'}(\sigma)$ are the effective-niche columns of the model vignette, at
one bandwidth $\sigma$.

# Stage 1: one joint slope fit per (sample, index type)

Stage 1 asks, within one sample $s$ and one index type $t$ at a time: across
the cells of this type in this core, how does this gene's expression co-vary
with the local density of each surrounding cell type? The answer is a vector
of slopes with variances, from a **joint regression on the niche columns**.
The three `stage1` options share the design and differ in the model fitted on
it.

## The joint fit

Within the subset of cells of index type $t$ in sample $s$ (kept only if it
has at least `min.cells` cells), let $\tilde N$ be the niche columns
**centred within the subset** and with the index type's own column removed —
the same self-interaction rule the full design applies. All three paths share
the design $D = [\,\mathbf 1 \mid \tilde N\,]$; the intercept absorbs each
gene's level. For the negative binomial paths the slopes are maximum-likelihood
estimates with variances from the per-gene Fisher information,
$v^{(s,t)}_{g,t'} = [(D^\top W_g D)^{-1}]_{t't'}$ with
$W_g = \mathrm{diag}(\hat\mu_{c,g}/(1+\hat\psi_g\hat\mu_{c,g}))$; a gene whose
information matrix is singular in a thin subset drops out as `NA` rather than
failing the run. For `"ols"`, with unit weights, they are the least-squares
quantities

$$
\hat\beta^{(s,t)}_{g} = \big(D^\top D\big)^{-1} D^\top \varepsilon_g,
\qquad
v^{(s,t)}_{g,t'} = \hat\varphi_g \Big[\big(D^\top D\big)^{-1}\Big]_{t't'},
\qquad
\hat\varphi_g = \frac{\sum_c r_{c,g}^2}{n_{s,t} - p},
$$

with $r_{c,g}$ the residuals, $n_{s,t}$ the subset's cell count and $p$ the
number of design columns. Degenerate niche columns (no variance within the
subset) are dropped and report `NA`.

**Why joint, not one niche at a time.** Niche densities are spatially smooth
and therefore correlated with each other. A marginal slope per niche lets one
true association bleed into every correlated niche column, and triplet
attribution becomes unreliable. Estimated jointly, correlated columns compete
for the same variation, so a smooth trend loads onto the column that carries
it. A practical consequence: restricting `niche =` changes the *fit*, not just
the report — every remaining slope is re-estimated in the smaller design.

## The three working responses

- **`"ols"`** regresses library-size-scaled log counts,
  $\varepsilon_{c,g} = \log(1 + Y_{c,g}\,\bar L / L_c)$ with $L_c$ the cell's
  total count and $\bar L$ their mean, with unit weights. It needs no stored
  normalisation fit and no dispersion, so it is dependency-free and the
  cheapest; the toy examples use it. It is also the best-measured stage-1
  path on calibration and power (calibration vignette), and the recommended
  one.
- **`"spanorm"`** (the default, for historical reasons) reads a stored
  `SpaNorm::SpaNorm()` fit from `metadata(spe)$SpaNorm` and uses its
  library-size / batch components as a fixed **offset** in a fresh negative
  binomial GLM on the raw counts of each subset,
  $\log\mu_{c,g} = a_{0,g} + \tilde N_c\beta_g + O_{c,g}$, with the
  dispersion from a per-sample pooled fit (`pool.psi = TRUE`). The technical
  effect enters as an offset, coefficient fixed at one, so it cannot absorb
  depth-correlated biology; the normalisation's biology component is
  deliberately *not* used — it anchors the normalisation and is not
  successfully modelled biology. The former `epsilon` argument is deprecated
  and ignored.
- **`"nb"`** fits a fresh negative binomial GLM per subset with a *fitted*
  log-library covariate in place of the offset. Its cost is per gene per
  subset, so it suits small `index` / `niche` restrictions rather than
  panel-wide runs.

# Between the stages: pooling cores into patients

When a patient contributed several cores, the per-core slopes are pooled by
inverse-variance weighting:

$$
\hat\beta_{g,p}
 = \frac{\sum_{s \in p} \hat\beta^{(s)}_{g} / v^{(s)}_{g}}
        {\sum_{s \in p} 1 / v^{(s)}_{g}},
\qquad
v_{g,p} = \Big(\sum_{s \in p} 1 / v^{(s)}_{g}\Big)^{-1}.
$$

Fixed-effect pooling is correct at this level — the cores share their
patient's true slope and differ only by sampling noise — while genuine
patient-to-patient variation enters at the next step through $\tau^2$.
`condition` is checked to be constant within each patient, and so must any
`patient.covariates`.

# Stage 2: the patient-level contrast

For each (index, niche) pair, stage 2 holds a genes × patients matrix of
pooled slopes $\hat\beta_{g,p}$ with variances $v_{g,p}$ and a patient-level
design $x_p = (1,\ r_p,\ \text{patient covariates})^\top$. The working model
is a random-effects meta-regression:

$$
\hat\beta_{g,p} = x_p^\top \theta_g + b_{g,p} + e_{g,p},
\qquad
b_{g,p} \sim \mathcal N(0, \tau^2),
\qquad
e_{g,p} \sim \mathcal N(0, v_{g,p}),
$$

| symbol | meaning |
|---|---|
| $\theta_g$ | the patient-level coefficients; its condition entry is the **tested quantity**, the difference in mean niche slope between conditions |
| $b_{g,p}$ | genuine patient-to-patient variation in the slope, one variance $\tau^2$ per (index, niche) pair |
| $e_{g,p}$ | stage-1 estimation noise, whose variance $v_{g,p}$ is known |

## The between-patient variance

$\tau^2$ is estimated by the **DerSimonian–Laird** moment method per gene,
then pooled. For one gene, fit the weighted regression with weights
$w_p = 1/v_{g,p}$, take residuals $r_p$, and form

$$
Q_g = \sum_p w_p\, r_p^2,
\qquad
\hat\tau^2_g
 = \max\!\Bigg(0,\;
   \frac{Q_g - (S_g - q)}
        {\sum_p w_p - \mathrm{tr}\!\big[(X^\top W X)^{-1} X^\top W^2 X\big]}
   \Bigg),
$$

with $S_g$ the patients with a finite slope for the gene, $q$ the number of
design columns, $X$ the stacked $x_p^\top$ and $W = \mathrm{diag}(w_p)$. If
the slopes scattered only by their stage-1 noise, $Q_g$ would average
$S_g - q$; any excess is between-patient variation. Per-gene estimates are
noisy, so — as the mixed model shares $\tau^2$ across genes — the reported
value is the **median over genes**, one per (index, niche) pair, floored at
zero.

## The weighted, moderated contrast

Each triplet is tested with `limma`: `lmFit()` on the slope matrix with
observation weights

$$
u_{g,p} = \frac{1}{v_{g,p} + \hat\tau^2},
$$

then `eBayes(robust = TRUE)`, reporting the moderated $t$ and p-value of the
condition column. The weight is the estimator's second load-bearing choice.
When $\tau^2$ dominates, $u_{g,p} \to 1/\tau^2$ — near-equal weights, the
validity of an unweighted patient-level test, and a guard against the
pseudobulk pitfall of over-weighting cell-rich patients (a patient with ten
times the cells has a smaller $v$ but not a more informative draw of
$b_{g,p}$). When $v$ dominates, the weights down-weight patients whose slopes
are poorly estimated. `eBayes` moderation shrinks each gene's residual
variance toward a gene-shared prior, lifting the reference df above the
handful of patients one gene provides; `robust = TRUE` protects the prior
from hypervariable genes. The p-values are corrected with a **flat
Benjamini–Hochberg across all tested triplets** — deliberately not the
hierarchical cascade of the full model, because a gene's triplets here are
mostly null and combining them dilutes a single true effect to no better than
Bonferroni.

# Diagnostics

The returned `SpiDEResults` carries three tables in `@diagnostics`, one per
known failure mode:

| table | what it reports | the failure mode it makes visible |
|---|---|---|
| `r2` | per (sample, index) subset and per basis: the $R^2$ of each niche column on that block of the normalisation design (`stage1 = "spanorm"` only) | overlap with the library-size basis warns that the fixed offset could remove niche-correlated variation |
| `inclusion` | per index type: which patients passed `min.cells`, with their condition | **informative dropout** — cell-type abundance can correlate with condition, silently biasing the tested patient subset; a Fisher test warns at $p < 0.05$ |
| `tau2` | the pooled $\hat\tau^2$ per (index, niche) | a large value flags triplets where patient heterogeneity, not stage-1 noise, limits power |

# A worked example

The toy data has no stored normalisation fit, so the example uses the
`stage1 = "ols"` path; each sample is its own patient:

```{r twostage_demo, message = FALSE}
data(toySpiDE)
spe2 <- buildNiches(toySpiDE, sigma = 30)
ts <- twoStageSpiDE(spe2, condition = "condition", sigma = 30,
                    min.cells = 10, stage1 = "ols", fdr = 0.3,
                    verbose = FALSE)
head(results(ts), 3)
head(ts$diagnostics$tau2, 3)
```

The results table keeps the schema of `fitSpiDE()` — `gene`, `ct_index`,
`ct_niche`, the slope contrast `coef`, its moderated `t`, `p.niche` and
`fdr.niche` — so downstream code is agnostic to which estimator produced it.
The returned object has an empty `fits` slot: there is no cell-level GLM to
inspect.

# What it fixes, what it does not, and when to use it

**Fixed: the error structure.** The condition contrast is made across
patients on patient-level numbers, so pseudo-replication cannot enter, and
the between-patient composition association cannot load onto a within-sample
slope.

**The price: power, and cells.** Each patient contributes exactly one slope
per triplet, so per-patient precision — cells — is what sharpens the test,
and below `min.cells` a patient contributes nothing. Thin subsets are also
where its calibration degrades, since nothing is borrowed across patients or
types. A flat `min.cells` that reaches calibration on a real cohort can
delete whole compartments; the calibration vignette gives the measured
trade-offs against `fitSpiDE()`.

**Not fixed: multiplicity.** A full-panel hypothesis space buries real
signal — BH needs the best p-value below $0.05/m$ — and combining a gene's
mostly-null triplets does not help. Restrict `index`, `niche` and the gene
set to a pre-specified hypothesis; a few thousand tests is the order at which
a $p \approx 10^{-5}$ effect survives. Restrict genes rather than index types:
the pathological tail of the null is a gene property, and shrinking the index
set only raises BH's threshold while leaving it in place.

**Use it** when the conclusion must survive a sceptical error audit on a
pre-specified hypothesis, on index types with complete patient inclusion and
enough cells per patient, with `stage1 = "ols"` and `pool.psi = TRUE`. **Use
`fitSpiDE()`** for discovery.

# Session info

```{r sessioninfo}
sessionInfo()
```
