---
title: "Speeding up the mixed-effects fit: the PQL subsampling trade-off"
author: "Dharmesh D. Bhuva"
package: spiDE
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Speeding up the mixed-effects fit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE, collapse = TRUE, comment = "#>", message = FALSE,
  warning = FALSE, fig.width = 7, fig.height = 4, dpi = 150, fig.retina = 2
)
```

```{r libs}
library(spiDE)
library(ggplot2)
source(system.file("spide_palette.R", package = "spiDE"))

bench <- function(f) {
  path <- system.file("extdata", "benchmark", f, package = "spiDE")
  if (file.exists(path)) utils::read.csv(path, check.names = FALSE) else NULL
}
```

# Overview

The mixed-effects fit (`fitSpiDE(..., random = "intercept" | "slope")`) corrects
the anti-conservative inference that cell-level pseudo-replication causes (see
*The spiDE model* vignette). It does so with a Schall/PQL loop that re-estimates
the shared random-effect variance components `tau2`, re-fitting the
negative-binomial GLM once per iteration — the expensive part of the fit.

Those variance components are a *single shared scalar per random-effect group* — a
population quantity that does not need every cell to estimate, exactly as
`SpaNorm::fitNB` already subsamples cells for its dispersion step. `spiDE` exposes
three knobs that can trade `tau2` accuracy for speed, of which only `re.prop` is
usually worth touching:

| argument       | default | effect |
|----------------|---------|--------|
| `re.prop`      | `1`     | fraction of cells sampled per cell type within each sample for the PQL loop; `1` disables subsampling |
| `re.maxit.psi` | `1`     | dispersion iterations inside each PQL step |
| `re.min.cells` | `100`   | per-stratum floor for `re.prop` |

Crucially the loop only supplies `tau2`; a single **final fit on all cells with
full dispersion** produces the coefficients and dispersion that inference uses. So
subsampling perturbs only the shared variance estimate, not the per-gene effect
sizes — but, as this vignette shows, "only the shared variance estimate" is not
the same as "safely, since it's just one number": `tau2` itself is what turns out
to be systematically biased by subsampling, not merely noisier. This vignette
quantifies the time saved and the accuracy cost, using two studies: a
single-run simulation benchmark (`research/`, shipped as this vignette's
figures) and a more rigorous real-cohort replicate study (5 refits per
`re.prop` level on the full 55-sample YTMA cohort) that changed the package
default, described below.

# Time versus accuracy across `re.prop`

The study fits the same realistic dataset at several values of `re.prop`, timing
each fit and correlating its response-niche t-statistics against the fully
accurate reference (`re.prop = 1`, all cells). Two figures tell the story:
subsampling makes the fit markedly faster while the t-statistics it feeds into
inference stay highly correlated with the full-data fit.

```{r pql, fig.height = 4, fig.cap = "Left: the mixed-effects fit is markedly faster once cells are subsampled for the PQL loop (about 1.8x versus the all-cell fit), after which it plateaus. Right: response-niche t-statistics stay highly correlated with the full-data reference (~0.96) across the sampling range."}
timing <- bench("pql_timing.csv")
acc <- bench("pql_accuracy.csv")

if (!is.null(timing)) {
  p_time <- ggplot(timing, aes(re.prop, elapsed)) +
    geom_line(colour = spide_pal$ref) +
    geom_point(size = 3, colour = spide_pal$accent) +
    scale_x_continuous(trans = "reverse") +
    labs(x = "re.prop (cell fraction, reversed)", y = "fit time (s)",
         title = "Fit time versus subsampling")

  if (!is.null(acc)) {
    # anchor the reference (re.prop = 1) at correlation 1
    acc <- rbind(acc, data.frame(re.prop = 1, cor = 1))
    p_acc <- ggplot(acc, aes(re.prop, cor)) +
      geom_line(colour = spide_pal$ref) +
      geom_point(size = 3, colour = spide_pal$accent) +
      scale_x_continuous(trans = "reverse") +
      coord_cartesian(ylim = c(min(0.9, min(acc$cor)), 1.001)) +
      labs(x = "re.prop (cell fraction, reversed)",
           y = "t-stat correlation vs full-data fit",
           title = "Accuracy versus subsampling")
    print(patchwork::wrap_plots(p_time, p_acc))
  } else {
    print(p_time)
  }
}
```

The coefficients come from the full final fit regardless of `re.prop`, so the
response-niche t-statistics stay highly correlated with the full-data reference
even under aggressive subsampling. This single-run simulation, taken on its
own, reads as a favourable trade: the subsample here underestimates `tau2`
(about 0.19 versus 0.25 at full data), and that feeds through to only a small
drop in t-statistic correlation — to roughly 0.96 rather than 1.

That reading turned out to be incomplete, which is why `re.prop = 1` (not
`0.2`) is now the package default. A single simulated timing/accuracy point
can't distinguish *noise* (would average out over replicates) from *bias*
(wouldn't). A follow-up study on the real 55-sample YTMA cohort — 5 independent
refits of identical data at each of `re.prop` = 0.2, 0.5, and 0.8, differing
only in the unseeded stratified cell subsample, plus the deterministic
`re.prop = 1` reference — found both, and the bias matters more:

* **The subsampling noise itself does not shrink** between `re.prop = 0.2` and
  `0.8` (replicate sd stayed in the 0.015-0.017 range throughout), and at every
  level tested it matched or exceeded the genuine between-patient signal (sd
  0.0145 from a 55-fold leave-one-sample-out jackknife) that `tau2` is meant to
  capture.
* **Mean `tau2` rose monotonically with `re.prop`** (0.034 -> 0.052 -> 0.059 ->
  0.074 at `re.prop` = 0.2/0.5/0.8/1) — systematic downward attenuation at every
  `re.prop < 1`, not random jitter around the true value. Averaging several
  `re.prop < 1` replicates would not fix this, because their common mean is
  itself biased low.

So a small drop in t-statistic *correlation* against one particular full-data
run (as in the simulation above) is the wrong lens: it can look reassuring even
while the underlying `tau2` is being systematically underestimated. GPU
backends make `re.prop = 1` affordable in practice (about 1.17x the fit time of
a heavily-subsampled fit, per the LOO study), so it is now the default; treat
`re.prop < 1` as a lower bound on `tau2`, not a point estimate, if used at all.

```{r tau2-table, echo = FALSE}
if (!is.null(timing)) {
  knitr::kable(timing[order(-timing$re.prop),
                      c("re.prop", "elapsed", "tau2")],
               digits = 3, row.names = FALSE,
               col.names = c("re.prop", "fit time (s)",
                             "tau2 (intercept)"),
               caption = "Fit time and the estimated intercept variance component across re.prop.")
}
```

# Scaling to realistic gene counts

A realistic dataset has ten-thousand-plus genes. Because the whole gene set is
fit in a single call (dispersion is moderated across all genes together), the
per-fit cost grows with both the number of genes and the number of cells. The
study records the fit and inference time as the number of cells per sample grows
at fixed gene count.

```{r timing, fig.height = 4, fig.cap = "Fit and inference wall-clock time as the number of cells grows, at a realistic gene count. Inference is the blockable, parallelisable stage."}
tm <- bench("timing.csv")
if (!is.null(tm) && nrow(tm) > 0) {
  long <- rbind(
    data.frame(n_cells = tm$n_cells, stage = "fit", seconds = tm$fit_seconds),
    data.frame(n_cells = tm$n_cells, stage = "inference",
               seconds = tm$test_seconds))
  ggplot(long, aes(n_cells, seconds, colour = stage)) +
    geom_line(linewidth = 0.8) + geom_point(size = 2) +
    scale_colour_manual(values = unname(spide_pal$method[c(1, 2)])) +
    labs(x = "total cells", y = "wall-clock time (s)", colour = NULL,
         title = sprintf("Scaling at ~%s genes",
                         format(round(mean(tm$n_genes), -2), big.mark = ",")))
} else {
  cat("*(Timing summary not found; run the benchmark study in `research/`.)*")
}
```

# Null calibration and the reference degrees of freedom

The mixed-effects fit corrects the fixed-effects anti-conservatism, but the
*reference degrees of freedom* it then uses decides how conservative the
corrected response-niche test is. The benchmark's `null` scenario (genuine
between-patient clustering, no planted effect) measures the type-I error of the
response-niche test across sample sizes for each `random =` mode; the figure
reads its shipped summary.

```{r calibration, fig.height = 4, fig.cap = "Null type-I error of the response-niche test across sample size S, per random-effects mode (gradient layout). Under the between-patient reference df, the mixed modes are over-conservative at small-to-moderate S, approaching the nominal 0.05 (dashed) only as S grows. df.method = 'satterthwaite' adds a per-mode arm whose per-coefficient df holds close to nominal across the whole S range instead."}
cal <- bench("null_type1.csv")
if (!is.null(cal)) {
  cal <- cal[cal$layout == "gradient", , drop = FALSE]
  # older summaries predate the df.method sweep; treat them as the between arm
  if (is.null(cal$df.method)) cal$df.method <- "between"
  cal$arm <- interaction(cal$method, cal$df.method, drop = TRUE, sep = " / ")
  ggplot(cal, aes(S, type1, colour = arm)) +
    geom_hline(yintercept = 0.05, linetype = 2, colour = spide_pal$ref) +
    geom_line(linewidth = 0.7) + geom_point(size = 2) +
    labs(x = "samples (S)", y = "response-niche type-I error", colour = NULL,
         title = "Null calibration of the response-niche test")
} else {
  cat("*(Null calibration summary not found; run the benchmark study in",
      "`research/`.)*")
}
```

Under the default between-patient reference (`df.method = "between"`) the mixed
modes are markedly **over-conservative** at small-to-moderate $S$: at $S=4$ the
$S-2=2$ reference df crushes the response-niche type-I to $\approx 0.001$, and it
climbs toward the nominal $0.05$ only as $S$ approaches the largest sampled size
($30$) — reaching $0.046$ (intercept) / $0.055$ (slope), still short of nominal.
That is the symptom the per-coefficient df targets — the niche coefficients
carry more information than $S-2$ credits them with, most acutely when patients
are few.

The full null study (`research/`, both `df.method` arms on identically seeded
data) confirms this: `df.method = "satterthwaite"` holds close to nominal
across the *entire* sampled $S$ range instead of only at the top end — $0.042$–$0.065$
(slope) and $0.047$–$0.059$ (intercept) from $S=4$ through $S=30$, compared to
between's $0.001$–$0.055$ span. It does **not** overshoot into severe
anti-conservatism the way the between reference undershoots at small $S$: the
worst-case satterthwaite type-I error ($\approx 0.065$ at $S=24$, slope mode) is
a mild, roughly 30% relative inflation over nominal, not the multi-fold
distortion between shows in the opposite direction at $S=4$. This mild
liberal drift at larger $S$ is a real, quantified residual (not a hypothetical
one) — closing it further is exactly what the Kenward–Roger covariance
correction (see *The spiDE model*) would target, if the residual bias turns out
to live there. In exchange for this modest miscalibration, satterthwaite
recovers substantial power the between reference cannot deliver at all in the
small-$S$ regime: in the paired `power` scenario (`fdr = 0.05`, category
`"all"`), between-reference type-I error near zero at $S \le 10$ comes with
type-I-error-*matched* power near zero too (TPR $\approx 0$–$0.002$), while
satterthwaite recovers real power immediately (TPR $\approx 0.10$ at $S=4$,
$\approx 0.32$ at $S=10$), and stays ahead through $S=30$ (TPR $\approx
0.58$–$0.60$ vs $0.50$–$0.52$) — a mean paired TPR gain of $+0.10$–$0.11$
across the sampled grid. Observed FDP is if anything slightly *lower* under
satterthwaite too, so this is not a power-for-validity trade: satterthwaite
improves both, at the cost of the mild large-$S$ liberal drift quantified
above. By construction it leaves the response *main* effect at its
$\approx 0.99\times$ calibration. See *The spiDE model* vignette for the
statistics behind the per-coefficient df.


# Model design: does `CellType:condition` change calibration or power?

The released niche design is

```
~ 0 + covariates + CellType + condition * CellType:(niches) + niches
```

which expands to `condition` + `CellType:niches` + `condition:CellType:niches`
but carries **no `CellType:condition`** term. A cell-type-specific response
*intercept* shift therefore has nowhere to go: the single global `condition`
column captures only the average, and the per-sample random intercept absorbs a
per-sample scalar. Because the niche covariates are not centred, the tested
three-way coefficient can and does absorb the missing shift, scaled by the mean
niche density.

The alternate design adds that term, cell-means coded (the `condition` main
effect is dropped, so there is one coefficient per cell type, each the
tested-level shift *within* that cell type, readable without forming a
contrast). The benchmark runs both as a `design` arm, over the families that can
discriminate between them — null calibration, power, and the combiner
comparison. `pql`, `timing` and `spatial` probe the random-effect machinery and
the layouts, which the design change does not touch, so they run once.

```{r design-cal, fig.height = 4, fig.cap = "Null type-I error of the response-niche test by sample size, for the released design and the design with CellType:condition restored (gradient layout, satterthwaite df). Restoring the term costs ~k degrees of freedom out of several hundred, so calibration should be unchanged; a visible gap would mean the added block is absorbing signal rather than nuisance."}
cal <- bench("null_type1.csv")
if (!is.null(cal) && !is.null(cal$design) &&
      length(unique(cal$design)) > 1) {
  d <- cal[cal$layout == "gradient", , drop = FALSE]
  if (!is.null(d$df.method)) d <- d[d$df.method == "satterthwaite", , drop = FALSE]
  d$arm <- interaction(d$method, d$design, drop = TRUE, sep = " / ")
  ggplot(d, aes(S, type1, colour = arm)) +
    geom_hline(yintercept = 0.05, linetype = 2, colour = spide_pal$ref) +
    geom_line(linewidth = 0.7) + geom_point(size = 2) +
    labs(x = "samples (S)", y = "response-niche type-I error", colour = NULL,
         title = "Null calibration by model design")
} else {
  cat("*(No `design` arm in the shipped summaries yet. Run the alternate arm:",
      "`SPIDE_DESIGN_ARM=celltype-response SPIDE_DESIGN_LIB=<lib>",
      "bash research/slurm/submit.sh`, then re-aggregate and reinstall.)*")
}
```

```{r design-power, fig.height = 4, fig.cap = "True-positive rate at FDR 0.05 versus effect size, by model design. The added term spends degrees of freedom, so a small power cost is expected; a large one would argue against restoring it."}
pw <- bench("power_fdr.csv")
if (!is.null(pw) && !is.null(pw$design) && length(unique(pw$design)) > 1) {
  d <- pw[pw$fdr == 0.05 & pw$sweep == "effect", , drop = FALSE]
  if (!is.null(d$df.method)) d <- d[d$df.method == "satterthwaite", , drop = FALSE]
  d$arm <- interaction(d$method, d$design, drop = TRUE, sep = " / ")
  ggplot(d, aes(beta, tpr, colour = arm)) +
    geom_line(linewidth = 0.7) + geom_point(size = 2) +
    labs(x = expression(paste("planted effect ", beta)),
         y = "TPR at FDR 0.05", colour = NULL,
         title = "Power by model design")
} else {
  cat("*(No `design` arm in the shipped power summaries yet.)*")
}
```

**What the arm actually varies.** Because `design` selects a whole package
build, the alternate arm bundles two changes that ship together: the added
`CellType:condition` block, *and* two-sided p-value combination under
`combine = "cauchy"` (Brown keeps one-sided input). Differences between the arms
in the figures above are therefore attributable to the pair, not to the design
term alone. That is deliberate — the two are intended to ship together — but it
means these panels answer "is the new build calibrated and no less powerful?"
rather than "what does `CellType:condition` do in isolation".

Sidedness is characterised on its own in
`vignette("spiDE-cauchy-vs-brown")`, which derives why one-sided ACAT cancels
on genes with opposing niche effects and shows it costs nothing where the
one-sided convention was already safe. The combiner table below carries both
`combine` and `design`, so the combiner axis can still be read within an arm.

# Summary

* Subsampling cells for the PQL loop (`re.prop < 1`) plus a single inner
  dispersion iteration (`re.maxit.psi = 1`) cuts the mixed-effects fit time
  substantially, while the final all-cell fit always keeps the coefficients
  exact — only the shared `tau2` is affected.
* That effect on `tau2` is not just noise: a real-cohort replicate study found
  it systematically biased downward at every `re.prop < 1` tested, an
  attenuation that gets worse (not better) the more the correlation-based
  simulation check above might suggest is "fine". Response-niche t-statistic
  correlation against a full-data fit is a poor proxy for this, since it can
  look reassuring even while `tau2` itself is off.
* `re.prop = 1` (no subsampling, fully reproducible, no seed required) is
  therefore the package default; GPU backends make it affordable. Lower it
  only when `tau2`'s absolute scale doesn't matter (e.g. a quick feasibility
  check), and set an external `set.seed()` if reproducibility of a
  `re.prop < 1` run matters.

Timing numbers reflect the machine that produced the benchmark tables; regenerate
them for your own hardware with the study in `research/`.

```{r session}
sessionInfo()
```
