---
title: "The spiDE model"
author: "Dharmesh D. Bhuva"
package: spiDE
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{The spiDE model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette explains what spiDE computes and why, one step at a time. Each
step starts in plain language and then gives the formula, so the text can be
read with or without the maths. The quick start (`vignette("spiDE")`) shows how
to run the functions; the calibration vignette (`vignette("spiDE-calibration")`)
shows how well the tests hold up on real data.

# The question

Take one cell type, the **index** type (say, tumour cells), and a second cell
type, the **niche** type (say, fibroblasts). spiDE asks whether a gene's
expression in the index cells changes with the number of niche cells around
them.

- The **pooled test** asks: across patients, is there on average such a
  change?
- The **condition-specific test** asks: does the change differ between two
  groups of patients (responders and non-responders, early and late stage)?

One gene, one index type and one niche type make a **triplet**. spiDE tests
every triplet.

Both questions are about patients, not cells. The condition belongs to a
patient, and so does the way a patient's tissue responds to its
neighbourhood. spiDE therefore treats the **patient as the unit of
replication**. A result is strong when patients agree, not when a patient
contributes many cells. Treating cells as replicates would make every result
look more certain the more cells were measured, even if patients disagreed.

# Step 1: the niche

For every cell, `buildNiches()` measures how many cells of each type surround
it. It places a Gaussian bump of width $\sigma$ (the *bandwidth*, in the units
of the spatial coordinates) on every cell of a type and adds them up, which
gives a smooth *density* of that type at every point of the tissue. The
density of type $n$ around cell $i$ is $d_{in}$.

spiDE uses $\ell_{in} = \log(1 + d_{in})$, which compresses very dense
neighbourhoods so that a few crowded areas do not dominate.

The bandwidth sets what "nearby" means: a small $\sigma$ looks at immediate
neighbours, a larger one at the surrounding tissue. Fit one bandwidth at a
time.

# Step 2: a regression within each index type

Each index type is analysed separately. Within it, each gene's counts are
modelled by a **negative binomial regression**, the standard model for
sequencing counts. Its mean can depend on covariates and its variance allows
for more spread than pure sampling noise. For gene $g$ in cell $i$ of
patient $s(i)$:

$$
\log \mathrm{E}[y_{gi}] \;=\;
\underbrace{a_{g,s(i)}}_{\text{patient's own level}} \;+\;
\underbrace{\gamma_g^\top z_i}_{\text{cell covariates}} \;+\;
\underbrace{\textstyle\sum_{n} \beta_{gn}\, \ell_{in}}_{\text{niche slopes}}
$$

- **$a_{gs}$, one intercept per patient.** Each patient gets its own baseline
  level of the gene. This absorbs everything that is constant within a
  patient: its average expression, its sample handling, its condition, and
  its average niche. What is left to explain the niche slopes is variation
  *within* each patient: cells of the same patient in denser or sparser
  neighbourhoods. So $\beta_{gn}$ is a **within-patient slope**. It answers "in
  the same tissue, do cells with more neighbours of type $n$ express more of
  the gene?", not "do patients with more type-$n$ cells express more of it?".
- **$z_i$, cell covariates.** By default the cell's log library size (its
  total count, centred), with a slope per gene. This adjusts for sequencing
  depth (`depth = "loglib"`; see the options below). Other cell-level
  measurements, such as cell area, are added with `covariates`.
- **$\ell_{in}$, the niche columns.** Every cell type other than the index
  type enters, so each slope is adjusted for the other niches. Only the types
  named in `niche` are tested. An index type is never tested against its own
  niche, including any `mergeNiches()` group it belongs to.
- **The dispersion $\psi_g$** measures how much more variable the gene is than
  pure counting noise. It is estimated per gene.

The fit uses `SpaNorm::polishNB()`, which fits each gene on its own. The patient
intercepts are handled algebraically, so a fit with many patients costs about
the same as one with few. Genes are independent, so `BPPARAM` fits them in
parallel.

**Depth options.** `depth = "loglib"` (the default) uses one slope on the log
library size. `"nonlinear"` lets the effect of depth bend (a spline with three
degrees of freedom). `"spatial_spline"` lets it vary smoothly across each
section. `"none"` leaves depth out, for use with `offset`, a precomputed
log-scale offset per gene and cell. Library size mixes technical capture with
biological RNA content, so adjusting for it also removes any part of the niche
effect that works through a cell's total RNA. A slope is then expression
*relative to the cell's total*.

# Step 3: each patient's own slope

The regression above gives one pooled slope per gene and niche. The **slopes
engine**, the default, then asks what each patient's own slope is.

Refitting the model for every patient would be slow and unstable for patients
with few cells. Instead, spiDE starts from the pooled fit and lets each
patient's own cells move the slope once toward what fits that patient best.
This is a single step of the Newton-type algorithm (Fisher scoring) used to fit
such models. With working weights $w_i = \mu_i/(1 + \psi\mu_i)$ and residuals
$r_i = (y_i - \mu_i)/(1 + \psi\mu_i)$ at the pooled fit, and $\tilde\ell_i$ the
niche columns centred within the patient:

$$
\hat\beta_{gs} = \hat\beta_g + I_s^{-1} u_s, \qquad
I_s = \sum_{i \in s} w_i \tilde\ell_i \tilde\ell_i^\top, \qquad
u_s = \sum_{i \in s} \tilde\ell_i\, r_i .
$$

$u_s$ is how strongly the patient's cells pull away from the pooled slope, and
$I_s$ is how much information they carry. A patient with many cells and a wide
range of niche densities has a large $I_s$ and a precise slope. A patient
whose niche columns barely vary, or vary together, gives no usable slope for
that gene and is left out of its test.

**How precise is a patient's slope?** Two answers are computed.

- The **model variance**, $v^{\text{model}}_{gs} = \mathrm{diag}(I_s^{-1})$,
  is what the regression would give if every cell were independent.
- Neighbouring cells are not independent: they share local tissue conditions,
  so their residual expression is similar. A **spatial variance** $v^{\text{tile}}$
  corrects for this. Each section is cut into square tiles (side `tile`,
  default $3\sigma$), the patient's score is summed per tile, and the slope's
  variance is computed from how much the tile sums vary:
  $v^{\text{tile}} = \mathrm{diag}\big(I_s^{-1} (\sum_t U_t U_t^\top) I_s^{-1}\big)\, T/(T-1)$
  over the $T$ tiles, with $U_t$ the score of tile $t$. A patient with fewer
  than `min.tiles` tiles takes the model variance scaled by the typical ratio.

The tile variance of a single gene is noisy, so spiDE summarises it as **one
inflation factor per patient and niche**, shared by all genes:
$f_{sn} = \max\{1, \mathrm{median}_g(v^{\text{tile}}_{gsn}/v^{\text{model}}_{gsn})\}$.
A patient's slope variance is then $v_{gs} = v^{\text{model}}_{gs} f_{sn}$.

Because the fit never sees the condition, any patient-level condition can be
tested afterwards (`testSpiDE(fit, condition = ...)`). Retesting under a
reshuffled condition reruns only the test, which makes permutation checks
cheap.

# Step 4: combining patients

Each gene now has one slope per patient, each with a variance. The tests
combine them across patients.

**How much patients disagree.** Patients' slopes differ for two reasons:
each is estimated with noise ($v_{gs}$), and patients genuinely differ. The
genuine spread is the **between-patient variance** $\tau^2_g$, estimated from
how much more the slopes scatter than their noise explains (the
DerSimonian-Laird estimator of meta-analysis, computed without the
condition).

**Weights.** Patient $s$ is weighted by

$$
w_{gs} = \frac{1}{v_{gs} + \tau^2_g}.
$$

A precise patient counts more, but never unboundedly more. When patients
genuinely disagree ($\tau^2_g$ large), every patient's weight approaches the
same value, because no single patient can speak for the rest.

## The pooled test

**Is the average slope zero?** The weighted average of the patients' slopes is
tested with `limma`'s moderated $t$-test. The moderated test borrows
information across genes to stabilise each gene's variance, which helps when
patients are few. Two refinements keep it honest:

- **Borrowing is matched by expression level.** A bright gene's variability is
  compared with other bright genes, not with the typical gene
  (`eBayes(trend = TRUE, robust = TRUE)` on the gene's log mean expression).
- **The degrees of freedom count effective patients.** When a few cell-rich
  patients carry most of the weight, the test should not behave as if every
  patient contributed equally. The Kish effective number of patients,
  $n_{\text{eff}} = (\sum_s w_s)^2 / \sum_s w_s^2$, equals the number of
  patients when the weights are equal and falls towards one as they
  concentrate. The $t$ statistic is referred to $\min(\nu_{\text{limma}},
  n_{\text{eff}} - 1)$ degrees of freedom.

## The condition-specific test

**Does the average slope differ between the two groups?** The patients' slopes
are regressed on the condition, with the same weights $w_{gs}$:

$$
\hat\beta_{gs} = b_0 + \Delta_g\, c_s + [\text{strata}] + \text{error},
$$

where $c_s$ is 1 for the second condition and 0 for the first. The estimate
$\hat\Delta_g$ is the difference in average slope, second level minus first.

Its standard error does not assume the weights are exactly right. It is a
**robust (HC2 sandwich) standard error**, computed from how far each patient's
slope actually sits from its group's fit, with small-sample **Bell-McCaffrey
degrees of freedom**. This matters when some patients' slopes are noisier than
their weights suggest. A marker gene that is patchy within the index type is
an example: its slope can swing with where the patches fall. Such a patient
widens the interval rather than producing a call.

The weights come from the gene-shared variances $v_{gs}$ and never from a
gene's own tile variance. A gene's own tile variance, computed at the pooled
fit, grows when a patient's slope sits far from the pooled slope. Weighting by
it would down-weight exactly the patients that differ most and shrink the
estimated difference toward zero.

`strata` (a patient-level slide or batch column) adds one term per stratum to
the condition model. Use it when the condition is partly confounded with
slide, so that slide differences in slope are not read as condition
differences. A condition entirely confounded with the strata cannot be tested
and is refused.

# Multiple testing

spiDE tests many triplets, so it controls the **false discovery rate** (the
expected share of false calls among the calls) with the Benjamini-Hochberg
procedure. `testSpiDE()` reports a $q$-value per row:

- **pooled**: over every tested triplet;
- **condition-specific**: by default (`procedure = "filtered"`) over the
  triplets whose pooled test passes at `fdr`. Rows outside that family have
  `q = NA`. The pooled test ignores the condition, so this family is the same
  under any relabelling of patients and the false discovery rate is preserved.
  The correction is then spent where there is a niche effect to modify.
  `procedure = "all"` uses every triplet instead.

The estimate is on the natural-log scale per unit $\log(1 + d)$. A pooled slope
of 0.2 means expression rises by about 22% ($e^{0.2}$) for each unit increase
in $\log(1 + \text{density})$, in the same patient.

# The sandwich engine

`engine = "sandwich"` takes a different route to the same question. It fits
the regression of Step 2 once, with the condition's interaction terms
$c_s \ell_{in}$ when given a condition, and tests the coefficients directly.
Its standard error is a **patient-clustered CR2 sandwich** (Bell and McCaffrey
2002; Pustejovsky and Tipton 2018) with Bell-McCaffrey degrees of freedom: it
lets patients differ without modelling how. It needs the condition at fit
time. It is simpler but less powerful when effects vary between patients, and
`spiGSEA()` needs the slopes engine.

# Gene sets: `spiGSEA()`

A gene set is tested exactly like a gene. In each patient, the set's slope is
the average of its genes' slopes, each gene first divided by its typical
spread between patients, $\sqrt{\tau^2_g + \mathrm{median}_s v_{gs}}$, so
that a highly expressed gene does not dominate.

- **Competitive** (the default): the same average over the other tested genes
  is subtracted. The test then asks whether the set moves *more than the
  typical gene*. This guards against a shift that moves every gene at once,
  such as signal spilling over from neighbouring cells.
- **Self-contained**: nothing is subtracted, and the test asks whether the set
  moves at all.

The patients' set slopes go through the same pooled and condition-specific
tests as genes' slopes. Because patients are the replicates, the correlation
between genes of a set is already reflected in how much the set's slope varies
from patient to patient, and needs no separate estimate. $q$-values are given
within each (index, niche, test) column and over every set tested
(`q.global`).

# A different question: niche abundance

The patient intercepts of Step 2 absorb one association on purpose. Patients
whose index cells sit, on average, among more niche cells may express a gene
differently *as a whole*. That is a comparison between patients, with as many
data points as patients. It can reflect biology, differences in tissue
composition, or signal spilling between neighbouring cells.
`testNicheAbundance()` asks it directly:

- per patient and index type, it averages expression into a pseudobulk
  profile (log counts per million) and computes the patients' mean niche
  density;
- it regresses one on the other across patients with `limma`, adding the
  difference between conditions when a condition is given.

A result from this test says "patients with more of this niche differ". A
result from `testSpiDE()` says "within a patient, cells with more of this niche
differ". Report each as what it is.

`plotNicheResponse(adjust = FALSE)` draws both on the example data. Each
segment is one patient's slope, the within-patient association spiDE tests; the
dashed line through the patients' means (open circles) is the between-patient
association the intercepts absorb:

```{r abundance-figure, fig.width = 6, fig.height = 3.5, message = FALSE}
library(spiDE)
data(toySpiDE)
spe <- buildNiches(toySpiDE, sigma = 30)
res <- spiDE(spe, condition = "condition", sigma = 30, index = "A", verbose = FALSE)
plotNicheResponse(spe, res, gene = "G1", index = "A", niche = "B", adjust = FALSE)
```

# What the model does not address

- **Smoothness of the niche itself.** A niche density is smooth over space.
  The spatial variance accounts for residual expression that is similar in
  neighbouring cells, and the calibration checks use nulls that keep the
  niche's smoothness, but the model does not describe it.
- **Straight-line effects.** Each slope is linear in $\log(1 + d)$. A gene that
  responds only above a threshold density is captured only partly.
- **Which average.** The pooled slope is a weighted average over patients,
  with weights that favour precise patients. When cell-rich patients have
  systematically different slopes from sparse ones, this average differs from
  a simple average over patients.
