---
title: "The spiDE model: neighbourhood-dependent DE and mixed-effects inference"
author: "Dharmesh D. Bhuva"
package: spiDE
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{The spiDE model and mixed-effects inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, collapse = TRUE, comment = "#>")
set.seed(1)
```

# Overview

This vignette documents the statistical model behind `spiDE` (Spatial Influence DE) and a **mixed-effects extension** that corrects the anti-conservative inference caused by cell-level pseudo-replication. It has three parts:

1.  the full `spiDE` model — how the effective niche is built and how gene expression is modelled as a function of cell type, neighbourhood, and clinical response;
2.  why a cell-level Wald test (`random = "none"`) is anti-conservative when many cells come from few patients; and
3.  the mixed-effects correction, implemented through the **ridge = random effects** equivalence so it reuses the same `SpaNorm::fitNB` engine.

`fitSpiDE()` and `spiDE()` **default to `random = "intercept"`**. On a null with
per-sample intercepts, `random = "none"` rejects at $\approx 0.71$ against a
nominal $0.05$, while the random intercept holds $\approx 0.04$ against a
calibrated pseudobulk reference of $\approx 0.04$. `"none"` remains available
for back-compatibility and for reproducing pre-correction results; it should not
be used for inference.

For a task-oriented quickstart, see the companion vignette *Neighbourhood-dependent differential expression with spiDE*.

```{r load, message = FALSE}
library(spiDE)
library(SpatialExperiment)
library(ggplot2)
source(system.file("spide_palette.R", package = "spiDE"))
```

# The full spiDE model

## Notation

Let $Y_{c,g}$ be the (SpaNorm-normalised) expression of gene $g \in \{1,\dots,G\}$ in cell $c \in \{1,\dots,C\}$. Each cell has a type $t_c \in \{1,\dots,T\}$ and belongs to a patient tissue sample $s_c \in \{1,\dots,S\}$. The clinical response of a sample is $r_{s} \in \{0 = \text{non-responder},\, 1 = \text{responder}\}$; because it is a property of the *patient*, $r_{s}$ is **constant across all cells of a sample** — a fact that drives everything below.

## Effective niche

Every quantity in this section is computed **within one sample at a time, entirely independently of every other sample**. Let $\mathcal C_s = \{c' : s_{c'} = s\}$ denote the cells belonging to sample $s$. For each cell $c \in \mathcal C_s$ we summarise its local neighbourhood by a kernel-smoothed intensity of every surrounding cell type, computed **from $\mathcal C_s$ alone**. With a Gaussian kernel $K_\sigma$ of bandwidth $\sigma$, the intensity of type $t'$ around cell $c$ is

$$
\Lambda_{c,t',s}(\sigma) \;=\; e_s(c) \!\!\sum_{c' \,\in\, \mathcal C_s}\!\!
  K_\sigma\!\left(d_{c,c'}\right)\,
  \mathbb{1}(t_{c'} = t'),
$$

where the sum runs **only over cells of the same sample** $s$, $d_{c,c'}$ is the (planar, within-sample) distance between two cells of $\mathcal C_s$, and $e_s(c)$ is Diggle's edge correction *for sample $s$'s own spatial window* — the smallest bounding box containing $\mathcal C_s$'s coordinates. Concretely, `buildNiches()` never forms a $C \times C$ inter-cell distance matrix or a single shared spatial window: `.speApply()` first splits the `SpatialExperiment` into one subset per sample, and `.effectiveNiche()` builds a fresh `spatstat` point pattern, window and kernel density surface **for that subset only**. Two cells from different samples that happen to share the same $(x,y)$ coordinates (routine, since each sample is imaged/segmented on its own coordinate system) therefore never interact in $\Lambda$ — there is no notion of "distance between a cell in sample 3 and a cell in sample 7". The expected log-count of type $t'$ in a disc of radius $\sigma$ is

$$
N_{c,t',s}(\sigma) \;=\; \log\!\left(\pi\sigma^2\,\Lambda_{c,t',s}(\sigma) + 1\right).
$$

Stacking over cell types gives the **effective niche composition** $N_{c,s}(\sigma) = \big(N_{c,1,s}(\sigma),\dots,N_{c,T,s}(\sigma)\big)$, still computed one sample at a time. Once every sample has been processed (in parallel, via `BiocParallel::bplapply`), the per-sample results are row-bound — not recomputed jointly — into one $C \times T$ matrix $N(\sigma)$ covering all samples. In the package this is `buildNiches()`, which stores $N(\sigma)$ as `reducedDim(spe, "Niche<sigma>")`.

## The negative binomial GLM

For a fixed bandwidth $\sigma$, gene $g$ is modelled with a negative binomial GLM with dispersion $\phi_g$ and log-mean

$$
\log \mathbb{E}\!\left(Y_{c,g}\mid s_c=s, t_c=t, r_s, N_{c,s}(\sigma)\right)
= \gamma_g L_s + \zeta_{g,t} + \eta_g\, r_s
 + \sum_{t'=1}^{T} N_{c,t',s}(\sigma)\,\alpha_{g,t,t'}
 + r_s \sum_{t'=1}^{T} N_{c,t',s}(\sigma)\,\beta_{g,t,t'} .
$$

Writing $\mu_{c,g} = \mathbb{E}(Y_{c,g}\mid\cdots)$ for the modelled mean, the negative-binomial **variance function** — how noisy a count is around its mean — is

$$
\mathrm{Var}(Y_{c,g}) = \mu_{c,g} + \phi_g\,\mu_{c,g}^2 ,
$$

i.e. Poisson noise ($\mu_{c,g}$) plus an extra, gene-specific overdispersion term that grows with the square of the mean. $\phi_g$ is estimated genome-wide (moderated across genes by `edgeR::estimateDisp()` inside `SpaNorm::fitNB`) and reappears throughout this vignette — it sets the IRLS working weights below and, later, the SE scaling of the mixed-effects fit.

The coefficients are:

| term | meaning |
|-----------------------------|-------------------------------------------|
| $\gamma_g L_s$ | sample-specific library-size effect (`computeSizeFactors()` offset) |
| $\zeta_{g,t}$ | cell-type-specific expression |
| $\eta_g$ | global response effect |
| $\alpha_{g,t,t'}$ | average effect of niche type $t'$ on gene $g$ in cell type $t$ |
| $\beta_{g,t,t'}$ | **response-specific** effect of niche type $t'$ in cell type $t$ |

The scientific quantity of interest is $\beta_{g,t,t'}$. We test

$$
H_0:\ \beta_{g,t,t'} = 0 \qquad\text{vs}\qquad H_1:\ \beta_{g,t,t'} \neq 0 .
$$

When $\beta_{g,t,t'} \neq 0$, niche cell type $t'$ has a response-specific association with the expression of gene $g$ in index cell type $t$.

In the package the design is assembled by `nicheDesign()` / `.buildNicheDesign()` as `~ 0 + <covariates> + CellType + CellType:condition + CellType:(niches) + CellType:condition:(niches) + niches`. This is **cell-means coding**: there is no bare `condition` main effect, so the $\eta_g$ response contrast is carried by the per-cell-type `"ResponseCellType"` columns, and the $\beta_{g,t,t'}$ (three-way `condition:CellType:niche`) columns are `"ResponseNiche"` — the ones carried through testing. The whole gene set is fit in one `SpaNorm::fitNB(Y, W, ...)` call (the dispersion $\phi_g$ is moderated across genes), and per-gene Wald statistics for $\eta_g$ / $\beta_{g,t,t'}$ are combined across correlated niches (Brown's method), across bandwidths (Cauchy combination), and corrected with a hierarchical FDR.

# The pseudo-replication problem

The GLM above treats each of the $C$ cells as an independent observation. That is fine for the within-sample coefficients ($\zeta$, $\alpha$), but the response-related coefficients $\eta_g$ and $\beta_{g,t,t'}$ are **contrasts on the patient-level label** $r_s$. There are only $S$ independent patients, yet the Wald information matrix

$$
\widehat{\mathrm{Var}}(\hat\theta) \;=\; \phi_g\left(W_\text{sub}^\top D_g\, W_\text{sub}\right)^{-1},
\qquad D_g = \mathrm{diag}\big(w_{c,g}\big),\ \ w_{c,g} = \frac{\mu_{c,g}}{1+\phi_g\mu_{c,g}},
$$

sums over all $C$ cells. The standard error of a response contrast therefore shrinks like $1/\sqrt{C}$ instead of $1/\sqrt{S}$, and with $C \gg S$ the Wald $z$-statistics are badly inflated. This reproduces the known behaviour of the original flat-script implementation and is documented as a property of `random = "none"`, which is why it is no longer the default.

We can see it directly. Below we simulate a **null** data set: every cell of a patient shares a per-patient random shift, but there is *no* response effect at all — so every response test should be null.

```{r sim}
sim_clustered <- function(n_samples = 10, n_per = 90, n_genes = 300,
                          field = 500, sd_patient = 0.7, seed = 1) {
  set.seed(seed)
  cts <- c("A", "B", "C"); gn <- sprintf("G%d", seq_len(n_genes))
  sids <- sprintf("S%d", seq_len(n_samples))
  cond <- rep(c("Responder", "Non-responder"), length.out = n_samples)
  names(cond) <- sids
  cells <- lapply(sids, function(sid) {
    x <- runif(n_per, 0, field); y <- runif(n_per, 0, field)
    ct <- ifelse(x > 0.6 * field & runif(n_per) < 0.7, "B",
                 sample(c("A", "C"), n_per, replace = TRUE))
    data.frame(sample_id = sid, condition = cond[[sid]],
               x = x, y = y, cell_type = ct)
  })
  cd <- do.call(rbind, cells); n <- nrow(cd)
  cd$cell_id <- sprintf("c%d", seq_len(n))
  base <- matrix(rnorm(n_genes * 3, 1.5, 0.6), n_genes, dimnames = list(gn, cts))
  lmu <- base[, cd$cell_type, drop = FALSE]
  # per-(gene, patient) random intercept: real clustering, NO response effect
  u <- matrix(rnorm(n_genes * n_samples, 0, sd_patient), n_genes,
              dimnames = list(gn, sids))
  lmu <- lmu + u[, cd$sample_id, drop = FALSE]
  counts <- matrix(rnbinom(length(lmu), mu = as.vector(exp(lmu)), size = 5),
                   n_genes, dimnames = list(gn, cd$cell_id))
  list(spe = SpatialExperiment(assays = list(counts = counts),
                               colData = S4Vectors::DataFrame(cd),
                               spatialCoords = as.matrix(cd[, c("x", "y")])),
       u = u)   # the true per-(gene, sample) random intercepts, kept for later
}

sim_null <- sim_clustered()
spe <- buildNiches(sim_null$spe, sigma = 30)
```

300 genes, not a handful, because $\tau^2$ below is a **single value shared across every
gene fit** (more on why in "The fitting procedure"): estimating a variance from
only a few genes is noisy, exactly the way estimating a variance from only a
few observations is noisy in any other setting.

A helper that fits one random mode once and returns both the response p-values
and the fitted object (so later sections can reuse it instead of re-fitting):

```{r respp}
fit_mode <- function(random) {
  # df.method = "between" (the scalar S - 2) is pinned here deliberately: this
  # section is about the mixed-effects *correction*, and the between reference
  # keeps the demo cheap to rebuild. The default "satterthwaite" per-coefficient
  # df is demonstrated in its own section below.
  f <- fitSpiDE(spe, "condition", sigma = 30, random = random,
                df.method = "between", verbose = FALSE)
  res <- testSpiDE(f, spe = spe)          # populates the Wald inference slots
  ff <- fits(res)[[1]]
  # Under cell-means coding there is no bare `condition` main effect: the
  # response contrast is carried by the CellType:condition columns, one per
  # cell type. Match on column NAMES -- `covtype` spans every design column
  # while `t_stat` holds only the tested ones.
  ct <- as.character(ff$covtype)
  rc <- intersect(colnames(ff$t_stat),
                  ff$coefmap$covariate[ct == "ResponseCellType"])
  # @df is a per-column NAMED VECTOR under the default df.method =
  # "satterthwaite", and NULL (fixed fit) or a scalar under "between".
  # Recycling a vector against a genes x columns matrix would run down rows,
  # not across columns, so expand it explicitly.
  tt <- ff$t_stat[, rc, drop = FALSE]
  df <- if (is.null(ff$df)) Inf else if (length(ff$df) == 1L) ff$df else ff$df[rc]
  dfm <- matrix(rep(df, each = nrow(tt)), nrow = nrow(tt))
  list(fit = ff, p = 2 * pt(-abs(tt), dfm))
}
fixed <- fit_mode("none")
mean(fixed$p < 0.05)  # should be ~0.05 under the null
```

The fixed-effects fit rejects a large fraction of these null tests at $\alpha = 0.05$ — grossly anti-conservative.

# Mixed-effects correction via ridge

If you know GLMs but have never fit or interpreted a random-effects (mixed)
model, this section is written for you. It builds the model up from the one
question a fixed-effects modeller always asks first, then walks through
*exactly* what `fitSpiDE()` computes, step by step, with a small worked
example you can run and modify yourself.

## Why not just add sample as a fixed effect?

The pseudo-replication problem above comes from ignoring that cells cluster by
patient. Your instinct from fixed-effects modelling is probably: *if
`sample_id` is a nuisance grouping variable, just add it to the design as
another factor*, the way you would `batch` or `site`. It is worth seeing
concretely why that does not work here, because the reason is exactly what
motivates a random effect instead.

```{r whynotfixed}
set.seed(42)
# a deliberately modest n_per and baseline mean: with abundant cells per
# sample, each sample's own data pins down its baseline so precisely that a
# random effect has little left to correct -- see "Shrinkage" below
S <- 12; n_per <- 12
toy_sample <- factor(rep(seq_len(S), each = n_per))
toy_resp   <- rep(rbinom(S, 1, 0.5), each = n_per)   # constant within a sample
true_tau2  <- 0.49                                    # matches sd_patient = 0.7
u_true     <- rnorm(S, 0, sqrt(true_tau2))
eta_true   <- 0.3 * toy_resp + u_true[as.integer(toy_sample)]
toy_y      <- rpois(length(eta_true), exp(eta_true))
toy <- data.frame(y = toy_y, resp = toy_resp, sample_id = toy_sample)

naive_fixed <- glm(y ~ sample_id + resp, family = poisson, data = toy)
coef(naive_fixed)["resp"]
```

`resp` comes back `NA`. This is not a numerical accident: because *every cell
of a sample shares the same response label*, `resp` is an exact linear
combination of the sample dummy columns (it is just "which samples are
Responders", encoded twice). A design that already contains a free column per
sample has nothing left to estimate the response contrast from — the two
sets of columns span the same space, and `glm()` correctly refuses to invent
information that is not there. (`spiDE` enforces the same logic the other way
around: `checkSample()` **errors** if you pass a sample-constant covariate
together with `random != "none"`, since the random intercept would make it
exactly as inestimable.)

Even setting the collinearity aside, giving every sample its own completely
free (unpenalised) parameter throws away the one thing a mixed model is
for: **borrowing strength across patients**. A free per-sample parameter fits
that one sample's data perfectly and tells you nothing about how much
patients differ *in general* — which is precisely the quantity ($\tau^2$
below) the corrected test needs.

What we actually want sits **between** two extremes:

-   *ignore the patient entirely* (fixed-effects default) — one shared
    parameter for everyone, badly anti-conservative, as shown above;
-   *give every patient a fully free parameter* (fixed factor) — inestimable
    here, and even when it isn't, it doesn't pool information across
    patients.

A **random effect** is the middle ground: each patient gets its own deviation
$u_s$, but those deviations are constrained to be draws from a shared
distribution, $u_s \sim \mathcal N(0, \tau^2)$, rather than free parameters.
$\tau^2$ — how much patients genuinely differ — is *estimated from the data*,
and each patient's own $\hat u_s$ is pulled ("shrunk") toward zero by an
amount that depends on how reliable its own data is relative to $\tau^2$. The
rest of this section makes that precise and shows it happening on real
numbers.

### An aside: nesting the samples, and does cell imbalance matter?

The collinearity above is an artefact of coding each sample as a *free,
independent* effect. You can instead **nest** the samples within their
condition group — code them with `contr.sum` inside each group — and then the
response contrast *is* estimable: with $S$ samples you fit $S-2$ within-group
contrasts plus `resp`, rather than $S-1$ collinear dummies. So estimability was
never the real obstacle.

The real obstacle is the **error stratum**: making `resp` estimable does not by
itself make its test honest. On the page-5 null (per-sample intercepts, no
response effect) we test `resp` four ways, and — because unequal cell counts are
an obvious worry — sweep a **cell-count imbalance from 1× to 10×** (the larger
samples get up to 10× the cells, balanced across conditions). Response-effect
null type-I at $\alpha = 0.05$ (should be $\approx 0.05$; a few folds shown, full
1×–10× sweep in `analysis/statistician-review/imbalance.R`):

```{r nested_imbalance_table, echo = FALSE}
knitr::kable(
  data.frame(
    imbalance = c("1x (balanced)", "3x", "5x", "10x"),
    `fixed (cell SE)` = c(0.730, 0.837, 0.898, 0.905),
    `nested, cell SE` = c(0.763, 0.832, 0.833, 0.825),
    `nested, between-sample df` = c(0.047, 0.048, 0.040, 0.038),
    `random intercept` = c(0.048, 0.053, 0.042, 0.045),
    check.names = FALSE),
  caption = paste("Response-effect null type-I across a cell-imbalance sweep.",
                  "Nesting alone (cell SE) does not help -- and the plain",
                  "cell-level fixed test only worsens as imbalance grows;",
                  "testing against the between-sample stratum, or equivalently a",
                  "random intercept, stays calibrated all the way to 10-fold."))
```

Two things to take from this. First, **the nesting alone is not the fix**:
tested at the cell SE it is as anti-conservative as ignoring samples entirely
($\approx 0.8$, and *worse* as imbalance grows); it is the between-sample stratum
($S-2$ df) — exactly what the random intercept supplies — that calibrates it.
Second, a properly-tested nested *fixed* model is a legitimate alternative to the
random effect **for the main effect**, and both stay calibrated across the entire
1×–10× imbalance sweep — so the choice between them is not about calibration. The
random effect's one statistical edge is that it pools the between-sample variance
across genes (a single $\tau^2$) and so borrows strength for the genes measured
least reliably; that advantage is largest with **few patients** and fades as the
cohort grows (each gene's own between-sample variance then has ample $S-2$ df).
We use the random-effects model because it delivers that pooling and — the part
that actually matters for the niche interactions — the per-coefficient degrees of
freedom (see *Per-coefficient degrees of freedom* below), and because it handles
the NB-GLM and covariates without hand-assembling error strata.

#### A nested fixed model was built, measured, and rejected

The "between-sample df" arm above is an **unweighted $t$-test on the $S$ sample
means**. That test is calibrated — it is essentially a pseudobulk analysis, and
holds $0.041$–$0.051$ across every design point we have measured. What does
*not* work is the natural way of folding that stratum back into the GLM, which
is worth recording because the reasoning is seductive.

The idea: keep the nested `contr.sum` block in the design, form the whole-plot
mean square from it, and substitute that for the dispersion in the Wald
standard error,

$$
\hat\sigma^2_{\mathrm{b},t}
  = \frac{1}{S-2}\sum_s W_{st}\bigl(m_{st} - \bar m_{t,q(s)}\bigr)^2,
\qquad
\mathrm{SE}_j = \sqrt{\hat\sigma^2_{\mathrm{b},t(j)}\,(\mathbf{A}^{-1})_{jj}},
$$

with $m_{st}$ the working-weighted mean residual of cell type $t$ in sample $s$,
centred within condition group. The justification is that
$\mathbb{E}[\hat\sigma^2_{\mathrm{b},t}] \approx \phi + \bar W_t\sigma^2_u$
while $\phi(\mathbf{A}^{-1})_{jj} = \tfrac{4}{S}\phi/\bar W_t$, so their ratio
is exactly the inflation $1 + \bar W_t\sigma^2_u/\phi$ that the contrast needs.

**That step silently assumes $(\mathbf{A}^{-1})_{jj}$ shrinks as $1/\bar W_t$,
so the two $\bar W_t$ terms cancel. They do not.** Measured against the true
sampling distribution of $\hat\beta$ (geometry and gene parameters held fixed
across replicates, so $\mathrm{sd}(\hat\beta)$ is a genuine sampling SD):

```{r stratum_decomp, echo = FALSE}
knitr::kable(
  data.frame(
    `cells per sample` = c(60, 160, 400),
    `inflation required` = c(2.63, 2.59, 2.35),
    `inflation applied` = c(3.63, 5.39, 7.60),
    `null type-I` = c(0.0198, 0.00043, 0.000),
    check.names = FALSE),
  caption = paste("The required inflation is CONSTANT in cells per sample;",
                  "the whole-plot mean square grows as sqrt(Wbar). Nominal",
                  "type-I is 0.05, and the pseudobulk oracle holds 0.042-0.057",
                  "at every one of these points."))
```

Across an 18-cell factorial (varying $S$, cells per sample, and $\sigma_u$, with
a calibrated oracle and a known-broken negative control in every cell) the
construction is calibrated at **exactly one** design point and collapses to zero
rejections as cells per sample grow. The whole-plot mean square is the **wrong
quantity** for a cell-means condition contrast, not a mis-scaled version of the
right one, so there is no implementation fix. Details and the full measurements
are in `research/reports/between-sample-stratum.html`.

Two intermediate findings from that work *were* correct and are reusable: the
stratum must be **cell-type specific** (pooling it to sample level over-inflates
the variance by a factor of $k$), and estimating it from residual aggregates
avoids both a $k(S-2)$ column blow-up — 4,752 columns at 24 cell types and 200
samples — and a double count that afflicts the design-column form.

## The model

We add **patient-level random effects** that soak up the between-patient variability the fixed-effects model wrongly attributes to residual (cell-level) noise. Crucially, we add them only for the *response-related* effects — the only ones we do inference on — by giving each response coefficient its patient-level random counterpart:

$$
\log \mathbb{E}(Y_{c,g}\mid\cdots)
= \underbrace{\gamma_g L_s + \zeta_{g,t}
 + \sum_{t'} N_{c,t',s}\,\alpha_{g,t,t'}}_{\text{fixed, within-sample}}
 + \underbrace{\eta_g\, r_s + r_s\!\sum_{t'} N_{c,t',s}\,\beta_{g,t,t'}}_{\text{fixed response effects (tested)}}
 + \underbrace{u_{g,s} + \sum_{t'} N_{c,t',s}\, v_{g,s,t,t'}}_{\text{patient random effects}},
$$

with

$$
u_{g,s} \sim \mathcal N(0,\tau^2_{\text{int}}),\qquad
v_{g,s,t,t'} \sim \mathcal N(0,\tau^2_{\text{slope}}).
$$

Note the variance components $\tau^2_{\text{int}}, \tau^2_{\text{slope}}$ carry
**no gene index**: a single value of each is shared by every gene (why, and
how that is estimated, is the subject of "The fitting procedure" below). The
realised deviations $u_{g,s}$ do vary by gene — each gene gets its own random
draw — only their common variance is pooled.

-   The random **intercept** $u_{g,s}$ is the patient-level counterpart of the response main effect $\eta_g$: it lets each patient's baseline float, so $\eta_g$ is identified from between-patient contrasts.
-   The random **slopes** $v_{g,s,t,t'}$ are the patient-level counterparts of the $\beta_{g,t,t'}$ terms: they let each patient's niche→expression slope float, so $\beta_{g,t,t'}$ is tested against between-patient *slope* variability. A random intercept alone would not protect $\beta$, because $\beta$ multiplies a covariate that varies *within* a patient.

`random = "intercept"` adds $u_{g,s}$ only; `random = "slope"` adds both. The slope model gives the fullest protection of $\beta$, but note the trade-off: the random slope $v_{g,s,t,t'}$ and the fixed effect $\beta_{g,t,t'}$ both describe how the niche→expression slope differs across patients, so with **few patients per response group** the random slope can absorb a genuine $\beta$ (a real effect seen in only a handful of responders is, statistically, hard to distinguish from patient-to-patient slope noise). With an adequate number of samples prefer `"slope"`; with only a few, `"intercept"` is a more powerful default and still corrects the dominant (baseline) source of pseudo-replication.

## Ridge $\equiv$ random effects: the conceptual bridge

If you know **ridge regression**, you already know 90% of what a random
effect *is* — you have just not called it that. Recall ordinary IRLS for a
GLM: at the current fit $\hat\theta$, form the linear predictor
$\eta_{c,g} = (W\theta)_{c,g}$, the **working response**

$$
z_{c,g} \;=\; \eta_{c,g} + \frac{Y_{c,g} - \mu_{c,g}}{\mu_{c,g}},
$$

(a first-order, log-link linearisation of the count around the current fit)
and the working weight $w_{c,g} = \mu_{c,g}/(1+\phi_g\mu_{c,g})$ from before.
One IRLS step is then an ordinary **weighted least-squares** regression of
$z_{c,g}$ on $W$ with weights $D_g = \mathrm{diag}(w_{c,g})$ — this is exactly
what `SpaNorm::fitNB` already does for `spiDE`'s fixed-effects fit, and it is
the "GLM" part you already know.

Write the expanded design as $\tilde W = [\,W \mid Z\,]$, where $Z$ holds the
random-effect columns: one indicator per sample for $u$, and sample-indicator
$\times$ `CellType:niche` columns for $v$. **A Gaussian random effect on a
group of columns is, on this same working scale, exactly an $L_2$ (ridge)
penalty applied only to those columns** — not an approximation, an identity.
Maximising the penalised weighted least-squares objective

$$
\big\lVert z - \tilde W\theta \big\rVert^2_{D_g} \;+\; \theta^\top \Lambda\, \theta,
\qquad \lVert a \rVert^2_D := a^\top D\, a,
\qquad
\Lambda = \mathrm{blockdiag}\!\left(0_{\text{fixed}},\; \lambda_{\text{int}} I,\; \lambda_{\text{slope}} I\right),
$$

gives the closed-form estimate

$$
\hat\theta = \big(\tilde W^\top D_g \tilde W + \Lambda\big)^{-1}\tilde W^\top D_g\, z,
\qquad\text{with}\qquad
\boxed{\ \lambda_k \;=\; 1/\tau^2_k\ }.
$$

Two equivalent ways to see *why* this identity holds, whichever you find more
natural:

-   **Bayesian / MAP view.** A ridge penalty $\lambda\lVert b\rVert^2$ is
    exactly the log-density (up to a constant) of a Gaussian prior
    $b \sim \mathcal N(0, 1/\lambda)$. Maximising the penalised likelihood is
    maximising a *posterior* under that prior — i.e. assuming $b_s$ are drawn
    from a shared Gaussian, which is the literal definition of a random
    effect. $\lambda = 1/\tau^2$ then just says: a *tighter* prior (small
    $\tau^2$, large $\lambda$) pulls estimates harder toward zero.
-   **Shrinkage view.** Without the penalty, $\hat\theta$ solves ordinary
    weighted least squares and each $\hat u_s$ is a free, unpooled estimate
    (the fixed-factor case that failed above). The penalty term pulls every
    $\hat u_s$ toward $0$ by an amount that grows with $\lambda$ — this pull
    is exactly what "borrowing strength across patients" means numerically,
    and $\hat\theta$ restricted to the random columns is called the model's
    **BLUP** (Best Linear Unbiased Predictor): "unbiased" in the sense of
    minimising mean-squared error *given* $\tau^2$, "predictor" because,
    unlike a fixed effect, $u_s$ is treated as a random draw being predicted,
    not a fixed unknown being estimated.

This is the standard penalised quasi-likelihood (PQL) construction (Breslow &
Clayton, 1993; Schall, 1991) — the same idea `lme4`, `mgcv` and
`MASS::glmmPQL()` all build on. It fits `spiDE` perfectly with **no change to
the fitting engine**: `SpaNorm::fitNB` already accepts a per-column ridge
penalty `lambda.a`, so we set it to $0$ on the fixed columns and $1/\tau^2_k$
on the random ones.

## The fitting procedure: a nested loop

The one piece the ridge equivalence does not hand us for free is $\tau^2$
itself — it is not data, it is exactly what we are trying to learn. `spiDE`
gets it with **two nested loops**, implemented in `.fitNBmixed()`
(`R/fitSpiDE.R`):

-   **Inner loop (existing machinery, unchanged).** For a *fixed* candidate
    $\tau^2$ (hence $\Lambda$), call `SpaNorm::fitNB(Y, W, lambda.a = ...)` —
    the package's ordinary IRLS engine, which already knows how to fit a
    penalised NB-GLM to the whole gene set at once. From `spiDE`'s point of
    view this is a black box: "give me the penalised MLE (the BLUPs
    $\hat u_{g,s}$ included) for this penalty."
-   **Outer loop (the new part).** Given those BLUPs, re-estimate $\tau^2$
    with a **method-of-moments** update, then re-penalise and re-fit. Repeat
    until $\tau^2$ stops moving.

The moment update, for random-effect group $k$ (e.g. `"SampleInt"`) of size
$q_k$ (its number of columns, i.e. the number of samples for a random
intercept):

$$
\hat\tau^2_k \;=\; \frac{\sum_g \lVert \hat b_{g,k}\rVert^2}{G\,\big(q_k - \lambda_k\,\mathrm{tr}\, C_{kk}\big)},
\qquad C = \big(\tilde W^\top \bar D\, \tilde W + \Lambda\big)^{-1}.
$$

New symbols here, defined:

| symbol | meaning |
|---|---|
| $\hat b_{g,k}$ | the fitted BLUPs of block $k$ for gene $g$ (e.g. $\hat u_{g,\cdot}$, the $S$ per-sample intercepts) |
| $G$ | number of genes — **every gene contributes to the same $\hat\tau^2_k$** |
| $\bar D = \mathrm{diag}(\bar w_c)$ | a single, **gene-averaged** working-weight matrix, $\bar w_c = \frac1G\sum_g w_{c,g}$, standing in for each gene's own (slightly different) $D_g$ |
| $C_{kk}$ | the block-$k$ diagonal sub-block of $C$ — the *penalised* covariance of $\hat\theta$ |
| $q_k - \lambda_k\,\mathrm{tr}\,C_{kk}$ | the block's **effective degrees of freedom** — less than $q_k$ because shrinkage means the BLUPs use up less than one full parameter each |

The numerator sums squared BLUPs **over every gene** — this is why the
earlier demo simulates 300 genes rather than a handful: $\tau^2$ is a single
number pooling information across the whole gene set, and pooling over more
genes is what makes that number stable (a single gene's own $\hat\tau^2$ would
be about as noisy as any variance estimated from a small sample — you'll see
this directly below). $\bar D$ exists for the same reason: since $\tau^2$
is shared, the outer loop needs *one* representative information matrix, not
$G$ separate ones.

Two implementation details worth knowing, because they affect what you should
expect to see:

-   **A convergence criterion on $\log\tau^2$.** The loop stops once
    $|\log\hat\tau^2_{\text{new}} - \log\hat\tau^2_{\text{old}}| < \texttt{re.tol}$
    — log scale because $\tau^2$ is a variance (strictly positive, and
    naturally multiplicative in how far off a bad starting guess is).
-   **Iteration cap (`re.maxit`, default `2`).** Lowered from 10 on
    measurement. For `random = "intercept"` a single iteration is
    indistinguishable from ten on null type-I error (to three decimal places)
    and on $\hat\tau^2$ (to two), because the loop converges in a couple of
    steps; the second iteration is kept as cheap insurance. The smaller cap also
    largely dissolves a hazard of the larger one: $\hat\tau^2$ can enter a
    **2-cycle** rather than converging, in which case the value returned depends
    on the *parity* of `re.maxit` — an unpleasant property for a default.

    **This evidence covers the intercept model only.** Under
    `random = "slope"` the slope variance component decays *monotonically*
    across all ten iterations without ever meeting `re.tol`, so a cap of 2
    leaves it orders of magnitude from where 10 leaves it. Pass
    `re.maxit = 10` for slope fits.

-   **Cell subsampling for speed (`re.prop`, default `1`, i.e. off).** The
    *inner* loop's fits can optionally run on a stratified subsample of cells
    (every sample $\times$ cell type stays represented) with a single
    dispersion iteration instead of every cell, trading speed for accuracy in
    the shared scalar $\tau^2$ — never the reported coefficients, since one
    **final fit on all cells with full dispersion** always supplies the
    $\hat\alpha_g$ / $\hat\phi_g$ that inference uses, regardless of
    `re.prop`. That trade is worse than it looks, though: a real-cohort
    replicate study found `tau2` systematically biased *downward* at every
    `re.prop < 1` tested, not just noisier, and averaging replicates does not
    fix it. `re.prop = 1` (no subsampling) is therefore the default; see the
    companion vignette *Speeding up the mixed-effects fit* for the full
    speed/accuracy trade-off.

### Watching it work: a worked example

To make this concrete rather than abstract, we fit the *same* algorithm by
hand on one gene's worth of the toy data above, using `MASS::glmmPQL()` — a
well-tested, textbook implementation of exactly this PQL/Schall construction
(it is not `spiDE`'s own code, but the same algorithm family; think of it as
a second, independent implementation you can cross-check against).

```{r pql_demo, message = FALSE}
library(MASS)
fit_pql <- glmmPQL(y ~ resp, random = ~ 1 | sample_id, family = poisson,
                   data = toy, verbose = FALSE)
summary(fit_pql)$tTable
```

Compare `resp`'s row to the collinear, undefined coefficient a few paragraphs
back: with the response now a random-intercept counterpart rather than a
free-for-all sample factor, it is perfectly estimable, and its reference `DF`
is $S - 2 = 10$ — the between-patient degrees of freedom formula used
throughout this vignette (below).

**The outer loop converging.** `glmmPQL()` exposes its outer-iteration cap as
`niter`; refitting with `niter` $= 1, 2, \dots$ replays the same trajectory
`.fitNBmixed()`'s own loop takes internally.

```{r tau2_trace, fig.small = TRUE, fig.cap = "The outer Schall/PQL loop converging on the patient-level variance component, for one gene. The dashed line is the true simulated value."}
tau2_trace <- vapply(1:8, function(k) {
  f <- suppressWarnings(glmmPQL(y ~ resp, random = ~ 1 | sample_id,
                                family = poisson, data = toy, niter = k,
                                verbose = FALSE))
  as.numeric(nlme::VarCorr(f)[1, 1])
}, numeric(1))
ggplot(data.frame(iteration = seq_along(tau2_trace), tau2 = tau2_trace),
      aes(iteration, tau2)) +
  geom_hline(yintercept = true_tau2, linetype = 2, colour = spide_pal$ref) +
  geom_line(colour = spide_pal$accent) + geom_point(colour = spide_pal$accent) +
  labs(x = "outer (Schall/PQL) iteration",
       y = expression(hat(tau)^2), title = "One gene's variance-component estimate")
```

It converges within a handful of iterations (`.fitNBmixed()`'s default cap is
`re.maxit = 10`) — but notice it does **not** land exactly on the dashed true
value. With only $S = 12$ patients contributing to a single gene's estimate,
$\hat\tau^2$ has real sampling noise, the same way any variance estimated from
12 numbers does. This is precisely why `spiDE` never estimates a per-gene
$\tau^2$: it pools the numerator ($\sum_g\lVert\hat b_{g,k}\rVert^2$) across
every gene being fit. Rerunning the *real* `fitSpiDE()` on the 300-gene `spe`
object from earlier makes the difference concrete:

```{r tau2_pooled}
mixed <- fit_mode("intercept")
mixed$fit$tau2   # pooled across 300 genes -- compare to true_tau2 = 0.49
```

**Do the random intercepts actually look Gaussian?** $\hat\tau^2$ being close
to $0.49$ confirms the model recovers the right *scale* of between-patient
variation, but not that the underlying assumption — $u_{g,s} \sim \mathcal
N(0, \tau^2)$ — genuinely holds. With only $S=12$ or $S=10$ patients, a
histogram of one gene's random intercepts would be far too few points to say
anything about a distribution's shape. What we *can* look at is every
$u_{g,s}$ the simulator actually drew: 300 genes $\times$ 10 samples $=$
3,000 independent draws, kept from `sim_clustered()` (`sim_null$u`) precisely
for this. Alongside them, the *fitted* BLUPs (extracted the same way as the
toy example, but now for all 300 genes and all `SampleInt` columns at once):

```{r gauss_check, fig.small = TRUE, fig.cap = "Empirical density of the true simulated random intercepts and the fitted BLUPs (pooled across all 300 genes and 10 samples), against the theoretical N(0, 0.49) the simulator drew from."}
blup_cols <- which(mixed$fit$re_group == "SampleInt")
blups <- mixed$fit$alpha[, blup_cols, drop = FALSE]
u_df <- data.frame(
  value = c(as.vector(sim_null$u), as.vector(blups)),
  source = rep(c("true simulated u (n = 3,000)", "fitted BLUPs (n = 3,000)"),
              c(length(sim_null$u), length(blups))))
theory <- data.frame(x = seq(-3, 3, length.out = 200))
theory$density <- dnorm(theory$x, 0, sqrt(true_tau2))

ggplot(u_df, aes(value, colour = source)) +
  geom_density(linewidth = 0.9) +
  geom_line(data = theory, aes(x, density), inherit.aes = FALSE,
           linetype = 2, colour = spide_pal$ref, linewidth = 0.8) +
  scale_colour_manual(
    values = c("true simulated u (n = 3,000)" = unname(spide_pal$method[["fixed"]]),
              "fitted BLUPs (n = 3,000)" = unname(spide_pal$method[["intercept"]])),
    name = NULL) +
  labs(x = "per-(gene, sample) random intercept", y = "density",
       title = "The dashed line is the theoretical N(0, 0.49) density")
```

The **true simulated values** (orange) track the dashed theoretical $\mathcal
N(0, 0.49)$ curve closely — unsurprising, since that is literally the
distribution `rnorm(n_genes * n_samples, 0, sd_patient)` draws from, but
useful to see directly rather than take on faith. The **fitted BLUPs**
(green) are visibly *narrower* than both — their empirical variance is
`r round(var(as.vector(blups)), 3)`, below the true $0.49$. This is
shrinkage again, now visible in aggregate
rather than one sample at a time: every individual BLUP is pulled toward
zero, so the *population* of fitted BLUPs is always somewhat narrower than
the population of true effects they estimate. `@tau2` is **not** simply the
variance of the BLUPs — the moment estimator's denominator
($q_k - \lambda_k\,\mathrm{tr}\,C_{kk}$, "The fitting procedure" above)
exists specifically to correct for this narrowing, which is why `@tau2`
lands close to $0.49$ even though `var(BLUPs)` does not.

**Shrinkage.** The other signature behaviour of a random effect is that its
BLUPs are pulled toward zero relative to what each patient's data would say
on its own. We can see this directly by comparing `glmmPQL()`'s BLUPs to the
*fully unpooled* per-sample estimate — the fixed, saturated model
`y ~ sample_id - 1` (one free parameter per sample, no response term, so no
collinearity issue this time).

**How much shrinkage to expect.** Shrinkage is not a fixed amount — it is
strongest exactly when a single patient's own data is *unreliable* relative to
$\tau^2$, and weak when it is not. Concretely, the naive per-sample estimate
has its own sampling variance shrinking like $1/(n_{\text{per}}\cdot\bar\mu)$
(more cells, or a higher mean count, means a more precise per-sample estimate
on its own); the smaller that is *relative to* $\tau^2$, the less the model
needs to borrow strength, and naive and BLUP converge to looking almost
identical. This is why the toy dataset above deliberately uses a modest
`n_per = 12` and a low baseline mean (about 2 counts/cell) rather than, say,
the 90 cells/sample used in `sim_clustered()` — with abundant, high-count
data per sample, shrinkage here would be barely visible at all, which is
itself worth knowing: **a housekeeping-like gene, well measured in every
sample, gets little help from the random effect and only a modest SE
correction from points 1–3 below; a sparse marker or background gene, where
each sample's own count is noisy, is exactly where shrinkage — and the
random-effect correction generally — does the most work.**

A scatter of naive against BLUP would put almost every point close to a
diagonal line no matter how much shrinkage is happening (12 numbers spanning
a wide range makes small, uniform pulls hard to see by eye), so instead we
plot each sample **twice** — its naive value and its BLUP — connected by an
arrow, samples ordered by their naive deviation:

```{r shrinkage, fig.width = 6.5, fig.height = 4, fig.cap = "Each sample's naive (unpooled) deviation and its BLUP from the mixed model, connected by an arrow. Every arrow points toward zero; the sample with the most extreme (least reliable) naive estimate is pulled in the most."}
naive_sat <- glm(y ~ sample_id - 1, family = poisson, data = toy)
naive_u <- coef(naive_sat) - mean(coef(naive_sat))
blup_u <- nlme::ranef(fit_pql)[[1]]
ord <- order(naive_u)
shrink_df <- data.frame(sample = factor(seq_len(S)[ord], levels = seq_len(S)[ord]),
                        naive = naive_u[ord], blup = blup_u[ord])
ggplot(shrink_df) +
  geom_segment(aes(x = sample, xend = sample, y = naive, yend = blup),
              arrow = arrow(length = unit(0.12, "cm")), colour = "grey55") +
  geom_hline(yintercept = 0, linetype = 2, colour = spide_pal$ref) +
  geom_point(aes(sample, naive, colour = "naive (unpooled)"), size = 2.2) +
  geom_point(aes(sample, blup, colour = "BLUP (mixed model)"), size = 2.2) +
  scale_colour_manual(
    values = c("naive (unpooled)" = unname(spide_pal$method[["fixed"]]),
              "BLUP (mixed model)" = unname(spide_pal$method[["intercept"]])),
    name = NULL) +
  labs(x = "sample (sorted by naive deviation)",
       y = "deviation from population mean",
       title = sprintf("var(naive) = %.2f, var(BLUP) = %.2f (%.0f%% reduction)",
                       var(naive_u), var(blup_u),
                       100 * (1 - var(blup_u) / var(naive_u)))) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

Every arrow points toward the dashed zero line — no sample's BLUP moves
*away* from the population mean. The pull is not uniform: the sample with the
largest naive deviation (far left) is pulled in substantially, while samples
already close to zero barely move, because they have little to correct. **How
to read it**: an arrow of length zero would mean the random effect is doing
nothing for that sample (its own data is already fully informative,
effectively $\tau^2 \to \infty$ in relative terms); if every BLUP collapsed
to exactly $0$, the model would be concluding patients don't really differ at
all ($\hat\tau^2 \to 0$). Real data sits in between, and, as the "how much
shrinkage to expect" note above explains, exactly where depends on how
reliable each sample's own naive estimate is relative to $\tau^2$. This is
the concrete, numerical form of "borrowing strength" — and it is what a
per-sample *fixed* effect, by construction, could never do.

## Corrected inference

With the random effects in place, three things change relative to the fixed-effects Wald test — each is a direct consequence of what fitting just did:

1.  **Covariance** uses the full penalised information and extracts the fixed response block, $\widehat{\mathrm{Var}}(\hat\theta_\text{resp}) = \varphi_g \big[(\tilde
    W^\top D_g \tilde W + \Lambda)^{-1}\big]_\text{resp}$. Because $Z$ absorbs the within-patient correlation, this block is inflated to a between-patient scale — the SEs get *larger*, which is the whole point: the fixed-effects fit's SEs were too small.
2.  **Dispersion.** $\varphi_g$ replaces the raw NB dispersion $\phi_g$ with an empirical, *post-fit* working (Pearson) dispersion,
    $$
    \varphi_g = \frac{1}{C - p}\sum_c \frac{(Y_{c,g} - \mu_{c,g})^2}{\mu_{c,g} + \phi_g\mu_{c,g}^2},
    \qquad p = \#\{\text{fixed (non-random) columns}\},
    $$
    i.e. a Pearson-residual-based scale factor, exactly analogous to the
    dispersion `summary(glm(..., family = quasipoisson))` reports — a
    familiar GLM device for saying "trust the assumed variance function, but
    rescale it by how much residual scatter this particular fit actually
    shows," now applied on top of the random-effects model rather than a
    plain GLM.
3.  **Reference distribution** is a $t$, not a normal. The default df
    (`df.method = "between"`) is a **between-patient** $S - 2$ — the same
    bookkeeping as an ordinary two-sample $t$-test (Responder vs. non-Responder)
    at the *patient* level: two group means estimated from the between-patient
    information leave $S-2$ residual df. This is exactly right for the response
    **main** effect $\eta_g$, a pure between-patient contrast. It is *too*
    conservative for the response–niche coefficients $\beta_{g,t,t'}$, which also
    draw on *within*-patient variation in the niche covariate and so carry a
    larger effective sample size — one $S-2$ cannot be right for both.
    `df.method = "satterthwaite"` replaces it with a **per-coefficient** df that
    adapts to each contrast (see *Per-coefficient degrees of freedom* below).

Together these test each response contrast against patient-level variability,
not cell-level variability. Compare the null calibration of the two fits (the
`fit_mode()` helper from earlier already fit both):

```{r mixed}
c(fixed = mean(fixed$p < 0.05), mixed = mean(mixed$p < 0.05))
```

```{r qq, fig.small = TRUE, fig.cap = "Null response-effect p-values. The fixed-effects fit is grossly anti-conservative; the mixed-effects fit is close to the uniform (dashed) reference."}
qq_points <- function(p, mode) {
  data.frame(mode = mode, expected = sort(-log10(ppoints(length(p)))),
             observed = sort(-log10(p)))
}
qq_df <- rbind(qq_points(fixed$p, "fixed"), qq_points(mixed$p, "intercept"))
qq_df$mode <- factor(qq_df$mode, levels = c("fixed", "intercept"),
                     labels = c("fixed", "mixed"))
ggplot(qq_df, aes(expected, observed, colour = mode)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = spide_pal$ref) +
  geom_point(size = 1.5) +
  scale_colour_manual(values = unname(spide_pal$method[c("fixed",
                                                         "intercept")]),
                      name = NULL) +
  labs(x = expression(-log[10]("expected p")),
       y = expression(-log[10]("observed p")))
```

## Per-coefficient degrees of freedom (Satterthwaite)

The single $S-2$ reference df is right for the response *main* effect but wrong
for the response–niche interactions, and it cannot be right for both. $\eta_g$
is a pure between-patient contrast (the condition is a patient-level label), so
its effective df genuinely is $\approx S-2$. Each $\beta_{g,t,t'}$, by contrast,
multiplies a niche covariate that varies *within* a patient, so it is informed
by far more than $S$ independent pieces of information — on real data the
per-coefficient effective sample size can span three orders of magnitude across
the tested columns. Referencing all of them against $S-2$ leaves the interaction
tests badly over-conservative.

`df.method = "satterthwaite"` (on `fitSpiDE()` / `spiDE()`) gives each tested
coefficient $j$ its own **Satterthwaite** degrees of freedom,
$$
\nu_j \;=\; \frac{2\,\big(\widehat{\mathrm{Var}}\,\hat\theta_j\big)^2}
                 {\mathrm{Var}\!\big(\widehat{\mathrm{Var}}\,\hat\theta_j\big)},
$$
the classical "how many independent variances is this SE really worth" ratio —
the same device `lmerTest` / `pbkrtest` use for small-sample linear mixed
models, here applied to the PQL working model. The denominator propagates (by
the delta method) the uncertainty in the *variance parameters* $\theta =
(\varphi,\ \tau^2_\text{int},\ \tau^2_\text{slope})$ through the SE, using their
covariance $\widehat{\mathrm{Cov}}(\hat\theta)$ — a small
$(1+\#\text{groups})$-square matrix from the reduced-form REML information of the
working model, built entirely from the $p\times p$ penalised inverse $M^{-1} =
(\tilde W^\top \bar D\,\tilde W + \Lambda)^{-1}$ (never an $n\times n$ matrix).

Two facts make this both cheap and safe:

-   **It is invariant to the per-gene dispersion $\varphi_g$.** The SE and every
    term of its variance scale identically in $\varphi$, so $\varphi$ cancels in
    the ratio $\nu_j$. What remains depends only on the *shared* working weights
    and variance components, so a single per-coefficient df vector — computed
    once per bandwidth from the representative penalised inverse — serves every
    gene, exactly as one shared $\tau^2$ does. Nothing is added to the per-gene
    inference loop.
-   **The main effect is preserved, by construction.** For the between-patient
    contrast $\eta_g$, $\widehat{\mathrm{Var}}\,\hat\eta_g$ is dominated by the
    $\tau^2_\text{int}$ direction, whose uncertainty is governed by the $\approx
    S$ patients — so $\nu$ returns to $\approx S-2$. The $S-2$ df is thus the
    *special case for the main effect*, not a global constant; the niche
    interactions move toward the (much larger) residual df.

Both are visible on the toy fit ($S=10$ patients, so $S-2=8$):

```{r satterthwaite, message = FALSE}
fs <- fitSpiDE(spe, "condition", sigma = 30, random = "intercept",
               df.method = "satterthwaite", verbose = FALSE)
ff <- fits(fs)[[1]]
ct <- as.character(ff$covtype)
resp  <- ff$coefmap$covariate[ct == "ResponseCellType"]
niche <- ff$coefmap$covariate[ct == "ResponseNiche"]
c(ResponseCellType.median = round(median(ff$df[resp]), 1),
  ResponseNiche.median = round(median(ff$df[niche]), 1))
```

The `CellType:condition` df sits at $\approx S-2 = 8$ — these are
between-patient contrasts, so they are left alone — while the niche-interaction
df is much larger, being the coefficients the flat $S-2$ was over-penalising.
Under the default `df.method = "satterthwaite"`, `@df` is therefore a **named
vector**, one entry per tested (`ResponseCellType` / `ResponseNiche`) column,
aligned to the columns of `@t_stat` / `@se`; passing `df.method = "between"`
restores the scalar $S-2$.

**If you consume `@df` directly**, note it is a *vector* by default. Index the
column you are testing — `fit@df[["<covariate>"]]` — rather than passing the
whole vector to `stats::pt()`, which would silently recycle it against your
per-gene statistics rather than erroring.

## The testing framework

How do we know the per-coefficient df is *calibrated* rather than merely
plausible? The check is a **null-simulation type-I study**: simulate data with
genuine between-patient clustering but *no* response effect ($\beta = 0$), fit
each `random =` / `df.method` combination, and record the fraction of response
tests significant at $\alpha = 0.05$, which should be $\approx 0.05$ for an
honest test. Two properties matter, and they pull in opposite directions:

-   **The main effect must stay $\approx$ nominal.** It calibrates close to
    nominal under `"between"`, and — since $\nu \approx S-2$ there by
    construction — `"satterthwaite"` must *not* move it. This is the
    load-bearing regression: whatever the df change does to the niche
    coefficients, it must leave the main effect where it was.
-   **The response–niche tests should approach nominal from below.** Under
    `"between"` they are over-conservative (type-I well under $0.05$); the
    larger `"satterthwaite"` df should raise them *toward* $0.05$. Were they to
    overshoot, the residual miscalibration would live in the covariance itself,
    and the principled next step would be a Kenward–Roger bias-correction of
    $\widehat{\mathrm{Cov}}(\hat\theta_\text{resp})$ — a heavier, second-order
    correction not enabled here.

### What the study found

The study has since been run at realistic scale (thousands of cells over tens of
patients, 40 replicates per design point) in the package's benchmark harness
(`research/`), sweeping both `df.method` arms on identically seeded data. The
full figures are in the *Simulating realistic spatial data and benchmarking
spiDE* vignette; the result matters enough to state here.

The first criterion is met and the second is met **only partially**, which is
worth being explicit about rather than glossing:

-   `"between"` is not merely conservative, it is *severely* so when patients
    are few: response-niche type-I error is $\approx 0.001$ at $S=4$ against a
    nominal $0.05$ — a 50-fold undershoot — climbing to only $\approx
    0.046$–$0.055$ by $S=30$. Power tracks it: recall is essentially zero at
    $S \le 10$ because nothing survives genome-wide correction.
-   `"satterthwaite"` holds type-I error in $0.042$–$0.065$ across the *whole*
    sampled range, and buys real power for it: a mean paired gain of
    $\approx +0.10$ TPR at 5% FDR, with observed FDP no worse.
-   But it does **mildly overshoot** at larger $S$ — worst measured
    $\approx 0.065$ against nominal $0.05$, a $\approx 30\%$ relative
    inflation. By the criterion set out above, that residual bias is located in
    the covariance, and **Kenward–Roger is therefore the indicated next step**,
    not merely a hypothetical one.

The trade is strongly favourable — a bounded $\approx 30\%$ liberal drift in
exchange for removing a 50-fold conservative one, plus the power that comes with
it — which is why `"satterthwaite"` is now the default. `"between"` remains
available for back-compatibility, or where strict conservatism is worth more
than power. On a small toy fixture the response–niche null is too noisy to read
as a calibration; the main-effect preservation shown above is the part that is
robust at toy scale.

## Interpreting a mixed `SpiDEFit`

A fit made with `random != "none"` populates three extra slots on each
`SpiDEFit` (`fits(res)[[1]]`, or `$` directly), all `NULL` for a plain
fixed-effects fit:

| slot | what it holds | how to read it |
|---|---|---|
| `@tau2` | the fitted variance component(s), named `SampleInt` (and `SampleSlope` under `random = "slope"`) | on the **log-mean scale**: a patient with $u_s = +\hat\tau$ has roughly $e^{\hat\tau}$-fold higher baseline expression than average. If `@tau2` is tiny, patients barely differ beyond sampling noise (the mixed fit will look similar to, but always slightly more conservative than, the fixed fit); if it is large relative to the within-patient (Poisson/NB) noise, a sizeable share of what a fixed-effects fit reported as "response signal" was actually between-patient noise leaking through |
| `@df` | the Wald reference degrees of freedom: a **named per-coefficient vector** under `df.method = "satterthwaite"` (the default; `ResponseCellType` $\approx S-2$, niche interactions larger), or a scalar between-patient $S - 2$ under `df.method = "between"` | small df (few patients) means a wide-tailed $t$ and hence conservative p-values *even for a real effect* — expected, not a bug; the default `"satterthwaite"` relieves this for the within-patient-informed niche coefficients while leaving the between-patient main effect untouched. Index it (`@df[[col]]`) rather than passing the vector to `pt()` |
| `@penalty` | the per-column ridge penalty actually used ($0$ on fixed columns, $1/\hat\tau^2_k$ on random ones) | mostly a diagnostic: a very small penalty on a random block means that group's $\hat\tau^2$ converged large, i.e. that source of between-patient variation is real and substantial |

For the toy fit above, `mixed$fit$tau2` pooled across 300 genes should sit
close to the simulated $0.49$. `mixed` pins `df.method = "between"` (see the
`fit_mode()` chunk), so `mixed$fit$df` is the flat scalar $S-2 = 8$ ($S=10$
patients). Under the package default `"satterthwaite"` it would instead be a
per-coefficient vector — as `fs` in the previous section shows — whose
`ResponseCellType` entries sit at $\approx 8$ while the niche interactions
carry more.

## Usage and caveats

```{r usage, eval = FALSE}
# random intercept + random niche slopes (fullest correction; use with an
# adequate number of samples). For few samples, random = "intercept" is safer.
res <- spiDE(spe, condition = "condition", sigma = c(10, 30, 50),
             random = "slope")
```

-   **Sample-level covariates.** A per-sample random intercept already adjusts for *all* between-sample nuisance variation, so it is confounded with measured sample-level covariates (Age, Sex, Stage) — exactly the collinearity demonstrated at the top of this section. `spiDE` therefore errors if such a covariate is passed together with `random != "none"` — drop it; the random intercept subsumes it.
-   **Number of samples.** The correction is only as good as the between-sample information: with very few samples the variance components (and the $S-2$ reference df) are unstable, and a warning is issued.
-   **Correct, not merely "fixed with extras".** The mixed fit does more than add a between-patient variance component: it also replaces `random = "none"`'s NB dispersion scaling and sub-design covariance with the correct **working (quasi-NB) dispersion** and **full penalised covariance**. So even when $\tau^2 \to 0$ (no genuine clustering) the mixed fit does *not* reproduce the `"none"` fit — it gives the correctly-scaled, and hence more conservative, standard errors. This is deliberate: `"none"` is known to be anti-conservative, so its larger apparent "power" is partly spurious. It is for this reason that `"intercept"` is now the default.
-   **Power needs samples and effect size.** Because the test is now honest, a real response-niche effect must be visible *between patients* to be declared. With few patients, or an effect seen in only a handful of responders, the between-patient df and the inflated SEs will (correctly) leave it non-significant. `df.method = "satterthwaite"` recovers legitimate power for niche coefficients that carry within-patient information — giving them their larger effective df instead of the flat $S-2$ — without touching the honest between-patient df of the main effect; larger sample sizes restore power further.

# Session info

```{r sessioninfo}
sessionInfo()
```
