---
title: "Speeding up the mixed-effects fit: the PQL subsampling trade-off"
author: "Dharmesh D. Bhuva"
package: spiDE
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Speeding up the mixed-effects fit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE, collapse = TRUE, comment = "#>", message = FALSE,
  warning = FALSE, fig.width = 7, fig.height = 4, dpi = 150, fig.retina = 2
)
```

```{r libs}
library(spiDE)
library(ggplot2)
source(system.file("spide_palette.R", package = "spiDE"))

bench <- function(f) {
  path <- system.file("extdata", "benchmark", f, package = "spiDE")
  # Large tables ship xz-compressed to stay inside Bioconductor's 5 MB package
  # cap (see research/R/install_results.R), so a requested .csv may exist only
  # as .rds. Fall back rather than returning NULL, which would blank the panel
  # instead of failing.
  if (!nzchar(path) && grepl("\\.csv$", f)) {
    path <- system.file("extdata", "benchmark", sub("\\.csv$", ".rds", f),
                        package = "spiDE")
  }
  if (!nzchar(path) || !file.exists(path)) return(NULL)
  if (grepl("\\.rds$", path)) readRDS(path) else
    utils::read.csv(path, check.names = FALSE)
}
```

# Overview

The mixed-effects fit (`fitSpiDE(..., random = "intercept" | "slope")`) corrects
the anti-conservative inference that cell-level pseudo-replication causes (see
*The spiDE model* vignette). It does so with a Schall/PQL loop that re-estimates
the shared random-effect variance components `tau2`, re-fitting the
negative-binomial GLM once per iteration — the expensive part of the fit.

Those variance components are a *single shared scalar per random-effect group* — a
population quantity that does not need every cell to estimate, exactly as
`SpaNorm::fitNB` already subsamples cells for its dispersion step. `spiDE` exposes
three knobs that can trade `tau2` accuracy for speed, of which only `re.prop` is
usually worth touching:

| argument       | default | effect |
|----------------|---------|--------|
| `re.prop`      | `1`     | fraction of cells sampled per cell type within each sample for the PQL loop; `1` disables subsampling |
| `re.maxit.psi` | `1`     | dispersion iterations inside each PQL step |
| `re.min.cells` | `100`   | per-stratum floor for `re.prop` |

Crucially the loop only supplies `tau2`; a single **final fit on all cells with
full dispersion** produces the coefficients and dispersion that inference uses. So
subsampling perturbs only the shared variance estimate, not the per-gene effect
sizes — but, as this vignette shows, "only the shared variance estimate" is not
the same as "safely, since it's just one number": `tau2` itself is what turns out
to be systematically biased by subsampling, not merely noisier. This vignette
quantifies the time saved and the accuracy cost, using two studies: a
single-run simulation benchmark (`research/`, shipped as this vignette's
figures) and a more rigorous real-cohort replicate study (5 refits per
`re.prop` level on the full 55-sample YTMA cohort) that changed the package
default, described below.

# Time versus accuracy across `re.prop`

The study fits the same realistic dataset at several values of `re.prop`, timing
each fit and correlating its response-niche t-statistics against the fully
accurate reference (`re.prop = 1`, all cells). Two figures tell the story:
subsampling makes the fit markedly faster while the t-statistics it feeds into
inference stay highly correlated with the full-data fit.

```{r pql, fig.height = 4, fig.cap = "Left: the mixed-effects fit is markedly faster once cells are subsampled for the PQL loop (about 1.8x versus the all-cell fit), after which it plateaus. Right: response-niche t-statistics stay highly correlated with the full-data reference (~0.96) across the sampling range."}
timing <- bench("pql_timing.csv")
acc <- bench("pql_accuracy.csv")

if (!is.null(timing)) {
  p_time <- ggplot(timing, aes(re.prop, elapsed)) +
    geom_line(colour = spide_pal$ref) +
    geom_point(size = 3, colour = spide_pal$accent) +
    scale_x_continuous(trans = "reverse") +
    labs(x = "re.prop (cell fraction, reversed)", y = "fit time (s)",
         title = "Fit time versus subsampling")

  if (!is.null(acc)) {
    # anchor the reference (re.prop = 1) at correlation 1
    acc <- rbind(acc, data.frame(re.prop = 1, cor = 1))
    p_acc <- ggplot(acc, aes(re.prop, cor)) +
      geom_line(colour = spide_pal$ref) +
      geom_point(size = 3, colour = spide_pal$accent) +
      scale_x_continuous(trans = "reverse") +
      coord_cartesian(ylim = c(min(0.9, min(acc$cor)), 1.001)) +
      labs(x = "re.prop (cell fraction, reversed)",
           y = "t-stat correlation vs full-data fit",
           title = "Accuracy versus subsampling")
    print(patchwork::wrap_plots(p_time, p_acc))
  } else {
    print(p_time)
  }
}
```

The coefficients come from the full final fit regardless of `re.prop`, so the
response-niche t-statistics stay highly correlated with the full-data reference
even under aggressive subsampling. This single-run simulation, taken on its
own, reads as a favourable trade: the subsample here underestimates `tau2`
(about 0.19 versus 0.25 at full data), and that feeds through to only a small
drop in t-statistic correlation — to roughly 0.96 rather than 1.

That reading turned out to be incomplete, which is why `re.prop = 1` (not
`0.2`) is now the package default. A single simulated timing/accuracy point
can't distinguish *noise* (would average out over replicates) from *bias*
(wouldn't). A follow-up study on the real 55-sample YTMA cohort — 5 independent
refits of identical data at each of `re.prop` = 0.2, 0.5, and 0.8, differing
only in the unseeded stratified cell subsample, plus the deterministic
`re.prop = 1` reference — found both, and the bias matters more:

* **The subsampling noise itself does not shrink** between `re.prop = 0.2` and
  `0.8` (replicate sd stayed in the 0.015-0.017 range throughout), and at every
  level tested it matched or exceeded the genuine between-patient signal (sd
  0.0145 from a 55-fold leave-one-sample-out jackknife) that `tau2` is meant to
  capture.
* **Mean `tau2` rose monotonically with `re.prop`** (0.034 -> 0.052 -> 0.059 ->
  0.074 at `re.prop` = 0.2/0.5/0.8/1) — systematic downward attenuation at every
  `re.prop < 1`, not random jitter around the true value. Averaging several
  `re.prop < 1` replicates would not fix this, because their common mean is
  itself biased low.

So a small drop in t-statistic *correlation* against one particular full-data
run (as in the simulation above) is the wrong lens: it can look reassuring even
while the underlying `tau2` is being systematically underestimated. GPU
backends make `re.prop = 1` affordable in practice (about 1.17x the fit time of
a heavily-subsampled fit, per the LOO study), so it is now the default; treat
`re.prop < 1` as a lower bound on `tau2`, not a point estimate, if used at all.

```{r tau2-table, echo = FALSE}
if (!is.null(timing)) {
  knitr::kable(timing[order(-timing$re.prop),
                      c("re.prop", "elapsed", "tau2")],
               digits = 3, row.names = FALSE,
               col.names = c("re.prop", "fit time (s)",
                             "tau2 (intercept)"),
               caption = "Fit time and the estimated intercept variance component across re.prop.")
}
```

# Scaling to realistic gene counts

A realistic dataset has ten-thousand-plus genes. Because the whole gene set is
fit in a single call (dispersion is moderated across all genes together), the
per-fit cost grows with both the number of genes and the number of cells. The
study records the fit and inference time as the number of cells per sample grows
at fixed gene count.

```{r timing, fig.height = 4, fig.cap = "Fit and inference wall-clock time as the number of cells grows, at a realistic gene count. Inference is the blockable, parallelisable stage."}
tm <- bench("timing.csv")
if (!is.null(tm) && nrow(tm) > 0) {
  long <- rbind(
    data.frame(n_cells = tm$n_cells, stage = "fit", seconds = tm$fit_seconds),
    data.frame(n_cells = tm$n_cells, stage = "inference",
               seconds = tm$test_seconds))
  ggplot(long, aes(n_cells, seconds, colour = stage)) +
    geom_line(linewidth = 0.8) + geom_point(size = 2) +
    scale_colour_manual(values = unname(spide_pal$method[c(1, 2)])) +
    labs(x = "total cells", y = "wall-clock time (s)", colour = NULL,
         title = sprintf("Scaling at ~%s genes",
                         format(round(mean(tm$n_genes), -2), big.mark = ",")))
} else {
  cat("*(Timing summary not found; run the benchmark study in `research/`.)*")
}
```

# Null calibration and the reference degrees of freedom

The mixed-effects fit corrects the fixed-effects anti-conservatism, but the
*reference degrees of freedom* it then uses decides how conservative the
corrected response-niche test is: under the default between-patient reference
(`df.method = "between"`), a scalar $S-2$ is markedly **over-conservative** at
small-to-moderate $S$, while `df.method = "satterthwaite"` (a per-coefficient
approximation) holds close to nominal across the whole sampled range and
recovers real power at small $S$ that the between reference cannot deliver at
all — without trading away calibration to get it (satterthwaite's worst-case
drift is a mild, bounded overshoot, not the severe distortion between shows in
the opposite direction). See the null-calibration and power-gain figures in
*Simulating realistic spatial data and benchmarking spiDE* for the full
between-vs-Satterthwaite comparison and the numbers behind this. The residual
liberal drift Satterthwaite shows at large $S$ is exactly what the
Kenward–Roger covariance correction (see *The spiDE model*) would target next,
if the residual bias turns out to live there.


# Model design: does `CellType:condition` change calibration or power?

The niche-only niche design is

```
~ 0 + covariates + CellType + condition * CellType:(niches) + niches
```

which expands to `condition` + `CellType:niches` + `condition:CellType:niches`
but carries **no `CellType:condition`** term. A cell-type-specific response
*intercept* shift therefore has nowhere to go: the single global `condition`
column captures only the average, and the per-sample random intercept absorbs a
per-sample scalar. Because the niche covariates are not centred, the tested
three-way coefficient can and does absorb the missing shift, scaled by the mean
niche density.

The alternate design adds that term, cell-means coded (the `condition` main
effect is dropped, so there is one coefficient per cell type, each the
tested-level shift *within* that cell type, readable without forming a
contrast). The benchmark runs both as a `design` arm, over the families that can
discriminate between them — null calibration, power, and the combiner
comparison. `pql`, `timing` and `spatial` probe the random-effect machinery and
the layouts, which the design change does not touch, so they run once.

```{r design-cal, fig.height = 4, fig.cap = "Null type-I error of the response-niche test by sample size, for the niche-only design and the design with CellType:condition restored (gradient layout, satterthwaite df). Restoring the term costs ~k degrees of freedom out of several hundred, so calibration should be unchanged; a visible gap would mean the added block is absorbing signal rather than nuisance."}
cal <- bench("null_type1.csv")
if (!is.null(cal) && !is.null(cal$design) &&
      length(unique(cal$design)) > 1) {
  d <- cal[cal$layout == "gradient", , drop = FALSE]
  if (!is.null(d$df.method)) d <- d[d$df.method == "satterthwaite", , drop = FALSE]
  d$arm <- interaction(d$method, d$design, drop = TRUE, sep = " / ")
  ggplot(d, aes(S, type1, colour = arm)) +
    geom_hline(yintercept = 0.05, linetype = 2, colour = spide_pal$ref) +
    geom_line(linewidth = 0.7) + geom_point(size = 2) +
    labs(x = "samples (S)", y = "response-niche type-I error", colour = NULL,
         title = "Null calibration by model design")
} else {
  cat("*(No `design` arm in the shipped summaries yet. Run the alternate arm:",
      "`SPIDE_DESIGN_ARM=celltype-response SPIDE_DESIGN_LIB=<lib>",
      "bash research/slurm/submit.sh`, then re-aggregate and reinstall.)*")
}
```

```{r design-power, fig.height = 4, fig.cap = "True-positive rate at FDR 0.05 versus effect size, by model design. The added term spends degrees of freedom, so a small power cost is expected; a large one would argue against restoring it."}
pw <- bench("power_fdr.csv")
if (!is.null(pw) && !is.null(pw$design) && length(unique(pw$design)) > 1) {
  d <- pw[pw$fdr == 0.05 & pw$sweep == "effect", , drop = FALSE]
  if (!is.null(d$df.method)) d <- d[d$df.method == "satterthwaite", , drop = FALSE]
  d$arm <- interaction(d$method, d$design, drop = TRUE, sep = " / ")
  ggplot(d, aes(beta, tpr, colour = arm)) +
    geom_line(linewidth = 0.7) + geom_point(size = 2) +
    labs(x = expression(paste("planted effect ", beta)),
         y = "TPR at FDR 0.05", colour = NULL,
         title = "Power by model design")
} else {
  cat("*(No `design` arm in the shipped power summaries yet.)*")
}
```

**What the arm actually varies.** Because `design` selects a whole package
build, the alternate arm bundles two changes that ship together: the added
`CellType:condition` block, *and* two-sided p-value combination under
`combine = "cauchy"` (Brown keeps one-sided input). Differences between the arms
in the figures above are therefore attributable to the pair, not to the design
term alone. That is deliberate — the two are intended to ship together — but it
means these panels answer "is the new build calibrated and no less powerful?"
rather than "what does `CellType:condition` do in isolation".

Sidedness is characterised on its own in
`vignette("spiDE-cauchy-vs-brown")`, which derives why one-sided ACAT cancels
on genes with opposing niche effects and shows it costs nothing where the
one-sided convention was already safe. The combiner table below carries both
`combine` and `design`, so the combiner axis can still be read within an arm.

## The scenario the niche-only design cannot express

The two panels above are measured on data whose ground truth is **purely
niche-driven**: `simulate_dataset()` plants `beta * niche_potential` in index
cells of Responders and nothing else. On such data the added
`CellType:condition` term has no real effect to capture, so it can only spend
degrees of freedom. A power cost there is expected and says nothing about
whether the term is worth having.

The `ctresp` scenario supplies the other half. It plants a **cell-type-specific,
niche-INDEPENDENT** response --- a flat shift in index cells of Responders, with
`beta = 0` so there is no neighbourhood effect at all --- on a gene block
disjoint from the niche-effect genes. Every niche-layer call on those genes is
therefore a false positive by construction.

Unlike the panels above this comparison **is** deconfounded: it pins
`combine = "brown"`, which is one-sided in both builds, and forwards
`df.method` explicitly, so the arms differ in the design term alone.

```{r design-ctresp, fig.height = 4, fig.cap = "Spurious niche calls on genes carrying a cell-type-specific response and no niche effect. Every call plotted is a false positive. Under the niche-only design the shift has nowhere to go but the three-way term and surfaces there as fabricated neighbourhood-dependence."}
ct <- bench("ctresp.csv")
if (!is.null(ct) && !is.null(ct$design) && length(unique(ct$design)) > 1) {
  ggplot(ct, aes(fdr, spurious, colour = design)) +
    geom_line(linewidth = 0.7) + geom_point(size = 2) +
    facet_wrap(~ paste("beta_ct =", beta_ct)) +
    labs(x = "nominal FDR", y = "fraction of planted genes called niche-dependent",
         colour = NULL, title = "Spurious niche-dependence by model design")
} else {
  cat("*(No `ctresp` results in the shipped summaries yet.)*")
}
```

The niche-only design produces spurious niche calls at every effect size, and its
rate climbs steeply with the FDR threshold --- at `beta_ct = 1` it rises from
8.3e-05 at FDR 0.01 to 1.6e-03 at FDR 0.20, roughly **9.5x** the
celltype-response arm, whose rate is nearly flat. The leaked signal sits just
below the gate and floods in as the gate loosens. At the weakest effect
(`beta_ct = 0.5`) the promoted design produces **no** spurious calls at any
threshold while the niche-only design produces them from FDR 0.05 upward.

**The honest reading is a trade, not a win.** Restoring `CellType:condition`
costs power when the truth is purely niche-driven, and prevents fabricated
niche-dependence when a cell-type-wide response shift exists. Which dominates is
an empirical question about the dataset: a cohort with no cell-type-level
response loses by the change, one with such a shift gains. Null calibration is
unaffected either way (median type-I 0.0450 against 0.0449 across 50 design
points per arm), so the choice is about power and specificity, not validity.

# Gene sets: why `spiGSEA()` defaults to the competitive test

`spiGSEA()` offers two nulls. The **self-contained** test asks whether a set's
mean statistic differs from zero; the **competitive** test asks whether it
differs from the genes outside the set, as `limma::camera` does. They answer
different questions, and the benchmark decides which is safe to default to.

The simulation plants nothing: 30 samples of 300 cells, 1,500 genes, and
gene-gene correlation induced at the rate seen in real spatial data
(rho ~ 0.011, calibrated against the *realised* residual correlation rather
than assumed). Sets are of three kinds -- `null` (random genes), `high`
(drawn from the top expression quintile) and `plant` (which carry signal only
in the power scenarios). With no signal anywhere, every call is a false one.

```{r gsea-null, fig.height = 3.6, fig.cap = "False call rate with nothing planted, at nominal FDR 0.05 and realistic inter-gene correlation. Every call shown is false. The self-contained test does not control error, and its errors concentrate in the high-expression sets."}
cl <- bench("gsea_calls.csv")
if (!is.null(cl)) {
  k <- cl[cl$arm == "null" & cl$fdr == 0.05 & cl$rho_mode == "estimated" &
            cl$load == 0.3 & cl$libsd == 0.2 &
            cl$set_class %in% c("null", "high"), ]
  g <- aggregate(call_rate ~ test + set_class, data = k, FUN = mean)
  g$set_class <- factor(g$set_class, c("null", "high"),
                        c("random sets", "high-expression sets"))
  ggplot(g, aes(set_class, call_rate, fill = test)) +
    geom_col(position = position_dodge(0.7), width = 0.6) +
    geom_hline(yintercept = 0.05, linetype = 2) +
    labs(x = NULL, y = "fraction of sets called (all false)", fill = NULL)
}
```

The dashed line is the nominal 5%. The competitive test sits at zero. The
self-contained test calls about 2.7% of random sets and 17.6% of
high-expression sets -- roughly 20 of 208 sets per replicate, all false, so its
realised false-discovery proportion is **1.00 against a reported 0.05**.

It is worth being precise about the cause, because the obvious suspect is
wrong. Setting `rho = 0` roughly quadruples the damage, so the
correlation term is doing real work -- but the self-contained test remains
badly anti-conservative *with* the correct `rho`. The difference is that it
assumes the gene-level statistics being averaged have unit spread. That holds
under the null and fails under signal (measured spread 1.0 null, 1.8 at the
largest effect). The competitive form divides by the observed spread, so it is
immune. With genuinely independent genes -- which requires switching off
per-cell library-size variation as well as the latent factor, since library
size alone leaves rho ~ 0.0036 -- **both** tests are calibrated, which is what
identifies correlation rather than the model or the FDR cascade as the cause.

Two consequences follow. Under a global shift affecting most genes, the
self-contained test called 96.3% of random sets against 0.7% for competitive.
And because the per-gene statistics track expression, its false calls land on
abundant sets: called sets sat at the 86th expression percentile against the
51st for random sets, closely reproducing what is seen on real data.

The competitive test is not bought with unusable sensitivity: at FDR 0.05 its
power is usable from about 25 genes per set at moderate effects and saturates
by 50 genes at larger ones.

# Summary

* Subsampling cells for the PQL loop (`re.prop < 1`) plus a single inner
  dispersion iteration (`re.maxit.psi = 1`) cuts the mixed-effects fit time
  substantially, while the final all-cell fit always keeps the coefficients
  exact — only the shared `tau2` is affected.
* That effect on `tau2` is not just noise: a real-cohort replicate study found
  it systematically biased downward at every `re.prop < 1` tested, an
  attenuation that gets worse (not better) the more the correlation-based
  simulation check above might suggest is "fine". Response-niche t-statistic
  correlation against a full-data fit is a poor proxy for this, since it can
  look reassuring even while `tau2` itself is off.
* `re.prop = 1` (no subsampling, fully reproducible, no seed required) is
  therefore the package default; GPU backends make it affordable. Lower it
  only when `tau2`'s absolute scale doesn't matter (e.g. a quick feasibility
  check), and set an external `set.seed()` if reproducibility of a
  `re.prop < 1` run matters.

Timing numbers reflect the machine that produced the benchmark tables; regenerate
them for your own hardware with the study in `research/`.

```{r session}
sessionInfo()
```
