---
title: "Is this fit calibrated? Reading lambda per cell type"
author: "Dharmesh D. Bhuva"
package: spiDE
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Calibration diagnostics: lambda per cell type}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, collapse = TRUE, comment = "#>", message = FALSE,
  warning = FALSE, fig.width = 7, fig.height = 4.2, dpi = 150, fig.retina = 2
)
set.seed(1)
```

```{r libs}
library(spiDE)
library(ggplot2)
source(system.file("spide_palette.R", package = "spiDE"))
```

# Overview

A `spiDE` fit tests thousands of genes against hundreds of
`celltype:condition:niche` coefficients, and those coefficients are **not
equally well estimated**. A tumour compartment with tens of thousands of cells
spread over every patient supports a very different quality of inference from a
dendritic-cell compartment with a few hundred cells concentrated in a handful of
patients -- yet both are reported in the same table, against the same FDR
threshold.

This vignette is about detecting that, using the genomic inflation factor
$\lambda$, and about what to do once you have. The short version:

- $\lambda$ measures whether your test statistics follow the null distribution
  you assumed. It is a property of the **test**, not of the biology.
- In `spiDE` it varies *within a single fit*, across cell types, because the
  cell types differ in how many patients contribute cells.
- $\lambda < 1$ costs **power, not validity**. A null result for a poorly
  calibrated cell type is weak evidence of absence.
- The right response is to restrict which coefficients you test, **not** to
  rescale the statistics and **not** to change the model.

# What lambda is

For a one-degree-of-freedom test, the squared statistic $t^2$ is a $\chi^2_1$
variate under the null. The genomic inflation factor compares the median you
observed with the median the null predicts:

$$\lambda = \frac{\mathrm{median}(t^2)}{\mathrm{median}(\chi^2_1)}
          = \frac{\mathrm{median}(t^2)}{0.4549}$$

The **median** is the important choice. A minority of true positives, however
large their statistics, barely moves it, so $\lambda$ reads the bulk of the
distribution -- which is overwhelmingly nulls. That is the whole design of the
statistic: it estimates whether your *null* is the null you assumed. It comes
from genome-wide association studies, where population stratification inflates
every association statistic at once (Devlin and Roeder, 1999).

| $\lambda$ | reading |
|---|---|
| $= 1$ | observed null matches the theoretical null; p-values mean what they say |
| $> 1$ | statistics too large: confounding or anti-conservatism, **or** genuinely widespread signal |
| $< 1$ | statistics too small: the test is conservative, p-values are too big |

```{r lambda-fn}
# lambda from any matrix or vector of t-statistics
lambda <- function(t) stats::median(t^2, na.rm = TRUE) / stats::qchisq(0.5, 1)
```

# Lambda below 1 does not mean "there is no signal"

This is the point most often got wrong, so it is worth demonstrating rather than
asserting. Under a pure null with a correctly calibrated test, $\lambda = 1$
**exactly**. Absence of biology gives you 1, not less than 1.

```{r pure-null}
z <- matrix(rnorm(200 * 50), 200, 50)     # pure null, perfectly calibrated
round(lambda(z), 3)
```

So $\lambda = 0.68$ is never a statement that a cell type is quiescent. It says
the reference distribution is wrong for the statistic's actual sampling
behaviour -- here, that the test is conservative. Mechanically, Wald tests are
asymptotic:
when few patients contribute cells the asymptotics have not arrived, the
variance is biased upward relative to the true sampling variance, and $|t|$
comes out systematically too small.

## What that costs

Because $|t|$ is scaled by $\sqrt{\lambda}$, a deflated test needs a
proportionally larger effect to clear the same bar. Take a real effect that
deserves $p = 0.001$ and ask what a fit at each $\lambda$ actually reports:

```{r cost-table}
t_true <- stats::qnorm(1 - 0.001 / 2)          # |t| = 3.29
lam <- c(1.00, 0.90, 0.80, 0.70, 0.60, 0.50)
cost <- data.frame(
  lambda      = lam,
  t_observed  = round(t_true * sqrt(lam), 2),
  p_reported  = signif(2 * stats::pnorm(-abs(t_true * sqrt(lam))), 2)
)
cost$times_worse <- round(cost$p_reported / 0.001, 1)
cost
```

A cell type at $\lambda = 0.5$ reports a genuine $p = 0.001$ effect as roughly
$p = 0.02$. Multiply that penalty by a multiple-testing correction over millions
of tests and such a compartment becomes effectively unable to produce a
discovery, however real its biology.

# What drives lambda in a spiDE fit

The `condition` contrast is **between patients**, and the random intercept
already absorbs patient-level shifts. What buys precision is therefore the
number of *patients* contributing usable cells of that type -- not the raw cell
count. The two are correlated but they come apart, and when they do it is the
patient count that tracks calibration.

$\lambda$ can depart from 1 in **either** direction, and which way tells you
which failure mode dominates:

- **Many cells spread over few patients** -- cell-level pseudo-replication. The
  reduced-design Wald standard errors treat cells as independent replicates, so
  statistics come out too *large* and $\lambda > 1$. This is what the
  mixed-effects correction exists to fix (see `vignette("spiDE-model")`).
- **Few cells and few patients** -- small-sample Wald conservatism. The
  asymptotics have not arrived, the variance is biased upward, statistics come
  out too *small* and $\lambda < 1$.

Both resolve towards 1 as information accumulates, from opposite sides.

The relationship is best seen across the cell types *within one real fit*, where
compartment sizes span orders of magnitude: abundant, widely distributed
compartments sit at $\lambda \approx 1$ while sparse ones fall well below it. A
small simulated design cannot show this -- its variance-component estimate is
itself unstable at small $n$, so repeated fits of the same configuration scatter
$\lambda$ more than changing the configuration does. A compartment with a healthy total cell count but
most of its cells concentrated in a few patients behaves like a much rarer
population -- the cell count flatters it, the patient count does not.

# Diagnosing your own fit

$\lambda$ should be read **per index cell type**, never as a single number for
the fit. A global value is an abundance-weighted mixture of well- and
poorly-calibrated compartments and hides exactly the heterogeneity you need.

```{r per-celltype, eval = FALSE}
lambda_by_celltype <- function(fit, type = "ResponseNiche") {
  cm <- as.data.frame(fit@coefmap)
  # `covtype` spans EVERY design column, including the random-effect block,
  # whereas `t_stat` holds only the tested ones. A logical index built from
  # covtype therefore overruns t_stat -- always match on column names.
  cols <- intersect(colnames(fit@t_stat), cm$covariate[cm$type == type])
  idx  <- cm$index[match(cols, cm$covariate)]
  out  <- lapply(split(cols, idx), function(cc) {
    data.frame(n_col = length(cc),
               lambda = lambda(fit@t_stat[, cc, drop = FALSE]))
  })
  cbind(index = names(out), do.call(rbind, out))
}

# ff <- fits(res)[[1]]
# lambda_by_celltype(ff)
```

Pair that with the number of patients contributing at least a handful of cells
of each type, which is the quantity that actually predicts it:

```{r patient-spread, eval = FALSE}
tab <- table(spe$cell_type, spe$sample_id)
data.frame(
  index       = rownames(tab),
  n_cells     = rowSums(tab),
  n_patients  = rowSums(tab > 0),
  n_pat_ge10  = rowSums(tab >= 10)
)
```

# What to do about it

## Do not rescale by lambda

Classical genomic control divides every statistic by $\lambda$. **That is the
wrong move here.** It assumes a single inflation factor applies to all tests,
whereas in a `spiDE` fit $\lambda$ varies substantially between cell types
*within the same fit*. A global correction deflates the well-calibrated
compartments and still leaves the sparse ones under-powered. The heterogeneity
is the finding, not noise to be divided out.

## Do not reach for a bigger model

A random slope, a different ridge penalty, or a refit will not move $\lambda$
when the cause is small-sample conservatism. Two checks are worth running before
concluding that the model is at fault:

- **Compare a fixed-effects fit against the mixed fit.** If the mixed model were
  over-shrinking, the fixed fit would be the better calibrated of the two. If it
  is instead *worse*, the random effect is helping and is not the culprit.
- **Look at the dispersion multiplier used at inference.** If its median is
  below 1 it is *shrinking* standard errors and improving $\lambda$, so it
  cannot be the source of deflation either.

Conservatism at small $n$ is expected behaviour for a Wald test on a sparse
negative-binomial fit. It is not a bug, and there is no model change that
removes it.

## Restrict what you test

The productive response is to stop testing coefficients that cannot support
inference, filtering on **patients contributing cells** rather than on total
cells. This helps twice: the anti-powered tail stops consuming the
multiple-testing budget, so the well-calibrated compartments gain power, and the
output stops presenting uninformative nulls as though they were evidence.

Whatever threshold you choose, report it, and treat a null result from an
excluded or poorly calibrated compartment as **weak evidence of absence** rather
than evidence of no effect.

# When lambda exceeds 1

$\lambda > 1$ is not automatically alarming. Because it is a median, it cannot
be pushed up by a handful of strong hits -- moving it requires a *large fraction
of all genes* to shift. Two situations produce that:

1. **Residual confounding or anti-conservative inference.** The classical
   concern, and the reason to check the design before interpreting.
2. **A genuine programme-level shift.** If a compartment really does change
   coordinately across a broad set of genes, the median moves with them and
   $\lambda$ exceeds 1 legitimately.

These are distinguishable. A confounding-driven inflation affects the null
genes indiscriminately; a real programme shift concentrates in one compartment
and one direction, and should reproduce in an independent contrast. Where a
single cell type shows $\lambda > 1$ while the rest of the fit sits at or below
1, a coordinated biological shift in that compartment is the more parsimonious
reading -- and it argues for interpreting the result at the level of the
programme rather than gene by gene.

# Session info

```{r session}
sessionInfo()
```
