---
title: "The spiDE model"
author: "Dharmesh D. Bhuva"
package: spiDE
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{The spiDE model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette explains what spiDE computes and why, one step at a time. Each
step starts in plain language and then gives the formula, so the text can be
read with or without the maths. The quick start (`vignette("spiDE")`) shows how
to run the functions; the calibration vignette (`vignette("spiDE-calibration")`)
shows how well the tests hold up on real data.

# The question

Take one cell type, the **index** type (say, tumour cells), and a second cell
type, the **niche** type (say, fibroblasts). spiDE asks whether a gene's
expression in the index cells changes with the number of niche cells around
them.

- The **pooled test** asks: across patients, is there on average such a
  change?
- The **condition-specific test** asks: does the change differ between two
  groups of patients (responders and non-responders, early and late stage)?

One gene, one index type and one niche type make a **triplet**. spiDE tests
every triplet.

Both questions are about patients, not cells. The condition belongs to a
patient, and so does the way a patient's tissue responds to its
neighbourhood. spiDE therefore treats the **patient as the unit of
replication**. A result is strong when patients agree, not when a patient
contributes many cells. Treating cells as replicates would make every result
look more certain the more cells were measured, even if patients disagreed.

# What kinds of effects can spiDE find?

A gene's expression in the index cells can be tied to a niche type in more
than one way, and a condition can change each of them. Each of spiDE's tests
is built for one kind of effect, and two common effects of a condition are
left to other tools. The first thing to separate is the scale of the
comparison:

- **Within a patient.** In the same tissue, do index cells with more niche
  cells around them express more (or less) of the gene than index cells with
  fewer? This compares cells that share everything that belongs to the
  patient: its genetics, its treatment, how its sample was handled, and how
  much of each cell type its tissue contains.
- **Between patients.** Do patients whose index cells sit, on average, among
  more of the niche type express more (or less) of the gene overall? This
  compares patients with one another, one value per patient.

Each of the two can be the same in both groups of patients or differ between
them. The figure below draws every case on mock data, together with two
effects of a condition that spiDE does not test.

```{r effect-types, echo = FALSE, fig.width = 7, fig.height = 8.6, out.width = "100%", fig.cap = "Mock expression trends of one gene in one index type, against the density of one niche type, for ten patients in two groups (early and late stage). Each faint dot is an index cell, each thin line a patient's own fitted slope and each open circle a patient's mean; a dashed line follows the patients' means. Each panel names the test that detects its effect. A and B: slopes within patients, tested by `testSpiDE()`. C and D: an association between patients, tested by `testNicheAbundance()`. E and F: effects of the condition that no spiDE test looks at."}
library(spiDE)
library(ggplot2)
set.seed(7)
groups <- c("early", "late")
panels <- c(
  A = "A  Within patients, in both groups\npooled test: testSpiDE()",
  B = "B  Within patients, in one group only\ncondition-specific test: testSpiDE(condition =)",
  C = "C  Between patients, in both groups\ntestNicheAbundance(), term \"niche\"",
  D = "D  Between patients, in one group only\ntestNicheAbundance(), term \"condition:niche\"",
  E = "E  Only the level differs between groups\nnot a spiDE test: pseudobulk DE",
  F = "F  Only the amount of niche differs\nnot a spiDE test: composition analysis")
n_pat <- 10
n_cell <- 40
late <- rep(c(FALSE, TRUE), each = n_pat / 2)
around <- function(lo, hi) runif(n_pat, lo, hi)
# both groups spread evenly over the same range of patient mean densities
spread <- c(seq(0.5, 2.7, length.out = n_pat / 2), seq(0.7, 2.9, length.out = n_pat / 2))
# per panel and patient: the mean niche density m, the level a at that
# density, and the within-patient slope b
truth <- list(
  A = list(m = around(1.2, 2.2), a = rnorm(n_pat, 1.6, 0.35), b = rnorm(n_pat, 0.65, 0.05)),
  B = list(m = around(1.2, 2.2), a = rnorm(n_pat, 1.6, 0.35),
           b = ifelse(late, 0.75, 0) + rnorm(n_pat, 0, 0.05)),
  C = list(m = spread, a = 0.7 + 0.6 * spread + rnorm(n_pat, 0, 0.08), b = rnorm(n_pat, 0, 0.05)),
  D = list(m = spread, a = ifelse(late, 0.3 + 0.9 * spread, 1.6) + rnorm(n_pat, 0, 0.08),
           b = rnorm(n_pat, 0, 0.05)),
  E = list(m = around(1.2, 2.2), a = ifelse(late, 2.4, 1.1) + rnorm(n_pat, 0, 0.12),
           b = rnorm(n_pat, 0, 0.05)),
  F = list(m = ifelse(late, around(2.0, 2.8), around(0.6, 1.4)), a = rnorm(n_pat, 1.6, 0.12),
           b = rnorm(n_pat, 0, 0.05)))
cells <- do.call(rbind, lapply(names(truth), function(p) {
  tr <- truth[[p]]
  do.call(rbind, lapply(seq_len(n_pat), function(s) {
    l <- pmax(0, tr$m[s] + rnorm(n_cell, 0, 0.4))
    data.frame(panel = p, patient = s, condition = groups[late[s] + 1L], density = l,
               expr = tr$a[s] + tr$b[s] * (l - tr$m[s]) + rnorm(n_cell, 0, 0.3))
  }))
}))
# each patient's own least-squares line over the central 90% of its
# densities, and the patient's mean
fits <- do.call(rbind, lapply(split(cells, list(cells$panel, cells$patient)), function(d) {
  cf <- coef(lm(expr ~ density, d))
  x <- quantile(d$density, c(0.05, 0.95), names = FALSE)
  data.frame(panel = d$panel[1], condition = d$condition[1], x = x[1], xend = x[2],
             y = cf[1] + cf[2] * x[1], yend = cf[1] + cf[2] * x[2],
             mean_density = mean(d$density), mean_expr = mean(d$expr))
}))
cells$panel <- factor(panels[cells$panel], panels)
fits$panel <- factor(panels[fits$panel], panels)
cells$condition <- factor(cells$condition, groups)
fits$condition <- factor(fits$condition, groups)
trend_both <- fits[fits$panel == panels[["C"]], ]
trend_by_group <- fits[fits$panel == panels[["D"]], ]
ggplot(cells, aes(density, expr, colour = condition)) +
  geom_point(size = 0.6, alpha = 0.3, shape = 16) +
  geom_segment(data = fits, aes(x = x, xend = xend, y = y, yend = yend), linewidth = 0.55) +
  geom_smooth(data = trend_both, aes(mean_density, mean_expr), inherit.aes = FALSE,
              method = "lm", formula = y ~ x, se = FALSE, colour = "black",
              linetype = "22", linewidth = 0.6) +
  geom_smooth(data = trend_by_group, aes(mean_density, mean_expr), method = "lm",
              formula = y ~ x, se = FALSE, linetype = "22", linewidth = 0.6) +
  geom_point(data = fits, aes(mean_density, mean_expr), shape = 21, fill = "white",
             size = 1.7, stroke = 0.6) +
  facet_wrap(~ panel, ncol = 2) +
  scale_colour_manual(values = setNames(spiDEColours("condition"), groups),
                      name = "patient group") +
  labs(x = "niche density around the index cell (log1p)",
       y = "expression in the index cell (log scale)") +
  theme_spiDE(rl = 0.9) +
  theme(strip.text = element_text(hjust = 0, lineheight = 1.05), legend.position = "bottom",
        axis.text = element_blank(), axis.ticks = element_blank()) +
  guides(colour = guide_legend(override.aes = list(alpha = 1, size = 2, linewidth = 0.8)))
```

## Within patients: what `testSpiDE()` tests

- **A. The same slope in every patient** (independent of the condition).
  Within each patient, the gene rises (or falls) with the density of the
  niche type, and the patients agree. Patients may differ in their overall
  level of the gene, as the lines in panel A sit at different heights; that
  does not matter, because only the slopes are compared. The **pooled test**
  detects it (`testSpiDE()`, then `results(res, test = "pooled")`). A call
  means: in the same tissue, index cells with more of this niche type around
  them express more (or less) of the gene, consistently across patients.
- **B. A slope that differs between the groups** (condition-specific). The
  slope is present in one group only (panel B), is stronger in one group, or
  runs in opposite directions. The **condition-specific test** detects it
  (`testSpiDE(condition = )`, then `results(res, test = "condition")`), as the
  difference in average slope, second level minus first. A call means: the
  index cells' response to each extra unit of niche density differs between
  the groups.

By default (`procedure = "filtered"`) the condition-specific test is run only
among triplets whose pooled test passes. A slope present in one group only
still has a non-zero average over all patients and can pass that filter. A
slope that reverses between the groups averages out close to zero and is
unlikely to pass; use `procedure = "all"` when such effects matter.

A within-patient slope is not always a response of the index cells.
Transcripts of neighbouring niche cells can be assigned to index cells when
cells are segmented, and they too rise with the niche's density.
`plotSpillover()` shows whether the calls are the niche type's own markers.

## Between patients: what `testNicheAbundance()` tests

spiDE's model gives each patient its own baseline level of every gene (the
patient intercepts of Step 2). Everything that differs between patients as a
whole is set aside before the slopes are estimated, including a real
association: patients whose index cells sit among more of the niche type may
express the gene differently overall. `testNicheAbundance()` asks that
question separately, on one pseudobulk profile per patient and index type,
against the patient's mean niche density.

- **C. Patients with more of the niche differ** (independent of the
  condition). Within each patient the gene is flat, but the patients' means
  rise with their mean niche density (panel C). The term `"niche"` of
  `testNicheAbundance()` detects it: across all patients without a condition,
  averaged over the two groups with one. A call means:
  patients whose index cells sit among more of this niche type have more (or
  less) of the gene in those cells.
- **D. That association differs between the groups** (condition-specific).
  The patients' means rise with their niche density in one group only
  (panel D). The term `"condition:niche"` of
  `testNicheAbundance(condition = )` detects it, as the difference between
  the groups, second level minus first.

These are statements about patients, not about cells responding to their
neighbours. Patients with more of a niche type differ in many other ways too:
in their overall composition, in sample quality, or in signal spilling
between cells across the whole section. A between-patient association cannot
be pinned on the niche, which is why spiDE's own tests work within patients.
Report each result as what it is, and never a result of one test as a result
of the other.

## What spiDE does not test

Two effects of a condition are absorbed by the patients' own levels, and no
spiDE test looks at them:

- **E. The gene's level differs between the groups**, whatever the
  neighbourhood (panel E). This is ordinary differential expression between
  conditions. Test it on pseudobulk profiles with a differential expression
  method such as `limma` or `edgeR`.
- **F. The amount of the niche differs between the groups** (panel F): a
  change in tissue composition, here more of the niche type around the index
  cells of late-stage patients, while expression does not change with it.
  Test it with a composition method, such as `sccomp` or `speckle` for
  cell-type proportions.

This matters most for a condition such as disease progression, which changes
tissue in many ways at once. Most of those changes are differences between
patients: late-stage tissue has more of some cell types and less of others
(F), and its cells express some genes more or less (E), often both together
and often in cell types that interact. Neither is a change in how cells
respond to their neighbours, and the condition-specific test sees neither. It
sees only a changed **response per unit of neighbour density** (B): within
late-stage tissue, index cells with more niche cells around them differ from
index cells with fewer by more, by less or in the opposite direction than
within early-stage tissue. A changed interaction between two cell types
reaches this test only when it changes that response. "Late-stage tumours
have more fibroblasts, and their tumour cells express more of a gene" is F
and E, not B, even when both happen in the same patients.

## The effects in the model

Reduced to one niche type, with each patient's niche densities centred at
their mean, the model of Step 2 reads, for gene $g$ in index cell $i$ of
patient $s$:

$$
\log \mathrm{E}[y_{gi}] \;=\; a_{gs} \;+\; \beta_{gs}\,(\ell_i - \bar\ell_s),
$$

where $\ell_i$ is the cell's niche density on the log1p scale, $\bar\ell_s$
the mean of $\ell_i$ over the patient's index cells, $\beta_{gs}$ the
patient's within-patient slope and $a_{gs}$ its level at its own mean
density. With $c_s$ the patient's group, each effect is a statement about
how these quantities vary across patients:

| Effect | In the model | Detected by | A call means |
|---|---|---|---|
| A. slope in every patient | the average $\beta_{gs}$ is not 0 | `testSpiDE()`, pooled test | within a patient, index cells with more of the niche around them express more (or less) of the gene |
| B. slope differs between groups | the average $\beta_{gs}$ differs with $c_s$ | `testSpiDE(condition = )`, condition-specific test | that within-patient response differs between the groups |
| C. patients with more niche differ | $a_{gs}$ changes with $\bar\ell_s$ | `testNicheAbundance()`, term `"niche"` | patients whose index cells sit among more of the niche differ in the gene's level |
| D. that association differs between groups | the change of $a_{gs}$ with $\bar\ell_s$ differs with $c_s$ | `testNicheAbundance(condition = )`, term `"condition:niche"` | the between-patient association differs between the groups |
| E. level differs between groups | $a_{gs}$ differs with $c_s$ | not spiDE: pseudobulk differential expression | the gene's overall level differs between the groups |
| F. amount of niche differs between groups | $\bar\ell_s$ differs with $c_s$ | not spiDE: composition analysis | the groups differ in how much of the niche surrounds their index cells |

spiDE's tests read only the $\beta_{gs}$. The intercepts $a_{gs}$ absorb C to
E, and F changes only where along the density axis each patient's slope is
measured.

# Step 1: the niche

For every cell, `buildNiches()` measures how many cells of each type surround
it. It places a Gaussian bump of width $\sigma$ (the *bandwidth*, in the units
of the spatial coordinates) on every cell of a type and adds them up, which
gives a smooth *density* of that type at every point of the tissue. The
density of type $n$ around cell $i$ is $d_{in}$.

spiDE uses $\ell_{in} = \log(1 + d_{in})$, which compresses very dense
neighbourhoods so that a few crowded areas do not dominate.

The bandwidth sets what "nearby" means: a small $\sigma$ looks at immediate
neighbours, a larger one at the surrounding tissue. Fit one bandwidth at a
time.

# Step 2: a regression within each index type

Each index type is analysed separately. Within it, each gene's counts are
modelled by a **negative binomial regression**, the standard model for
sequencing counts. Its mean can depend on covariates and its variance allows
for more spread than pure sampling noise. For gene $g$ in cell $i$ of
patient $s(i)$:

$$
\log \mathrm{E}[y_{gi}] \;=\;
\underbrace{a_{g,s(i)}}_{\text{patient's own level}} \;+\;
\underbrace{\gamma_g^\top z_i}_{\text{cell covariates}} \;+\;
\underbrace{\textstyle\sum_{n} \beta_{gn}\, \ell_{in}}_{\text{niche slopes}}
$$

- **$a_{gs}$, one intercept per patient.** Each patient gets its own baseline
  level of the gene. This absorbs everything that is constant within a
  patient: its average expression, its sample handling, its condition, and
  its average niche. What is left to explain the niche slopes is variation
  *within* each patient: cells of the same patient in denser or sparser
  neighbourhoods. So $\beta_{gn}$ is a **within-patient slope**. It answers "in
  the same tissue, do cells with more neighbours of type $n$ express more of
  the gene?", not "do patients with more type-$n$ cells express more of it?".
- **$z_i$, cell covariates.** By default the cell's log library size (its
  total count, centred), with a slope per gene. This adjusts for sequencing
  depth (`depth = "loglib"`; see the options below). Other cell-level
  measurements, such as cell area, are added with `covariates`.
- **$\ell_{in}$, the niche columns.** Every cell type other than the index
  type enters, so each slope is adjusted for the other niches. Only the types
  named in `niche` are tested. An index type is never tested against its own
  niche, including any `mergeNiches()` group it belongs to.
- **The dispersion $\psi_g$** measures how much more variable the gene is than
  pure counting noise. It is estimated per gene.

The fit uses `SpaNorm::polishNB()`, which fits each gene on its own. The patient
intercepts are handled algebraically, so a fit with many patients costs about
the same as one with few. Genes are independent, so `BPPARAM` fits them in
parallel.

**Depth options.** `depth = "loglib"` (the default) uses one slope on the log
library size. `"nonlinear"` lets the effect of depth bend (a spline with three
degrees of freedom). `"spatial_spline"` lets it vary smoothly across each
section. `"none"` leaves depth out, for use with `offset`, a precomputed
log-scale offset per gene and cell. Library size mixes technical capture with
biological RNA content, so adjusting for it also removes any part of the niche
effect that works through a cell's total RNA. A slope is then expression
*relative to the cell's total*.

# Step 3: each patient's own slope

The regression above gives one pooled slope per gene and niche. The **slopes
engine**, the default, then asks what each patient's own slope is.

Refitting the model for every patient would be slow and unstable for patients
with few cells. Instead, spiDE starts from the pooled fit and lets each
patient's own cells move the slope once toward what fits that patient best.
This is a single step of the Newton-type algorithm (Fisher scoring) used to fit
such models. With working weights $w_i = \mu_i/(1 + \psi\mu_i)$ and residuals
$r_i = (y_i - \mu_i)/(1 + \psi\mu_i)$ at the pooled fit, and $\tilde\ell_i$ the
niche columns centred within the patient:

$$
\hat\beta_{gs} = \hat\beta_g + I_s^{-1} u_s, \qquad
I_s = \sum_{i \in s} w_i \tilde\ell_i \tilde\ell_i^\top, \qquad
u_s = \sum_{i \in s} \tilde\ell_i\, r_i .
$$

$u_s$ is how strongly the patient's cells pull away from the pooled slope, and
$I_s$ is how much information they carry. A patient with many cells and a wide
range of niche densities has a large $I_s$ and a precise slope. A patient
whose niche columns barely vary, or vary together, gives no usable slope for
that gene and is left out of its test.

**How precise is a patient's slope?** Two answers are computed.

- The **model variance**, $v^{\text{model}}_{gs} = \mathrm{diag}(I_s^{-1})$,
  is what the regression would give if every cell were independent.
- Neighbouring cells are not independent: they share local tissue conditions,
  so their residual expression is similar. A **spatial variance** $v^{\text{tile}}$
  corrects for this. Each section is cut into square tiles (side `tile`,
  default $3\sigma$), the patient's score is summed per tile, and the slope's
  variance is computed from how much the tile sums vary:
  $v^{\text{tile}} = \mathrm{diag}\big(I_s^{-1} (\sum_t U_t U_t^\top) I_s^{-1}\big)\, T/(T-1)$
  over the $T$ tiles, with $U_t$ the score of tile $t$. A patient with fewer
  than `min.tiles` tiles takes the model variance scaled by the typical ratio.

The tile variance of a single gene is noisy, so spiDE summarises it as **one
inflation factor per patient and niche**, shared by all genes:
$f_{sn} = \max\{1, \mathrm{median}_g(v^{\text{tile}}_{gsn}/v^{\text{model}}_{gsn})\}$.
A patient's slope variance is then $v_{gs} = v^{\text{model}}_{gs} f_{sn}$.

Because the fit never sees the condition, any patient-level condition can be
tested afterwards (`testSpiDE(fit, condition = ...)`). Retesting under a
reshuffled condition reruns only the test, which makes permutation checks
cheap.

# Step 4: combining patients

Each gene now has one slope per patient, each with a variance. The tests
combine them across patients.

**How much patients disagree.** Patients' slopes differ for two reasons:
each is estimated with noise ($v_{gs}$), and patients genuinely differ. The
genuine spread is the **between-patient variance** $\tau^2_g$, estimated from
how much more the slopes scatter than their noise explains (the
DerSimonian-Laird estimator of meta-analysis, computed without the
condition).

**Weights.** Patient $s$ is weighted by

$$
w_{gs} = \frac{1}{v_{gs} + \tau^2_g}.
$$

A precise patient counts more, but never unboundedly more. When patients
genuinely disagree ($\tau^2_g$ large), every patient's weight approaches the
same value, because no single patient can speak for the rest.

## The pooled test

**Is the average slope zero?** The weighted average of the patients' slopes is
tested with `limma`'s moderated $t$-test. The moderated test borrows
information across genes to stabilise each gene's variance, which helps when
patients are few. Two refinements keep it honest:

- **Borrowing is matched by expression level.** A bright gene's variability is
  compared with other bright genes, not with the typical gene
  (`eBayes(trend = TRUE, robust = TRUE)` on the gene's log mean expression).
- **The degrees of freedom count effective patients.** When a few cell-rich
  patients carry most of the weight, the test should not behave as if every
  patient contributed equally. The Kish effective number of patients,
  $n_{\text{eff}} = (\sum_s w_s)^2 / \sum_s w_s^2$, equals the number of
  patients when the weights are equal and falls towards one as they
  concentrate. The $t$ statistic is referred to $\min(\nu_{\text{limma}},
  n_{\text{eff}} - 1)$ degrees of freedom.

## The condition-specific test

**Does the average slope differ between the two groups?** The patients' slopes
are regressed on the condition, with the same weights $w_{gs}$:

$$
\hat\beta_{gs} = b_0 + \Delta_g\, c_s + [\text{strata}] + \text{error},
$$

where $c_s$ is 1 for the second condition and 0 for the first. The estimate
$\hat\Delta_g$ is the difference in average slope, second level minus first.

Its standard error does not assume the weights are exactly right. It is a
**robust (HC2 sandwich) standard error**, computed from how far each patient's
slope actually sits from its group's fit, with small-sample **Bell-McCaffrey
degrees of freedom**. This matters when some patients' slopes are noisier than
their weights suggest. A marker gene that is patchy within the index type is
an example: its slope can swing with where the patches fall. Such a patient
widens the interval rather than producing a call.

The weights come from the gene-shared variances $v_{gs}$ and never from a
gene's own tile variance. A gene's own tile variance, computed at the pooled
fit, grows when a patient's slope sits far from the pooled slope. Weighting by
it would down-weight exactly the patients that differ most and shrink the
estimated difference toward zero.

`strata` (a patient-level slide or batch column) adds one term per stratum to
the condition model. Use it when the condition is partly confounded with
slide, so that slide differences in slope are not read as condition
differences. A condition entirely confounded with the strata cannot be tested
and is refused.

# Multiple testing

spiDE tests many triplets, so it controls the **false discovery rate** (the
expected share of false calls among the calls) with the Benjamini-Hochberg
procedure. `testSpiDE()` reports a $q$-value per row:

- **pooled**: over every tested triplet;
- **condition-specific**: by default (`procedure = "filtered"`) over the
  triplets whose pooled test passes at `fdr`. Rows outside that family have
  `q = NA`. The pooled test ignores the condition, so this family is the same
  under any relabelling of patients and the false discovery rate is preserved.
  The correction is then spent where there is a niche effect to modify.
  `procedure = "all"` uses every triplet instead.

The estimate is on the natural-log scale per unit $\log(1 + d)$. A pooled slope
of 0.2 means expression rises by about 22% ($e^{0.2}$) for each unit increase
in $\log(1 + \text{density})$, in the same patient.

# The sandwich engine

`engine = "sandwich"` takes a different route to the same question. It fits
the regression of Step 2 once, with the condition's interaction terms
$c_s \ell_{in}$ when given a condition, and tests the coefficients directly.
Its standard error is a **patient-clustered CR2 sandwich** (Bell and McCaffrey
2002; Pustejovsky and Tipton 2018) with Bell-McCaffrey degrees of freedom: it
lets patients differ without modelling how. It needs the condition at fit
time. It is simpler but less powerful when effects vary between patients, and
`spiGSEA()` needs the slopes engine.

# Gene sets: `spiGSEA()`

A gene set is tested exactly like a gene. In each patient, the set's slope is
the average of its genes' slopes, each gene first divided by its typical
spread between patients, $\sqrt{\tau^2_g + \mathrm{median}_s v_{gs}}$, so
that a highly expressed gene does not dominate.

- **Competitive** (the default): the same average over the other tested genes
  is subtracted. The test then asks whether the set moves *more than the
  typical gene*. This guards against a shift that moves every gene at once,
  such as signal spilling over from neighbouring cells.
- **Self-contained**: nothing is subtracted, and the test asks whether the set
  moves at all.

The patients' set slopes go through the same pooled and condition-specific
tests as genes' slopes. Because patients are the replicates, the correlation
between genes of a set is already reflected in how much the set's slope varies
from patient to patient, and needs no separate estimate. $q$-values are given
within each (index, niche, test) column and over every set tested
(`q.global`).

# A different question: niche abundance

The patient intercepts of Step 2 absorb one association on purpose. Patients
whose index cells sit, on average, among more niche cells may express a gene
differently *as a whole*. That is a comparison between patients, with as many
data points as patients. It can reflect biology, differences in tissue
composition, or signal spilling between neighbouring cells.
`testNicheAbundance()` asks it directly:

- per patient and index type, it averages expression into a pseudobulk
  profile (log counts per million) and computes the patients' mean niche
  density;
- it regresses one on the other across patients with `limma`, adding the
  difference between conditions when a condition is given.

Without a condition, the term `"niche"` is the association across all patients
(effect C of *What kinds of effects can spiDE find?*). With a condition,
`"niche"` is the association averaged over the two conditions, and the term
`"condition:niche"` is its difference between them, second level minus first
(effect D). The two terms are separate multiple-testing families.

With a few dozen patients, one patient at the edge of a niche type's
abundance can carry a whole between-patient slope; if a gene is also unusual
in that patient, the slope looks far more certain than it is. So patients
whose leverage in the regression exceeds `max.leverage` times the average are
down-weighted until none does. The weights come from the patients' niche
densities and conditions only, never from expression, so the tests stay
exact. Each row reports the largest leverage before this cap and how many
patients were down-weighted.

A result from this test says "patients with more of this niche differ". A
result from `testSpiDE()` says "within a patient, cells with more of this niche
differ". Report each as what it is.

`plotNicheResponse(adjust = FALSE)` draws both on the example data. Each
segment is one patient's slope, the within-patient association spiDE tests; the
dashed line through the patients' means (open circles) is the between-patient
association the intercepts absorb:

```{r abundance-figure, fig.width = 6, fig.height = 3.5, message = FALSE}
library(spiDE)
data(toySpiDE)
spe <- buildNiches(toySpiDE, sigma = 30)
res <- spiDE(spe, condition = "condition", sigma = 30, index = "A", verbose = FALSE)
plotNicheResponse(spe, res, gene = "G1", index = "A", niche = "B", adjust = FALSE)
```

# What the model does not address

- **Smoothness of the niche itself.** A niche density is smooth over space.
  The spatial variance accounts for residual expression that is similar in
  neighbouring cells, and the calibration checks use nulls that keep the
  niche's smoothness, but the model does not describe it.
- **Straight-line effects.** Each slope is linear in $\log(1 + d)$. A gene that
  responds only above a threshold density is captured only partly.
- **Which average.** The pooled slope is a weighted average over patients,
  with weights that favour precise patients. When cell-rich patients have
  systematically different slopes from sparse ones, this average differs from
  a simple average over patients.
