---
title: "Calibration: reading lambda, and what the benchmarks measured"
author: "Dharmesh D. Bhuva"
package: spiDE
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Calibration: reading lambda, and what the benchmarks measured}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, collapse = TRUE, comment = "#>", message = FALSE,
  warning = FALSE, fig.width = 7, fig.height = 4.2, dpi = 150, fig.retina = 1
)
set.seed(1)
```

```{r libs}
library(spiDE)
library(ggplot2)
source(system.file("spide_palette.R", package = "spiDE"))
```

# Overview

This vignette has two halves. The first is a diagnostic: how to read the
genomic inflation factor $\lambda$ from your own fit, per cell type, and what
to do about it. The second is the evidence: every number this package quotes
about its defaults — null calibration, power, and what was measured on a real
cohort — collected in one place with pointers to the reports that carry the
full tables (https://bhuvalab.github.io/spiDE-research/). The model those
numbers are about is in `vignette("spiDE-model")`.

# Reading lambda

A `spiDE` fit tests thousands of genes against hundreds of
`celltype:condition:niche` coefficients, and those coefficients are **not
equally well estimated**. A tumour compartment with tens of thousands of cells
spread over every patient supports a very different quality of inference from
a dendritic-cell compartment with a few hundred cells concentrated in a
handful of patients — yet both are reported in the same table, against the
same FDR threshold. $\lambda$ detects that.

- $\lambda$ measures whether your test statistics follow the null
  distribution you assumed. It is a property of the **test**, not of the
  biology.
- In `spiDE` it varies *within a single fit*, across cell types, because cell
  types differ in how many patients contribute cells.
- $\lambda < 1$ costs **power, not validity**. A null result for a poorly
  calibrated cell type is weak evidence of absence.
- The right response is to restrict which coefficients you test, **not** to
  rescale the statistics and **not** to change the model.

## What lambda is

For a one-degree-of-freedom test, the squared statistic $t^2$ is a $\chi^2_1$
variate under the null. The genomic inflation factor compares the observed
median with the null's:

$$\lambda = \frac{\mathrm{median}(t^2)}{\mathrm{median}(\chi^2_1)}
          = \frac{\mathrm{median}(t^2)}{0.4549}$$

The median is the important choice: a minority of true positives, however
large, barely moves it, so $\lambda$ reads the bulk of the distribution, which
is overwhelmingly nulls (Devlin and Roeder, 1999).

| $\lambda$ | reading |
|---|---|
| $= 1$ | the observed null matches the theoretical one; p-values mean what they say |
| $> 1$ | statistics too large: confounding or anti-conservatism, **or** genuinely widespread signal |
| $< 1$ | statistics too small: the test is conservative |

```{r lambda-fn}
lambda <- function(t) stats::median(t^2, na.rm = TRUE) / stats::qchisq(0.5, 1)
```

## Lambda below 1 does not mean "no signal"

Under a pure null with a calibrated test, $\lambda = 1$ exactly. Absence of
biology gives 1, not less than 1:

```{r pure-null}
z <- matrix(rnorm(200 * 50), 200, 50)
round(lambda(z), 3)
```

So $\lambda = 0.68$ is never a statement that a cell type is quiescent. It
says the reference distribution is wrong for the statistic's sampling
behaviour — here, that the test is conservative. Wald tests are asymptotic:
when few patients contribute cells, the variance is biased upward and $|t|$
comes out systematically too small. Because $|t|$ scales with
$\sqrt\lambda$, a deflated test needs a proportionally larger effect to clear
the same bar:

```{r cost-table}
t_true <- stats::qnorm(1 - 0.001 / 2)          # |t| = 3.29 for p = 0.001
lam <- c(1.00, 0.90, 0.80, 0.70, 0.60, 0.50)
cost <- data.frame(lambda = lam,
                   t_observed = round(t_true * sqrt(lam), 2),
                   p_reported = signif(2 * stats::pnorm(-abs(t_true * sqrt(lam))), 2))
cost$times_worse <- round(cost$p_reported / 0.001, 1)
cost
```

A cell type at $\lambda = 0.5$ reports a genuine $p = 0.001$ effect as about
$p = 0.02$; after a genome-wide correction such a compartment is effectively
unable to produce a discovery.

## What drives it

The `condition` contrast is between patients, and the random intercept
absorbs patient-level shifts, so what buys precision is the number of
*patients* contributing usable cells of that type, not the raw cell count.
$\lambda$ can depart from 1 in either direction, and the direction says which
failure dominates:

- **Many cells over few patients** — pseudo-replication; statistics too
  large, $\lambda > 1$. This is what the mixed-effects correction fixes.
- **Few cells and few patients** — small-sample Wald conservatism;
  statistics too small, $\lambda < 1$.

Both resolve toward 1 as information accumulates. A third driver is not a
variance at all: a between-patient composition association loading onto the
niche slopes when the design has no (sample × cell type) intercept. It
inflates $\lambda$ in exactly the pattern a real programme shift would —
concentrated in the compartments with the most cells, largest for the
brightest genes — and `re.celltype = TRUE` (the default) removes it; refit
before interpreting an inflated compartment from an older fit. The measured
form is under *The real cohort* below. A small measured cost of the defaults
on a clean null also means a $\lambda$ a few percent above 1 is expected from
the method and is not evidence of a data problem.

## Diagnosing your own fit

Read $\lambda$ **per index cell type**, never as one number for the fit — a
global value is an abundance-weighted mixture that hides the heterogeneity.

```{r per-celltype, eval = FALSE}
lambda_by_celltype <- function(fit, type = "ResponseNiche") {
  cm <- as.data.frame(fit@coefmap)
  # `covtype` spans every design column, `t_stat` only the tested ones:
  # always match on column names
  cols <- intersect(colnames(fit@t_stat), cm$covariate[cm$type == type])
  idx  <- cm$index[match(cols, cm$covariate)]
  out  <- lapply(split(cols, idx), function(cc) {
    data.frame(n_col = length(cc), lambda = lambda(fit@t_stat[, cc, drop = FALSE]))
  })
  cbind(index = names(out), do.call(rbind, out))
}
# lambda_by_celltype(fits(res)[[1]])
```

Pair it with the number of patients contributing cells of each type, the
quantity that predicts it:

```{r patient-spread, eval = FALSE}
tab <- table(spe$cell_type, spe$sample_id)
data.frame(index = rownames(tab), n_cells = rowSums(tab),
           n_patients = rowSums(tab > 0), n_pat_ge10 = rowSums(tab >= 10))
```

## What to do about it

**Do not rescale by lambda.** Classical genomic control divides every
statistic by $\lambda$; it assumes one inflation factor for all tests, whereas
here $\lambda$ varies between cell types within a fit. A global correction
deflates the calibrated compartments and leaves the sparse ones
under-powered. The heterogeneity is the finding.

**Do not reach for a bigger model.** A random slope, a different penalty or a
refit will not move $\lambda$ when the cause is small-sample conservatism.
Two checks before blaming the model: a fixed-effects fit should be *worse*
calibrated than the mixed fit, not better; and the dispersion multiplier at
inference should have a median below 1, i.e. be shrinking standard errors.
Conservatism at small $n$ is expected behaviour for a Wald test on a sparse
negative binomial fit.

**Restrict what you test.** Stop testing coefficients that cannot support
inference, filtering on patients contributing cells rather than on total
cells. The under-powered tail stops consuming the multiple-testing budget, and
the output stops presenting uninformative nulls as evidence. Report the
threshold, and treat a null from an excluded compartment as weak evidence of
absence.

**When lambda exceeds 1.** Because it is a median, a handful of strong hits
cannot move it; a large fraction of genes must shift. Residual confounding
affects null genes indiscriminately; a genuine programme-level shift
concentrates in one compartment and one direction and reproduces in an
independent contrast. Where one cell type shows $\lambda > 1$ and the rest of
the fit sits at or below 1, a coordinated shift in that compartment is the
more parsimonious reading — and it argues for interpreting at the level of
the programme rather than gene by gene.

# What the benchmarks measured

Each paragraph is one measured fact with the number that matters, in the
order the defaults are introduced in the model vignette. The simulation study
uses a realistic simulator (Gamma gene means, a mean–variance dispersion
trend, library sizes of about 1,500–2,000 counts per cell, configurable
spatial layouts), 40 replicates per design point, sample sizes $S$ from 4 to
30 and 300 cells per sample unless stated. Full tables: the *Simulation study*
report on the benchmark site. All figures are for `random = "intercept"` with
`df.method = "satterthwaite"` unless stated.

## The random intercept

On a null with per-sample intercepts and no response effect,
`random = "none"` rejects the response contrast at about **0.71** against a
nominal 0.05, worse as cell counts become imbalanced across samples; the
random intercept holds about **0.04**, matching a calibrated pseudobulk
reference. That is why `"none"` is not the default and should not be used for
inference.

## The reference degrees of freedom

`df.method = "between"` tests every coefficient against the scalar $S - 2$
and is severely over-conservative for the niche slopes when patients are few:
null type-I about **0.001** at $S = 4$ against 0.05, with near-zero power.
`"satterthwaite"` holds type-I in **0.042–0.065** over the whole sampled range
and gains about **0.10** in recall, with a mild liberal drift at larger $S$
(worst 0.065). A Kenward–Roger correction of the covariance is the indicated
refinement.

## The small-sample regime

`fitSpiDE(random = "intercept")` is a powerful estimator once samples are
plentiful — raw power **0.66** at $S = 30$ — but it fails at small $S$ in a way
the null table does not show: realised FDP **0.35** at $S = 4$ against a nominal 0.05, recovering
to 0.04 by $S = 16$. Any sub-analysis that thins the patient count re-enters
this regime.

## The 0.99.17 defaults: nested intercept and per-gene convergence

The two defaults were adopted on real-cohort evidence (below). Measured as
switch arms on the simulation study, which places niche cells by the same
potential in every sample and therefore plants **no** composition effect, they
can only cost there, and they do:

| null type-I, mean over $S \ge 10$ | at $S = 4$ |
|---|---|
| design term alone: 0.058 | 0.036–0.049 |
| convergence only: 0.066 | 0.086–0.095 |
| nested block only: 0.068 | 0.037–0.043 |
| both (the defaults): 0.071 | 0.085–0.096 |

(Per-cell standard error about 0.0016.) The two switches act at different
sample sizes. Convergence's cost is a small-$S$ effect — the per-gene
dispersion gives up cross-gene moderation, which matters most when data are
few — and fades onto the design-term arm by $S = 24$. The nested block is
harmless at $S = 4$ and holds a constant $+0.010$ from $S = 16$, which points
at its reference df or the estimation of its variance component. The pair
gains recall (**0.402** against 0.361 at $S = 30$) with slightly better FDP
from $S = 10$ and a worse one at $S = 4$ (0.59 against 0.35). None of this is
a reason to revert: the defaults were chosen for a confound real cohorts
carry, because patients differ in composition. Of the two candidate
refinements, the dispersion at the converged mean is measured next; the
nested block's reference df is not yet.

## The dispersion rule and the standard-error scale (0.99.18)

Four arms on the same benchmark, all with the nested block and the
convergence step, cross the dispersion rule of the convergence stage
(profile-ML per gene, or `fitNB`'s cross-gene moderated value kept) with the
scale of the standard errors (the Pearson working dispersion, or the
quasi-likelihood dispersion). Null type-I at a nominal 0.05, mean of both
layouts, 40 replicates:

| $S$ | profile $\psi$ + Pearson (0.99.17) | moderated $\psi$ + Pearson | moderated $\psi$ + QL (0.99.18) | profile $\psi$ + QL |
|---|---|---|---|---|
| 4 | 0.091 | 0.091 | 0.050 | 0.055 |
| 10 | 0.076 | 0.074 | 0.055 | 0.055 |
| 16 | 0.071 | 0.071 | 0.055 | 0.054 |
| 30 | 0.068 | 0.068 | 0.055 | 0.053 |

The dispersion rule is irrelevant — the two Pearson arms agree to the third
decimal, and so do the two QL arms — so the small-$S$ convergence cost was
never the loss of moderation. The scale is the whole difference, and the QL
scale is the first configuration measured here that holds the nominal level
at every $S$, including $S = 4$. Its recall at FDR 0.05 is that of the design
with neither switch (**0.360** against 0.361 at $S = 30$; the Pearson scale's
0.402 is bought with its inflated null: raw power at 0.05 is 0.666, 0.660
and 0.694 respectively) and its realised FDP sits far below nominal (0.014
at $S = 30$, 0.000 at $S = 4$ against 0.59), which says the hierarchical
cascade is conservative on better-calibrated p-values. On the real cohort
the QL and Pearson variants share 110 of about 115 calls and both keep 81 to
83 of the earlier 84. The moderated dispersion runs the cohort's real grid
in two thirds of the wall time (178 against 266 minutes) and the QL pre-pass
adds nothing measurable at cohort scale (about 5% at $S \ge 16$ on the
benchmark, more only on tiny designs). Both are the defaults from 0.99.18.

## The polish's two dispersion rules, timed, and the FDR procedure (0.99.20)

The polish stage of 0.99.20 converges each gene once from `fitNB`'s output
and then re-polishes warm after every variance-component step, with the loop
Steffensen-accelerated to its fixed point rather than capped. Its dispersion
rule is `psi = "profile"` (re-estimated by profile likelihood at the
converged mean, the default) or `psi = "moderated"` (`fitNB`'s cross-gene
value kept, which skips the profile search and the two re-polishes it
triggers). The two were measured from **one fit per simulated dataset**,
polished both ways, so the rule is the only difference (arm `polish-rules`,
40 replicates per point, random intercept, Satterthwaite df):

| $S$ | null type-I, profile / moderated | TPR at FDR 0.05 | realised FDP | raw power |
|---|---|---|---|---|
| 4 | 0.055 / 0.052 | 0.001 / 0.001 | 0.033 / 0.042 | 0.247 / 0.228 |
| 10 | 0.059 / 0.060 | 0.069 / 0.062 | 0.014 / 0.012 | 0.437 / 0.428 |
| 16 | 0.058 / 0.060 | 0.200 / 0.193 | 0.012 / 0.015 | 0.548 / 0.538 |
| 30 | 0.058 / 0.060 | 0.379 / 0.366 | 0.014 / 0.019 | 0.681 / 0.675 |

They are equivalent within the study's resolution (standard errors of
0.002 to 0.003 on the null and about 0.01 on recall), and on the real
cohort's full transcriptome their per-gene calibrated statistics agree at
$r = 0.999$ over two million triplets with the same six exceedances at the
calibrated threshold. The difference is cost. The polish is the whole cost
of the pipeline, three to six times the fit at ten thousand genes on one
core, and the moderated rule saves 13 to 19% of it here (at 24,000 cells:
fit 17.7 min, polish 100.9 against 87.3, inference 4.2) and half of it on
the cohort's bright genes (651 against 332 minutes on 64 workers).

Every fit of that arm was also scored with classic Benjamini–Hochberg over
all triplet tests, from the same $t$-statistics and reference df, beside the
hierarchical cascade `testSpiDE()` uses. At $S = 30$ under the profile rule
the cascade gives TPR 0.379 and realised FDP 0.014 at nominal 0.05 against
0.371 and 0.011 for classic BH, and 0.476 / 0.101 against 0.461 / 0.062 at
nominal 0.20: the cascade is the slightly less conservative and slightly
more powerful of the two at every level. Both sit far below the nominal
line, so the conservatism seen since 0.99.18 is a property of the
QL-scaled p-values, not of the multiplicity step; a different correction
does not reclaim it.

## The `CellType:condition` term, and a design that looks more powerful

The design spiDE shipped before 0.99.7 had no `CellType:condition` column,
and on the simulation study it reports far more niche calls: recall **0.36**
against 0.17 at FDR 0.05 on the same dataset, concentrated in marker genes.
That is not a better test. One power dataset ($S = 24$, $\beta = 2$) was fit
under four package builds including one that differs from that design *only*
by the added column; the build makes no difference (slopes correlate at
1.000), the column is the whole gap. Without it the niche slope exceeds the
term-restored one by **0.25×** the flat-response coefficient, gene by gene
($r = 0.88$ on planted genes, 0.92 on null genes) — the mean of the planted
effect counted as niche dependence — which inflates the estimate by only
about 10%; the $t$ is 1.5–1.6× larger because the standard error is
**0.67–0.72×**, a regression through the origin. On flat-shift genes with no
niche effect the same channel produces fabricated niche calls; the recall
saturates at about 0.5 from $\beta = 2$; and on the real cohort the
term-restored design recovered **41% more** triplets. Two-sided Cauchy
combination gave more discoveries than one-sided Brown in every build, so
sidedness is not involved. The recall of a design is comparable across
designs only when they estimate the same quantity.

## The combiner

Cauchy (two-sided) matches or beats Brown's method on power while holding
type-I error under correlation without estimating the correlation matrix;
report: *Combining niche p-values*.

## The gene-set layer

`spiGSEA(test = "competitive")` is calibrated; the self-contained form is
not (it called 20.6 of 208 null sets per replicate, all false) and is kept
only to reproduce older scripts. Report: *Gene-set inference*.

## The real cohort

The sharpest test in this package's history, and the one to repeat before any
claim of niche-dependent DE: a **niche-shuffle null** on a 55-patient CosMx
cohort, permuting the rows of the niche matrix within (sample, cell type) so
that the true slope is zero by construction, in `free` (random rows) and
`block` (toroidal shift, preserving spatial smoothness) forms. Report: *The real
cohort*.

- **Under the design shipped before 0.99.17, the real data were
  indistinguishable from their null**: $\mathrm{sd}(t)$ 1.100 against a
  shuffle maximum of 1.101. The heavy tail was per-gene — standardising each
  gene by its own null spread took the $p < 10^{-6}$ excess from 59.9× to
  2.6× — and confined to the brightest genes in the most populous index
  types. Twelve variance-side explanations were refuted. The cause was the
  composition confound of the model vignette: with a gene's converged fit
  held fixed, the permutation $t$ had the expected spread but a per-column
  mean up to $\pm 3$ that equalled the real-data $t$ column by column
  ($r = 0.66$); the shuffle preserves each group's mean niche density and so
  preserves the confound, which is why real and null matched.
- **With the nested intercept and convergence** the `free` null is flat at
  **0.96–0.98** in every expression band with zero $|t| > 4.89$ exceedances
  of 101,508 where the old design produced 95; the `block` null keeps a
  0.99 → 1.21 gradient in the ~100 brightest genes, the spatially smooth
  component, so calibrate against `block`. The real cohort now exceeds its
  null — 97 exceedances against a replicated null range of 5–11, with 100
  random control genes inside the range — and gives about a hundred calls at
  a per-gene empirical FDP of 0.05.
- **The count is not a set of findings.** Splitting Plasma out of the B cell
  compartment keeps the count and loses every immunoglobulin call; five
  imaging covariates keep 84 of 104 and drop the keratin-in-immune calls;
  bandwidths 10, 50 and 70 share only four triplets. One triplet survives
  every perturbation. The patient-level composition test on the same cohort
  shows neighbour types' own markers rising in the index cells' pseudobulk
  with the neighbours' density, uniformly across conditions — which is what
  the old design's immunoglobulin-in-tumour calls were.
- **Raw counts with a global library-size slope** are the best-calibrated
  substrate measured: the expression gradient of the null went from 1.207 to
  1.149 and the median null $\mathrm{sd}(t)$ from 1.055 to 1.008. A
  back-transformed assay is not counts, and cell-type-specific size factors
  buy nothing.

# Session info

```{r session}
sessionInfo()
```
