---
title: "Simulation study: power and FDR control of the spiDE mixed-effects correction"
author: "Dharmesh D. Bhuva"
package: spiDE
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Simulation study: power and FDR control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4.5)
```

# Overview

This is a **research/benchmarking vignette**, not user-facing documentation of the package API. It answers a concrete question that comes up when exploring `spiDE`'s mixed-effects correction (`random = "intercept"` / `"slope"`, see the companion vignette *The spiDE model: neighbourhood-dependent DE and mixed-effects inference*): if I simulate more samples, or more cells per sample, why doesn't power obviously go up?

The short answer, demonstrated quantitatively below: statistical power under the mixed-effects correction is driven almost entirely by the number of **samples** (patients), not the number of **cells**. That is not a bug — it is the price of correcting cell-level pseudo-replication. Cells within a sample share the same response label, so they cannot contribute independent evidence about a between-sample contrast; only more samples can. The bundled toy dataset (`.toySPE()` / `data(toySpiDE)`) obscures this because (a) it reseeds the *entire* dataset when you change `n_samples`/`n_per`, so effect size is confounded with sample size, and (b) it only plants one true gene among \~20, so the gene-level FDR gate swallows the signal. Here we build a **fair** simulation harness that isolates the two questions: does the null stay calibrated, and does power scale correctly with each knob.

All code needed to reproduce this study is self-contained in this file (no internal `spiDE:::` helpers are used) so it can be lifted out and run independently of the package if desired.

```{r load, message = FALSE}
library(spiDE)
library(SpatialExperiment)
library(BiocParallel)
library(ggplot2)

# keep the default vignette build light; bump these for a real study (see the
# final section for suggested "full" settings and expected runtime)
RUN_FULL <- TRUE
B <- if (RUN_FULL) 40 else 6              # replicate seeds per design point
NPAR <- if (RUN_FULL) 8 else 1            # parallel workers
BPPARAM <- if (NPAR > 1) MulticoreParam(NPAR, RNGseed = 1) else SerialParam()
set.seed(1)  # fitSpiDE is stochastic (fitNB subsamples cells for dispersion)
```

`MulticoreParam` parallelises by forking. SpaNorm's GPU path (used by `fitSpiDE()` when `backend = "auto"`, the default) initialises `torch`'s Metal/MPS runtime, and that runtime is **not fork-safe** — a forked worker crashes with `+[MPSGraphObject initialize] ... fork() ... Crashing instead`. All fits below therefore pass `backend = "cpu"`, which keeps SpaNorm off the GPU so the forked workers are safe. (The models here are small; the CPU path is the right choice anyway. On a machine with a usable GPU, keep the GPU by parallelising with the non-forking `SnowParam(NPAR)` instead.)

# Data generators

Both generators plant a **patient-level random intercept** (`sd_patient`) on top of the cell-type baseline, so — unlike `.toySPE()` — every simulated dataset has genuine between-sample clustering, the actual source of pseudo-replication. `sim_null()` has no response effect at all (used for Type-I error / calibration); `sim_alt()` additionally plants a known, gene-indexed A-index / B-niche response effect in a specified fraction of genes (used for power and FDR).

```{r generators}
.simulate_spe <- function(S, n_per, n_genes, field, sd_patient, seed,
                          nonnull = integer(0), beta = 0, sd_slope = 0) {
  set.seed(seed)
  cts <- c("A", "B", "C")
  gn <- sprintf("G%d", seq_len(n_genes))
  sids <- sprintf("S%d", seq_len(S))
  cond <- rep(c("Responder", "Non-responder"), length.out = S)
  names(cond) <- sids

  cells <- lapply(sids, function(sid) {
    x <- runif(n_per, 0, field)
    y <- runif(n_per, 0, field)
    ct <- ifelse(x > 0.6 * field & runif(n_per) < 0.7, "B",
                sample(c("A", "C"), n_per, replace = TRUE))
    data.frame(sample_id = sid, condition = cond[[sid]],
              x = x, y = y, cell_type = ct, stringsAsFactors = FALSE)
  })
  cd <- do.call(rbind, cells)
  n <- nrow(cd)
  cd$cell_id <- sprintf("c%d", seq_len(n))

  # per-cell-type baseline log-mean
  base <- matrix(rnorm(n_genes * length(cts), 1.5, 0.6), n_genes,
                 dimnames = list(gn, cts))
  lmu <- base[, cd$cell_type, drop = FALSE]

  # per-(gene, sample) random intercept: real clustering, no response effect
  u <- matrix(rnorm(n_genes * S, 0, sd_patient), n_genes,
             dimnames = list(gn, sids))
  lmu <- lmu + u[, cd$sample_id, drop = FALSE]

  # planted A-index / B-niche response effect in the `nonnull` genes; x/field
  # is a proxy for local B-niche density (B cells sit at high x)
  if (length(nonnull) > 0 && beta != 0) {
    is_A <- cd$cell_type == "A"
    is_resp <- cd$condition == "Responder"
    slope_shift <- if (sd_slope > 0) {
      # per-sample random slope noise, shared across the nonnull genes'
      # per-sample deviation but independent across samples
      v <- setNames(rnorm(S, 0, sd_slope), sids)
      v[cd$sample_id]
    } else 0
    eff <- (beta + slope_shift) * is_A * is_resp * (cd$x / field)
    for (g in nonnull) lmu[g, ] <- lmu[g, ] + eff
  }

  mu <- exp(lmu)
  counts <- matrix(rnbinom(length(mu), mu = as.vector(mu), size = 5), n_genes,
                   dimnames = list(gn, cd$cell_id))
  cd$Area <- exp(rnorm(n, sd = 0.2))
  cd$nCount <- colSums(counts)

  SpatialExperiment(assays = list(counts = counts),
                    colData = S4Vectors::DataFrame(cd),
                    spatialCoords = as.matrix(cd[, c("x", "y")]))
}

sim_null <- function(S = 10, n_per = 90, n_genes = 30, field = 500,
                     sd_patient = 0.7, seed = 1) {
  .simulate_spe(S, n_per, n_genes, field, sd_patient, seed)
}

sim_alt <- function(S = 10, n_per = 90, n_genes = 150, field = 500,
                    prop_nonnull = 0.15, beta = 2, sd_patient = 0.7,
                    sd_slope = 0, seed = 1) {
  n_nonnull <- max(1, round(prop_nonnull * n_genes))
  nonnull <- seq_len(n_nonnull) # G1..Gk are the true positives, by construction
  spe <- .simulate_spe(S, n_per, n_genes, field, sd_patient, seed,
                       nonnull = nonnull, beta = beta, sd_slope = sd_slope)
  list(spe = spe, truth = seq_len(n_genes) %in% nonnull, nonnull_genes = nonnull)
}
```

# Shared driver

For a given simulated dataset we fit all three `random` modes at a single bandwidth and pull out two levels of evidence:

-   **raw Wald p-values** on the `ResponseNiche` (A-index x B-niche) coefficient, gate-free — the cleanest signal for calibration/power curves;
-   **end-to-end discoveries** from `results()`, thresholded on `fdr.niche` — what a user actually sees.

```{r driver}
sigma_fixed <- 50

.raw_AB_p <- function(fit1) {
  cn <- colnames(fit1@t_stat)
  col <- which(grepl("CellTypeA", cn) & grepl("B$", cn) & grepl("Responder", cn))
  stopifnot(length(col) == 1)
  df <- if (is.null(fit1@df)) Inf else fit1@df
  t <- fit1@t_stat[, col]
  se <- fit1@se[, col]
  list(t = t, p = 2 * stats::pt(-abs(t), df), coef = t * se)
}

run_one <- function(spe, random, fdr = 0.05) {
  spe <- buildNiches(spe, sigma = sigma_fixed)
  # backend = "cpu" is required for the parallel runs below: SpaNorm's GPU path
  # initialises torch's Metal/MPS runtime, which is not fork-safe, so the default
  # ("auto") crashes the forked MulticoreParam workers (see the setup chunk).
  fit <- fitSpiDE(spe, "condition", sigma = sigma_fixed, random = random,
                  backend = "cpu", verbose = FALSE, re.prop = 0.2)
  res <- testSpiDE(fit, spe = spe, fdr = fdr)
  f1 <- fits(res)[[1]]
  raw <- .raw_AB_p(f1)
  tab <- results(res)
  hit <- tab$gene[tab$ct_index == "A" & tab$ct_niche == "B" & tab$fdr.niche <= fdr]
  list(raw_p = raw$p, discoveries = unique(hit))
}

methods <- c(fixed = "none", intercept = "intercept", slope = "slope")
```

# Part A — calibration under the null

No response effect is planted (`sim_null()`); every response test should be null, so p-values should be $\mathrm{Unif}(0,1)$ and the rejection rate at $\alpha = 0.05$ should be $\approx 0.05$ for a correctly calibrated method.

```{r partA-run}
S_grid_null <- if (RUN_FULL) c(4, 6, 10, 16, 24) else c(6, 16)

partA <- bplapply(S_grid_null, function(S) {
  reps <- lapply(seq_len(B), function(b) {
    spe <- sim_null(S = S, seed = 1000 * S + b)
    p <- vapply(methods, function(m) mean(run_one(spe, m)$raw_p < 0.05), numeric(1))
    p
  })
  reps <- do.call(rbind, reps)
  data.frame(S = S, method = names(methods), reject_rate = colMeans(reps))
}, BPPARAM = BPPARAM)
partA <- do.call(rbind, partA)
partA
```

```{r partA-plot, fig.cap = "Type-I error (rejection rate at alpha=0.05) under the null, by number of samples S. Dashed line = nominal 0.05. The fixed-effects fit is inflated regardless of S; the mixed-effects fits are close to nominal and improve as S grows."}
ggplot(partA, aes(S, reject_rate, colour = method)) +
  geom_hline(yintercept = 0.05, linetype = "dashed", colour = "grey40") +
  geom_line() + geom_point(size = 2) +
  scale_y_log10() +
  labs(y = "P(raw p < 0.05 | null)", title = "Null calibration vs number of samples") +
  theme_bw()
```

A QQ-plot makes the same point at the level of the full p-value distribution
rather than a single threshold. Two subtleties matter for reading it correctly,
and a *single* simulated dataset gets both wrong:

* The variance components (`tau2`) are **shared across genes and estimated from
  only `S` samples**, so all `n_genes` response-niche p-values in one dataset
  ride on the *same* `tau2` estimate. One unlucky variance-component draw tilts
  the entire curve — the departure is coherent, not gene-by-gene noise. This is
  most pronounced for `random = "slope"`, whose extra per-sample slope variance
  is the least stable component. So we **pool over all `B` datasets**.
* `fitSpiDE()` is **not deterministic by default**: `fitNB` subsamples cells for
  the dispersion estimate using the RNG, so the same data + model can give
  visibly different p-values run to run. We therefore `set.seed()` before every
  fit (identically across the three methods, so they see the same draw).

```{r partA-qq, fig.cap = "QQ-plot of null response-niche p-values at S = 10, pooled over all B datasets with the fit RNG fixed. The fixed-effects fit is badly anti-conservative; the mixed-effects fits are far closer to the uniform (dashed) reference, though both remain mildly anti-conservative at this small S (the residual tail shrinks as S grows, cf. Part A)."}
p_qq <- bplapply(seq_len(B), function(b) {
  spe <- sim_null(S = 10, seed = 7000 + b)
  lapply(methods, function(m) { set.seed(b); run_one(spe, m)$raw_p })
}, BPPARAM = BPPARAM)
qq_df <- do.call(rbind, lapply(names(methods), function(m) {
  p <- sort(unlist(lapply(p_qq, `[[`, m), use.names = FALSE))
  data.frame(method = m, expected = -log10(ppoints(length(p))),
             observed = -log10(p))
}))
ggplot(qq_df, aes(expected, observed, colour = method)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(size = 1, alpha = 0.5) +
  labs(x = "expected -log10(p)", y = "observed -log10(p)") +
  theme_bw()
```

Two practical takeaways from this null study:

* **Reproducibility, not just averaging.** Because `fitNB` subsamples cells for
  the dispersion estimate from the RNG, the same data and model can give
  materially different response-niche p-values (and `tau2` estimates) from run to
  run. Set a seed before fitting — as done above and in the setup chunk — if you
  need reproducible variance components. (Raising `fitNB`'s `maxn.psi` to cover
  all cells removes the subsampling entirely and makes the fit deterministic, at
  some cost.)
* **`random = "slope"` is the less stable model at small `S`.** Its extra
  per-sample slope variance is collinear with the tested fixed effect and is
  estimated from only `S` samples, so it is heavy-tailed and, per dataset, tends
  to be a little more anti-conservative than `"intercept"` (visible as the slope
  curve sitting slightly above intercept). Both converge to nominal as `S` grows
  (Part A); flooring or clamping `tau2` does *not* fix the small-`S` gap. Prefer
  `"intercept"` when you have few samples and no reason to expect the niche-slope
  itself to vary between patients.

# Part B — power vs samples and vs cells

Now we plant a real effect (`sim_alt()`, `beta = 2`, 15% of genes true). Power is the fraction of the `nonnull_genes` recovered as raw-p `< 0.05`. We sweep `S` at fixed `n_per`, and `n_per` at fixed `S`, holding the effect size fixed in both sweeps (unlike the toy dataset, the same `beta`/`seed` structure is reused, only `S`/`n_per` change — no confound).

```{r partB-run}
S_grid <- if (RUN_FULL) c(4, 8, 12, 20, 30) else c(6, 12, 24)
npc_grid <- if (RUN_FULL) c(40, 80, 160, 320, 640) else c(80, 200, 400)

power_curve <- function(grid, vary = c("S", "n_per"), S_fixed = 10, npc_fixed = 90) {
  vary <- match.arg(vary)
  res <- bplapply(grid, function(v) {
    S <- if (vary == "S") v else S_fixed
    npc <- if (vary == "n_per") v else npc_fixed
    reps <- lapply(seq_len(B), function(b) {
      sim <- sim_alt(S = S, n_per = npc, seed = 2000 * v + b)
      vapply(methods, function(m) {
        p <- run_one(sim$spe, m)$raw_p
        mean(p[sim$nonnull_genes] < 0.05)
      }, numeric(1))
    })
    reps <- do.call(rbind, reps)
    data.frame(value = v, method = names(methods), power = colMeans(reps))
  }, BPPARAM = BPPARAM)
  do.call(rbind, res)
}

partB_S <- power_curve(S_grid, vary = "S")
partB_npc <- power_curve(npc_grid, vary = "n_per")
```

```{r partB-plot-S, fig.cap = "Power to detect the planted A-index/B-niche effect vs number of samples, at fixed cells/sample. Power rises for the mixed-effects methods as S grows."}
ggplot(partB_S, aes(value, power, colour = method)) +
  geom_line() + geom_point(size = 2) + ylim(0, 1) +
  labs(x = "number of samples (S)", y = "power (P(raw p < 0.05 | true effect))") +
  theme_bw()
```

```{r partB-plot-npc, fig.cap = "Power vs cells per sample, at fixed S. Under the mixed-effects correction power is essentially flat in cell count -- extra within-sample cells cannot substitute for extra samples, because the response label is constant within a sample."}
ggplot(partB_npc, aes(value, power, colour = method)) +
  geom_line() + geom_point(size = 2) + ylim(0, 1) +
  labs(x = "cells per sample", y = "power (P(raw p < 0.05 | true effect))") +
  theme_bw()
```

The contrast between the two panels is the whole answer to the motivating question: power climbs with `S` but is flat (or only weakly increasing, via the dispersion estimate) with `n_per`. The fixed-effects curve looks "more powerful" in both panels, but Part A already showed that its apparent power is inflated by the same anti-conservative bias that broke its null calibration -- comparing raw power across methods is only meaningful once Type-I error is matched, which is what Part C does properly via realized FDR.

# Part C — FDR control end-to-end

Using the full pipeline (`results()` gated on `fdr.niche`) with many genes (mostly null), we check whether the *realized* false discovery proportion (FDP) tracks the nominal FDR threshold, and how many of the true positives are recovered (TPR) at each threshold.

```{r partC-run}
fdr_grid <- if (RUN_FULL) c(0.01, 0.05, 0.1, 0.2, 0.3) else c(0.05, 0.2)
S_fdr <- if (RUN_FULL) 16 else 12

partC <- bplapply(fdr_grid, function(target) {
  reps <- lapply(seq_len(B), function(b) {
    sim <- sim_alt(S = S_fdr, n_genes = 150, prop_nonnull = 0.15, beta = 2,
                   seed = 3000 * (target * 100) + b)
    vapply(methods, function(m) {
      spe <- buildNiches(sim$spe, sigma = sigma_fixed)
      fit <- fitSpiDE(spe, "condition", sigma = sigma_fixed, random = m,
                      backend = "cpu", verbose = FALSE)
      res <- testSpiDE(fit, spe = spe, fdr = target)
      tab <- results(res)
      hit <- tab$gene[tab$ct_index == "A" & tab$ct_niche == "B" &
                        tab$fdr.niche <= target]
      hit <- unique(hit)
      is_fp <- !(hit %in% sprintf("G%d", sim$nonnull_genes))
      fdp <- if (length(hit) > 0) mean(is_fp) else 0
      tpr <- mean(sprintf("G%d", sim$nonnull_genes) %in% hit)
      c(fdp = fdp, tpr = tpr)
    }, numeric(2))
  })
  reps <- simplify2array(reps) # 2 x method x B
  data.frame(
    target = target, method = rep(names(methods), 2),
    metric = rep(c("fdp", "tpr"), each = length(methods)),
    value = c(rowMeans(reps["fdp", , ]), rowMeans(reps["tpr", , ]))
  )
}, BPPARAM = BPPARAM)
partC <- do.call(rbind, partC)
```

```{r partC-plot, fig.cap = "Realized false discovery proportion (top) and true-positive rate (bottom) vs the nominal FDR threshold. The dashed diagonal in the top panel is the y=x reference: points below it mean FDR is controlled; above it means it is exceeded."}
p1 <- ggplot(subset(partC, metric == "fdp"), aes(target, value, colour = method)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_line() + geom_point(size = 2) +
  labs(x = "nominal FDR", y = "realized FDP") + theme_bw()
p2 <- ggplot(subset(partC, metric == "tpr"), aes(target, value, colour = method)) +
  geom_line() + geom_point(size = 2) + ylim(0, 1) +
  labs(x = "nominal FDR", y = "TPR (recall)") + theme_bw()
p1
p2
```

# Part D — intercept vs slope: the few-responders trade-off

The model vignette notes a trade-off: `random = "slope"` gives the fullest protection against pseudo-replication, but with few samples the random slope can absorb a genuine effect, because both describe between-sample variation in the same niche-dependent slope. We reproduce that here by shrinking `S` at fixed effect size and comparing `"intercept"` vs `"slope"`.

```{r partD-run}
S_small_grid <- if (RUN_FULL) c(4, 6, 8, 12) else c(4, 8)

partD <- bplapply(S_small_grid, function(S) {
  reps <- lapply(seq_len(B), function(b) {
    sim <- sim_alt(S = S, n_genes = 60, prop_nonnull = 0.15, beta = 2.5,
                   sd_slope = 0.4, seed = 4000 * S + b)
    vapply(c(intercept = "intercept", slope = "slope"), function(m) {
      p <- run_one(sim$spe, m)$raw_p
      mean(p[sim$nonnull_genes] < 0.05)
    }, numeric(1))
  })
  reps <- do.call(rbind, reps)
  data.frame(S = S, method = c("intercept", "slope"), power = colMeans(reps))
}, BPPARAM = BPPARAM)
partD <- do.call(rbind, partD)
```

```{r partD-plot, fig.cap = "Power of random='intercept' vs random='slope' as the number of samples shrinks, with genuine per-sample slope heterogeneity present. With few samples, the random slope can absorb part of the true effect, making 'intercept' the more powerful (if less fully protective) default; the gap narrows as S grows."}
ggplot(partD, aes(S, power, colour = method)) +
  geom_line() + geom_point(size = 2) + ylim(0, 1) +
  labs(x = "number of samples (S)", y = "power") + theme_bw()
```

# Discussion

Putting the four pieces together answers the original question -- "I increased samples and cells and power didn't move":

1.  **Cells cannot buy power under the mixed-effects correction, by design** (Part B, right panel). The response label is constant within a sample, so extra cells are pseudo-replicates; the Wald test for the response-niche effect is evaluated in the between-sample stratum (`df = S - 2`, and standard errors that floor at the between-sample variance scale), matching the between-sample degrees of freedom documented in the model vignette.
2.  **Samples do buy power** (Part B, left panel), and this study isolates that cleanly because `beta`/`sd_patient`/seeds are held structurally fixed across the sweep -- something the bundled toy generator does not do, since it draws its entire dataset (coordinates, cell types, baselines, noise) from one `set.seed()` call that depends on `n_samples`/`n_per` themselves.
3.  **The fixed-effects fit's apparent extra "power" is not real power** -- Part A shows it is badly anti-conservative under the null regardless of sample size, so any power comparison against it is unfair until Type-I error is matched; Part C's realized-FDR curves make the honest comparison and show the mixed-effects fits controlling FDR near the nominal level while the fixed-effects fit overshoots.
4.  **Gene-level gating can hide a real, well-powered niche-level hit.** In the original toy example the gene-level FDR was pinned near 1 (Brown-combining one strong interaction with several null ones, then BH over only \~20 genes) even when the niche-level test (`fdr.niche`) was already strongly significant. This study reports power/FDR at the niche level throughout to avoid that dilution; Part C confirms the same gating still yields controlled FDR when there are enough true positives to estimate it.
5.  **With very few samples, `random = "slope"` can cost power relative to `"intercept"`** (Part D), because a real per-sample slope effect is statistically hard to distinguish from slope noise when there are only a handful of patients to average over.

## Scaling up this study

The chunks above default to a small grid (`RUN_FULL <- FALSE`) so the vignette renders quickly. To reproduce a publication-quality version, set `RUN_FULL <- TRUE` at the top (finer grids, `B = 40` replicate seeds) and raise `NPAR` to the number of available cores; expect a runtime of tens of minutes to a few hours depending on hardware, dominated by the `S/n_per` sweeps in Parts B-C. All simulation parameters (`sd_patient`, `beta`, `sd_slope`, `n_genes`, `prop_nonnull`) are exposed as generator arguments and can be swept further to probe other regimes (e.g. weaker effects, higher between-sample noise).

# Session info

```{r sessioninfo}
sessionInfo()
```