---
title: "Speeding up the mixed-effects fit: the PQL subsampling trade-off"
author: "Dharmesh D. Bhuva"
package: spiDE
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Speeding up the mixed-effects fit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4, dpi = 90
)
set.seed(1)
```

# Overview

The mixed-effects fit (`fitSpiDE(..., random = "intercept" | "slope")`) corrects
the anti-conservative inference that cell-level pseudo-replication causes (see
the *mixed-effects model* vignette). It does so with a Schall/PQL loop that
re-estimates the shared random-effect variance components `tau2`, and that loop
is the expensive part: it re-fits the negative-binomial GLM once per iteration.

The variance components are a *single shared scalar per random-effect group* — a
population quantity. Estimating them does not need every cell, exactly as
`SpaNorm::fitNB` already subsamples cells for its dispersion step. `spiDE`
exposes three knobs that trade a little `tau2` accuracy for a large speed-up:

| argument       | default | effect |
|----------------|---------|--------|
| `re.prop`      | `0.1`   | fraction of cells sampled **per cell type within each sample** for the PQL loop |
| `re.maxit.psi` | `1`     | dispersion iterations inside each PQL step (the final fit uses full dispersion) |
| `re.min.cells` | `100`   | per-stratum floor for `re.prop` |

Crucially the loop only supplies `tau2`; a single **final fit on all cells with
full dispersion** produces the coefficients and dispersion that inference uses.
So subsampling perturbs only the shared variance estimate, not the per-gene
effect sizes. This vignette quantifies the time saved and the accuracy cost.

```{r libs, message = FALSE}
library(spiDE)
library(ggplot2)
```

# Benchmark data

We generate a data set large enough that subsampling actually engages (strata
above the 100-cell floor), with both a planted patient-level random intercept
(so `tau2 > 0`) and a recoverable `response x niche` signal on gene `G1` in index
cell type `A` (so the response-niche t-statistics have a real range to compare).

```{r makedata}
makeBenchData <- function(n_samples = 6, n_per = 600, n_genes = 80,
                          field = 500, sd_patient = 0.6, beta = 1.2, seed = 1) {
  set.seed(seed)
  cts <- c("A", "B", "C")
  genes <- sprintf("G%d", seq_len(n_genes))
  sids <- sprintf("S%d", seq_len(n_samples))
  cond <- rep(c("Responder", "Non-responder"), length.out = n_samples)
  names(cond) <- sids

  cells <- lapply(sids, function(s) {
    x <- runif(n_per, 0, field); y <- runif(n_per, 0, field)
    ct <- ifelse(x > 0.6 * field & runif(n_per) < 0.7, "B",
                 sample(c("A", "C"), n_per, replace = TRUE))
    data.frame(sample_id = s, condition = cond[[s]], x = x, y = y,
               cell_type = ct, stringsAsFactors = FALSE)
  })
  cd <- do.call(rbind, cells)
  cd$cell_id <- sprintf("cell%d", seq_len(nrow(cd)))

  base <- matrix(rnorm(n_genes * 3, 1.5, 0.6), n_genes,
                 dimnames = list(genes, cts))
  lmu <- base[, cd$cell_type, drop = FALSE]
  # patient-level random intercept shared by every cell of a sample
  u <- matrix(rnorm(n_genes * n_samples, 0, sd_patient), n_genes,
              dimnames = list(genes, sids))
  lmu <- lmu + u[, cd$sample_id, drop = FALSE]
  # planted response x niche signal: G1 up in A cells, in Responders, with x
  is_A <- cd$cell_type == "A"; is_r <- cd$condition == "Responder"
  lmu["G1", ] <- lmu["G1", ] + beta * is_A * is_r * (cd$x / field)

  counts <- matrix(rnbinom(length(lmu), mu = exp(as.vector(lmu)), size = 5),
                   n_genes, dimnames = list(genes, cd$cell_id))
  cd$Area <- exp(rnorm(nrow(cd), sd = 0.2)); cd$nCount <- colSums(counts)
  SpatialExperiment::SpatialExperiment(
    assays = list(counts = counts),
    colData = S4Vectors::DataFrame(cd),
    spatialCoords = as.matrix(cd[, c("x", "y")]))
}

spe <- buildNiches(makeBenchData(), sigma = 30, verbose = FALSE)
dim(spe)
table(spe$sample_id, spe$cell_type)
```

We fix `backend = "cpu"` throughout so the timings reflect compute alone. (With
`backend = "auto"` each `fitNB` call also pays a one-off GPU-probe cost that is
independent of the model and would obscure the comparison on a GPU-less machine.)

# A single fit-and-test helper

```{r helper}
runOne <- function(spe, re.prop, re.maxit.psi, seed = 1) {
  set.seed(seed)
  t <- system.time(
    fit <- fitSpiDE(spe, "condition", sigma = 30, random = "slope",
                    backend = "cpu", re.prop = re.prop,
                    re.maxit.psi = re.maxit.psi, verbose = FALSE)
  )[["elapsed"]]
  res <- testSpiDE(fit, spe = spe, verbose = FALSE)
  f <- fits(res)[[1]]
  rn <- as.character(f@covtype) == "ResponseNiche"
  list(
    time = t,
    tau2 = f@tau2,
    tstat = f@t_stat[, f@coefmap$covariate[rn], drop = FALSE],
    # gene-level combined p for G1 (the planted hit) and its rank
    p_G1 = res@p.cauchy.pos["G1", "Gene"],
    rank_G1 = rank(res@p.cauchy.pos[, "Gene"])[["G1"]]
  )
}
```

# Time versus accuracy across `re.prop`

The reference is the fully accurate configuration (`re.prop = 1`, full
dispersion). Each subsampled fit is compared against it: we report the wall-clock
speed-up, the correlation of the response-niche t-statistics, and whether the
planted `G1` hit stays top-ranked.

```{r sweep}
ref <- runOne(spe, re.prop = 1, re.maxit.psi = 25)

props <- c(1, 0.5, 0.25, 0.1, 0.05)
sweep <- lapply(props, function(p) {
  o <- if (p == 1) runOne(spe, 1, 1) else runOne(spe, p, 1)
  data.frame(
    re.prop = p,
    time = o$time,
    speedup = ref$time / o$time,
    tstat_cor = cor(as.vector(o$tstat), as.vector(ref$tstat)),
    tau2_int = o$tau2[["SampleInt"]],
    tau2_slope = o$tau2[["SampleSlope"]],
    rank_G1 = o$rank_G1
  )
})
sweep <- do.call(rbind, sweep)
knitr::kable(sweep, digits = 3, row.names = FALSE)
```

```{r plot-time, fig.cap = "Wall-clock time falls sharply as cells are subsampled; the fully-accurate reference is the leftmost point."}
ggplot(sweep, aes(re.prop, time)) +
  geom_line(colour = "grey60") +
  geom_point(aes(colour = factor(re.prop)), size = 3) +
  scale_x_continuous(trans = "reverse") +
  labs(x = "re.prop (cell fraction, reversed)", y = "fit time (s)",
       colour = "re.prop") +
  theme_bw()
```

```{r plot-acc, fig.cap = "Response-niche t-statistics stay almost perfectly correlated with the full-data reference across the whole sampling range."}
ggplot(sweep, aes(re.prop, tstat_cor)) +
  geom_line(colour = "grey60") +
  geom_point(aes(colour = factor(re.prop)), size = 3) +
  scale_x_continuous(trans = "reverse") +
  coord_cartesian(ylim = c(min(0.9, min(sweep$tstat_cor)), 1)) +
  labs(x = "re.prop (cell fraction, reversed)",
       y = "t-stat correlation vs reference", colour = "re.prop") +
  theme_bw()
```

Accuracy stays high because the coefficients come from the full final fit;
`re.prop` only moves the shared `tau2`, and `tau2` moves little (below).

# Sampling (RNG) variation at the default `re.prop = 0.1`

Because no seed is set internally, repeated runs subsample different cells. We
quantify how much that jitters the estimates by repeating the default fit over
several seeds. Small spread here is the justification for keeping sampling on by
default.

```{r rng}
seeds <- 1:10
rng <- lapply(seeds, function(s) {
  o <- runOne(spe, re.prop = 0.1, re.maxit.psi = 1, seed = s)
  data.frame(seed = s, tau2_int = o$tau2[["SampleInt"]],
             tau2_slope = o$tau2[["SampleSlope"]], rank_G1 = o$rank_G1,
             tstat_cor = cor(as.vector(o$tstat), as.vector(ref$tstat)))
})
rng <- do.call(rbind, rng)
knitr::kable(rng[1:5, ], digits = 4, row.names = FALSE)

data.frame(
  metric = c("tau2 (intercept)", "t-stat cor vs ref"),
  mean = c(mean(rng$tau2_int), mean(rng$tstat_cor)),
  sd = c(sd(rng$tau2_int), sd(rng$tstat_cor)),
  cv = c(sd(rng$tau2_int) / mean(rng$tau2_int), NA)
) |> knitr::kable(digits = 4)
```

```{r plot-rng, fig.width = 6, fig.height = 3.5, fig.cap = "Across random subsamples the intercept variance component is tightly clustered around the reference (dashed line)."}
ggplot(rng, aes(x = "re.prop = 0.1", y = tau2_int)) +
  geom_boxplot(width = 0.3, fill = "grey90") +
  geom_jitter(width = 0.05, size = 2, colour = "steelblue") +
  geom_hline(yintercept = ref$tau2[["SampleInt"]], linetype = 2,
             colour = "firebrick") +
  labs(x = NULL, y = expression(tau^2 ~ "(SampleInt)")) +
  theme_bw()
```

# Summary

* Subsampling cells for the PQL loop (`re.prop`) plus a single inner dispersion
  iteration (`re.maxit.psi = 1`) cuts the mixed-effects fit time substantially
  while the final all-cell fit keeps the coefficients exact.
* Response-niche t-statistics and the planted-hit ranking are essentially
  unchanged across `re.prop`, and the shared `tau2` varies little across random
  subsamples — so `re.prop = 0.1` is a safe default.
* Set `re.prop = 1` for a fully reproducible (seed-independent) fit, at the cost
  of speed. Set an external `set.seed()` to make the subsampled fit reproducible.

```{r session}
sessionInfo()
```
