---
title: "Neighbourhood-dependent differential expression with spiDE"
author: "Dharmesh D. Bhuva"
package: spiDE
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Neighbourhood-dependent differential expression with spiDE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

`spiDE` finds **neighbourhood-dependent differential expression** in spatial
transcriptomics. Within an *index* cell type, it asks whether a gene's
expression changes with the local density (the *niche*) of another cell type.
One hypothesis is one (gene, index type, niche type) *triplet*, and spiDE asks
two questions of it:

- **the pooled test**: within patients, does the gene's expression in the
  index type change with the density of the niche type, consistently across
  patients?
- **the condition-specific test**: does that dependence differ between two
  conditions (responders and non-responders, early and late stage)?

Both are questions about patients. The condition is a property of a patient,
not of a cell, so spiDE takes the patient as its unit of replication: a
triplet's evidence is the variation of its niche slope *between* patients,
never the number of cells.

Both tests look at slopes *within* patients. Patients whose tissue holds more
of a niche type may also differ as a whole; `testNicheAbundance()` tests that
between-patient association separately. A condition that only shifts a gene's
overall level, or only changes how much of a cell type patients have, is not
tested by spiDE at all: those are questions for pseudobulk differential
expression and for composition analysis. The model vignette's section *What
kinds of effects can spiDE find?* draws each kind of effect on mock data and
names the test that detects it.

Three steps:

1. **Niches**: `buildNiches()` evaluates, per sample, a Gaussian kernel
   density of every cell type at every cell.
2. **Fit**: `fitSpiDE()` fits, for every index cell type, one negative
   binomial GLM per gene of its expression on the niche densities, with an
   intercept per patient, so each niche coefficient is a *within-patient*
   slope.
3. **Test**: `testSpiDE()` tests those slopes across patients. It always runs
   the pooled test, and runs the condition-specific test when it is given a
   condition.

`spiDE()` runs all three. The model is stated in full in the *model*
vignette (`vignette("spiDE-model")`), and what its calibration rests on in
the *calibration* vignette (`vignette("spiDE-calibration")`).

# Data

spiDE needs a `SpatialExperiment` with

- **raw integer counts** (the `counts` assay), not normalised or
  back-transformed values;
- a **cell type** per cell (`cell_type`);
- a **patient** per cell (`sample_id`). One sample per patient: if a patient
  has several sections, give the patient in `sample_id` and the section in
  `section`, so the spatial variance is estimated per section;
- for the condition-specific test, a **two-level condition that is constant
  within a patient**.

The shipped example has 16 patients (eight responders, eight
non-responders), three cell types and 20 genes. It plants one effect: gene
`G1` is up-regulated in index type `A`, in responders only, in proportion to
the local density of niche type `B`.

```{r data, message = FALSE}
library(spiDE)
library(SpatialExperiment)
data(toySpiDE)
toySpiDE
patients <- unique(as.data.frame(colData(toySpiDE))[, c("sample_id", "condition")])
table(patients$condition)
table(toySpiDE$cell_type)
```

# Niches

`buildNiches()` computes the niche covariates at one or more bandwidths
`sigma` (in the units of the spatial coordinates) and stores each as a
`reducedDim` named `Niche<sigma>`:

```{r niches}
spe <- buildNiches(toySpiDE, sigma = 30, verbose = FALSE)
head(reducedDim(spe, "Niche30"), 3)
```

The bandwidth sets the scale of a neighbourhood. Fit one bandwidth at a time;
comparing results across bandwidths is a sensitivity analysis, not a
multiplicity-free search. `mergeNiches()` pools niche columns into groups
(e.g. all lymphocytes) before fitting, and spiDE never tests an index type
against its own niche, merged or not.

# Fit

```{r fit}
fit <- fitSpiDE(spe, condition = "condition", sigma = 30, verbose = FALSE)
fit
```

`fitSpiDE()` fits every index type with at least `min.cells` cells in at
least `min.patients` patients (`index` restricts it), and tests every other
cell type as a niche (`niche` restricts it; the others stay in the model as
adjustments). Genes detected in fewer than `min.detect` of an index type's
cells are not tested in it. Sequencing depth enters as the centred log
library size of each cell (`depth = "loglib"`); cell-level covariates are
added with `covariates`.

## The two engines

`engine` chooses how the between-patient error is estimated:

- **`"slopes"`** (the default) fits the model once, without the condition,
  then estimates every patient's own niche slopes, each with a variance that
  allows for spatially correlated cells. `testSpiDE()` combines them across
  patients, allowing the slope to vary between patients. Because the fit
  never sees the condition, any patient-level condition can be tested
  afterwards, and a permutation of the condition reruns only the test.
- **`"sandwich"`** fits the condition-specific model directly and tests its
  coefficients with a patient-clustered sandwich standard error. It needs the
  condition at fit time.

The slopes engine is the more powerful, most of all when effects vary between
patients, and it is needed for `spiGSEA()`. The model vignette explains both
engines, and the calibration vignette shows how they hold up on real data.

## Each patient's slopes

The slopes engine keeps every patient's slope, which is the evidence the
tests weigh. `plotPatientSlopes()` draws them, one column per gene: each
point is a patient's slope with its interval, sized by the patient's weight
in the tests; the grey band is the pooled estimate, and each condition's
weighted mean is a coloured line.

```{r slopes, fig.width = 7, fig.height = 5}
ps <- patientSlopes(fit, gene = "G1", index = "A", niche = "B")
head(ps[, c("patient", "condition", "slope", "var_spatial", "ncells")])
res <- testSpiDE(fit)
plotPatientSlopes(res, gene = c("G1", "G2"), index = "A", niche = "B")
```

# Test

```{r test}
res
head(results(res, test = "pooled"))
```

With a condition (here recorded by `fitSpiDE()`; it can also be given to
`testSpiDE()`), the condition-specific test runs as well. The contrast is the
second level minus the first:

```{r condition}
res$contrast
cond <- results(res, test = "condition")
cond[cond$in_family, ]
```

By default (`procedure = "filtered"`) the condition-specific test is carried
out only among the triplets whose pooled test passes at `fdr`: the pooled
test never looks at the condition, so the filter is exact under any
relabelling of patients, and the multiplicity correction is spent on
triplets that have a niche dependence to modify. Rows outside that family
have `q = NA`. `procedure = "all"` tests every triplet.

The planted triplet, G1 in A against B, is the strongest pooled association
(`t` = `r round(results(res, test = "pooled")$t[1], 2)`), and in the
condition-specific test its slope is larger in responders by
`r round(cond$estimate[cond$gene == "G1" & cond$index == "A" & cond$niche == "B"], 2)` log units
per unit `log1p` density (`q` =
`r signif(cond$q[cond$gene == "G1" & cond$index == "A" & cond$niche == "B"], 2)`). `spiDE()` runs the three steps in one call:

```{r spiDE, eval = FALSE}
res <- spiDE(toySpiDE, condition = "condition", sigma = 30)
```

# Gene sets

`spiGSEA()` tests gene sets the way `testSpiDE()` tests genes. Each patient
gets a slope for the set: the average of its genes' slopes, each gene put on a
common scale. By default the average over all other tested genes is
subtracted, so a set is called when it moves *more than the typical gene*
(`type = "competitive"`). The set slopes then go through the pooled and the
condition-specific test, so the correlation between genes of a set is
accounted for without being estimated. Any named list of gene symbols works,
for example MSigDB collections from the `msigdb` package:

```{r gsea}
sets <- list(planted = c("G1", "G2", "G3"), other = paste0("G", 4:8),
             rest = paste0("G", 9:20))
gs <- spiGSEA(res, sets, min.size = 3)
head(gs[, c("set", "index", "niche", "test", "size", "estimate", "p", "q", "q.global")])
```

`q` adjusts within each (index, niche, test) column and `q.global` over every
set, index and niche of a test. With many sets and many columns, `q.global`
is the one to read. `plotTripletHeatmap()` shows the sets' t across every
(index, niche) column, a dot where `q.global` passes:

```{r gsea-plot, fig.width = 6, fig.height = 3}
plotTripletHeatmap(gs, test = "pooled")
```

# A different question: niche abundance

spiDE's slopes are within-patient. A patient whose index cells sit among many
niche cells may also express a gene differently, as a whole, from a patient
whose index cells do not; that between-patient association is absorbed by
the patient intercepts and is invisible to `testSpiDE()`. It is real, and it
is a different claim: that patients with more of a niche type differ, not
that a cell responds to its neighbours. `testNicheAbundance()` asks it on
pseudobulk profiles, one per patient and index type, against the patient's
mean niche density:

```{r abundance}
ab <- testNicheAbundance(spe, condition = "condition", sigma = 30, verbose = FALSE)
head(ab)
```

With a condition, the rows with `term` `"niche"` give the association averaged
over the two conditions, and those with `"condition:niche"` test whether it
differs between them (second level minus first). Each term is its own
multiple-testing family (`q.global`), and a patient who alone carries most of
a niche type's spread is down-weighted (`max.leverage`), so that one patient
cannot make a call on its own.

Never report a result of one as a result of the other.

# Plots

Every plot is a ggplot2 object in one house style (`theme_spiDE()`) and one
colour-blind-safe palette (`spiDEColours()`): amber and violet for a slope that
rises or falls with the niche's density, teal and rufous for the first and
second level of the condition. Each returns its data in `p$data`, so a plot is
changed by adding layers or a theme, and any two combine with patchwork.

Where the calls are, and the strongest of them. With `spe`, a triangle marks a
gene several-fold higher in the niche type than in the index type:

```{r plots-results, fig.width = 7, fig.height = 3.5}
plotCallMap(res, test = "pooled")
plotTopTriplets(res, test = "pooled", n = 12, spe = spe)
```

The cells behind a slope. By default each patient's fitted intercept is
divided out, so the patients share one level and the slopes the tests compare
are visible; `adjust = FALSE` shows the raw data, where the trend through the
patients' means is the between-patient association of the previous section:

```{r plots-response, fig.width = 7, fig.height = 3.5}
plotNicheResponse(spe, res, gene = c("G1", "G2"), index = "A", niche = "B")
plotNicheResponse(spe, res, gene = "G1", index = "A", niche = "B", adjust = FALSE)
```

The niche covariate in one sample (the bandwidth is right when the field is
neither speckled nor flat), and whether the calls are the niche type's own
markers, as transcripts leaking across segmentation boundaries would make
them:

```{r plots-niche, fig.width = 7, fig.height = 3.5}
plotNicheMap(spe, niche = "B", sample = "S1", index = "A")
plotSpillover(res, spe)
```

Diagnostics: the p-values by expression band (a sanity check; the calibration
vignette has the real one), and what the patients' intercepts capture. The
intercepts absorb every difference between patients by design; with `niche`,
the same plot on the slopes shows whether a batch or slide structures them, in
which case it belongs in `strata`:

```{r plots-diagnostics, fig.width = 7, fig.height = 4}
plotPValues(res)
plotPatientEffects(res, index = "A")
```

Figures combine with patchwork:

```{r plots-patchwork, fig.width = 9, fig.height = 4, eval = requireNamespace("patchwork", quietly = TRUE)}
library(patchwork)
plotPatientSlopes(res, gene = "G1", index = "A", niche = "B") +
  plotNicheResponse(spe, res, gene = "G1", index = "A", niche = "B") +
  plot_annotation(tag_levels = "A")
```

# Practicalities

- **Parallelism.** Genes are fitted independently given an index type, so
  `BPPARAM` parallelises over genes. Forked workers run one BLAS thread each
  when `RhpcBLASctl` is installed; set `OMP_NUM_THREADS=1` in the job
  otherwise.
- **Counts must be integers.** A back-transformed normalised assay is not a
  count and is refused.
- **Objects from spiDE 0.99.22 and earlier** came from a different model and
  are refused by this version, with a message naming the archive that reads
  them.

# Session information

```{r session}
sessionInfo()
```
