---
title: "Combining correlated niche p-values: Brown's method vs Cauchy combination"
author: "Dharmesh D. Bhuva"
package: spiDE
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Brown vs Cauchy combination of niche p-values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE, collapse = TRUE, comment = "#>", message = FALSE,
  warning = FALSE, fig.width = 7, fig.height = 4.2, dpi = 150, fig.retina = 2
)
set.seed(1)
```

```{r libs}
library(spiDE)
library(ggplot2)
source(system.file("spide_palette.R", package = "spiDE"))
```

# Overview

Within a gene, `spiDE` tests several `celltype:condition:niche` (*ResponseNiche*)
coefficients — one per surrounding cell type — and must collapse them into a
single gene-level p-value and one p-value per index cell type. These coefficients
are **correlated**, because neighbouring cell-type densities are themselves
correlated, so the combination step has to remain calibrated under dependence.

`spiDE` offers two combiners, selected with the `combine` argument:

```{r api, eval = FALSE}
res <- spiDE(spe, condition = "condition", combine = "cauchy") # default
res <- spiDE(spe, condition = "condition", combine = "brown")  # alternative
```

* **Brown's method** is a correlation-aware generalisation of Fisher's method: it
  estimates the coefficient correlation matrix and adjusts the combined
  statistic's reference distribution accordingly. It is exact under multivariate
  normal dependence but needs a well-conditioned correlation estimate.
* **Cauchy combination** (ACAT) forms an average of tangent-transformed p-values.
  Because a weighted sum of Cauchy variates is again Cauchy *regardless of
  dependence*, its tail stays valid under arbitrary correlation, with no
  correlation matrix to estimate.

This vignette compares them in two settings: a controlled Monte-Carlo experiment
that feeds both combiners identical, exactly correlated p-values, and the full
`spiDE` pipeline on realistic simulated data.

# A controlled comparison

We draw $m$ correlated z-scores with exchangeable correlation $\rho$, convert
them to one-sided p-values, and combine. Under the null the combined p-value
should be Uniform(0, 1); we sweep the correlation $\rho$ and the number of
coefficients $m$.

```{r combiners}
# exchangeable-correlated normals via a shared common factor: cor = rho
rmvnorm_exch <- function(n, k, rho, mean = 0) {
  u <- rnorm(n)
  e <- matrix(rnorm(n * k), n, k)
  sqrt(rho) * u + sqrt(1 - rho) * e + mean
}
cauchy_combine <- function(P) {
  P <- pmin(pmax(P, 1e-15), 1 - 1e-15)
  1 - pcauchy(rowMeans(tan((0.5 - P) * pi)))
}
brown_combine <- function(P, R) {
  Rm <- poolr::mvnconv(R, target = "m2lp", side = 1, cov2cor = FALSE)
  apply(P, 1, function(p) poolr::fisher(p, side = 1, R = Rm,
                                        adjust = "generalized")$p)
}
```

```{r sim-grid}
nsim <- 3000
rhos <- c(0, 0.3, 0.6, 0.9)
ks   <- c(3, 5, 10)

simulate_null <- function(k, rho) {
  R <- matrix(rho, k, k); diag(R) <- 1
  P <- pnorm(rmvnorm_exch(nsim, k, rho), lower.tail = FALSE)
  data.frame(Brown = brown_combine(P, R), Cauchy = cauchy_combine(P))
}
null_res <- do.call(rbind, lapply(ks, function(k) do.call(rbind,
  lapply(rhos, function(rho) {
    d <- simulate_null(k, rho)
    data.frame(k = k, rho = rho,
               method = rep(c("Brown", "Cauchy"), each = nsim),
               p = c(d$Brown, d$Cauchy))
  }))))
```

## Both combiners stay calibrated under correlation

On a QQ plot a calibrated combiner lies on the diagonal. At $\rho = 0$ both do;
as correlation grows, Brown (handed the true correlation matrix) stays on the
diagonal while Cauchy bends slightly below it in the body — mildly conservative —
but the two converge again in the extreme tail that actually drives FDR control.

```{r qq, fig.height = 4.4, fig.cap = "Null QQ plots for m = 5 correlated p-values across four correlation levels. Both combiners are calibrated or conservative; neither is anti-conservative."}
qq_df <- do.call(rbind, by(
  subset(null_res, k == 5),
  list(null_res$method[null_res$k == 5], null_res$rho[null_res$k == 5]),
  function(d) { d <- d[order(d$p), ]
    d$expected <- (seq_len(nrow(d)) - 0.5) / nrow(d); d }))

ggplot(qq_df, aes(-log10(expected), -log10(p), colour = method)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = spide_pal$ref) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ paste0("rho = ", rho), nrow = 1) +
  scale_colour_spide("combine") +
  labs(x = expression(-log[10]("expected")),
       y = expression(-log[10]("observed")), colour = NULL,
       title = "Null calibration (m = 5 correlated p-values)")
```

## Type-I error is controlled at every correlation

```{r typeI, fig.cap = "Empirical type-I error at alpha = 0.05 versus correlation, for three values of m. Both combiners control error; Cauchy is marginally conservative as correlation approaches 0.9."}
typeI <- aggregate(p ~ method + k + rho, null_res, function(x) mean(x <= 0.05))
names(typeI)[names(typeI) == "p"] <- "typeI"
ggplot(typeI, aes(rho, typeI, colour = method)) +
  geom_hline(yintercept = 0.05, linetype = 2, colour = spide_pal$ref) +
  geom_line(linewidth = 0.8) + geom_point(size = 1.8) +
  facet_wrap(~ paste0("m = ", k), nrow = 1) +
  scale_colour_spide("combine") + ylim(0, 0.1) +
  labs(x = expression(rho), y = "type-I error (nominal 0.05)", colour = NULL,
       title = "Type-I error versus correlation")
```

## Power is essentially identical

We now shift two of the $m = 5$ columns by a common effect $\delta$ at a
challenging correlation $\rho = 0.6$ and measure the detection rate.

```{r power, fig.cap = "Power versus effect size. The two combiners are near-identical; Cauchy sacrifices no meaningful power for its distribution-free calibration."}
delta_grid <- seq(0, 3, by = 0.5)
k <- 5; rho <- 0.6; n_signal <- 2
power_res <- do.call(rbind, lapply(delta_grid, function(delta) {
  R <- matrix(rho, k, k); diag(R) <- 1
  mu <- c(rep(delta, n_signal), rep(0, k - n_signal))
  P <- pnorm(rmvnorm_exch(nsim, k, rho) + matrix(mu, nsim, k, byrow = TRUE),
             lower.tail = FALSE)
  data.frame(delta = delta, method = c("Brown", "Cauchy"),
             power = c(mean(brown_combine(P, R) <= 0.05),
                       mean(cauchy_combine(P) <= 0.05)))
}))
ggplot(power_res, aes(delta, power, colour = method)) +
  geom_line(linewidth = 0.8) + geom_point(size = 1.8) +
  scale_colour_spide("combine") +
  labs(x = expression(paste("effect size ", delta, " (2 of 5 columns)")),
       y = "power at alpha = 0.05", colour = NULL,
       title = "Power versus effect size (rho = 0.6)")
```

# In the full pipeline at realistic scale

The controlled experiment isolates the combiners; we also compare them inside the
complete `spiDE` pipeline on realistic simulated data (ten-thousand-plus genes,
low counts, a planted niche effect in 5% of genes). The table below, from the
benchmark study (`research/`), reports the realised false discovery proportion
and true-positive rate under each combiner at several nominal FDR thresholds.

```{r pipeline}
cb <- local({
  path <- system.file("extdata", "benchmark", "combiner.csv", package = "spiDE")
  if (file.exists(path)) utils::read.csv(path, check.names = FALSE) else NULL
})
```

```{r pipeline-plot, eval = !is.null(cb), fig.height = 4, fig.cap = "Full-pipeline comparison on realistic data. Left: realised FDP versus nominal FDR under the alternative (dashed y = x is the control boundary). Right: recall at each threshold."}
alt <- subset(cb, arm == "alt")
alt$combine <- factor(tools::toTitleCase(alt$combine),
                      levels = names(spide_pal$combine))
p_fdp <- ggplot(alt, aes(fdr, fdp, colour = combine)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = spide_pal$ref) +
  geom_line(linewidth = 0.8) + geom_point(size = 2) +
  scale_colour_spide("combine") +
  labs(x = "nominal FDR", y = "realised FDP", colour = NULL,
       title = "FDR control")
p_tpr <- ggplot(alt, aes(fdr, tpr, colour = combine)) +
  geom_line(linewidth = 0.8) + geom_point(size = 2) +
  scale_colour_spide("combine") +
  labs(x = "nominal FDR", y = "TPR (recall)", colour = NULL, title = "Recall")
patchwork::wrap_plots(p_fdp, p_tpr, guides = "collect") &
  ggplot2::theme(legend.position = "bottom")
```

```{r pipeline-note, echo = FALSE, results = "asis", eval = is.null(cb)}
cat("*(The full-pipeline summary table was not found; run the benchmark study",
    "in `research/` and `install_results.R` to populate it.)*\n")
```

In the full pipeline the two combiners diverge sharply, and in Cauchy's favour.
Cauchy recovers far more of the planted genes at every threshold — recall around
0.30 at a nominal 20 % FDR against roughly 0.02 for Brown — while Brown makes few
discoveries and a large fraction of them are false, its realised FDP sitting near
0.4 regardless of the threshold. Cauchy's own FDP does exceed the nominal level
at the *strictest* thresholds, where only a handful of genes are called and the
proportion is dominated by one or two false positives, but it falls below the
line as more true genes accumulate. Both combiners inherit their absolute FDP
level from the underlying mixed-effects fit; this comparison holds that fit fixed
and isolates the combination step, and on that axis Cauchy is clearly preferable.


# Sidedness: why Cauchy needs two-sided input

Everything above combines **one-sided** p-values, testing each direction
separately at `fdr/2` so that a gene can be labelled Up, Down or Both. That
convention is safe for Brown's method but **not** for Cauchy combination, and the
reason is the transform each uses.

| combiner | transform | value as $p \to 1$ |
|---|---|---|
| Brown / Fisher | $-2\log p$ | $\to 0$ — contributes nothing |
| Cauchy / ACAT | $\tan((0.5 - p)\pi)$ | $\to -\infty$ — cancels the signal |

Consider a gene that is strongly **up** near one niche cell type and strongly
**down** near another. In the up-direction combination the down coefficient
contributes a one-sided $p \approx 1$. Fisher's transform maps that to
$\approx 0$, so it is simply ignored. Cauchy's maps it to a large *negative*
number that annihilates the large positive contribution from the genuine signal.

```{r sided-terms}
d_grid <- c(3, 8, 15, 25)
data.frame(
  `effect (t)` = d_grid,
  `opposing one-sided p` = signif(pt(-d_grid, 50, lower.tail = FALSE), 4),
  `Cauchy term` = signif(tan((0.5 - pmin(pt(-d_grid, 50, lower.tail = FALSE),
                                         1 - 1e-15)) * pi), 3),
  `Fisher term` = signif(-2 * log(pmin(pt(-d_grid, 50, lower.tail = FALSE),
                                       1 - 1e-15)), 3),
  check.names = FALSE
)
```

The consequence is not subtle: the stronger the opposing signal, the more
completely Cauchy discards the gene.

```{r sided-sim}
k <- 5; df_t <- 50; nsim_s <- 2000
# gene-level p as the pipeline forms it: one-sided combiners test each direction
# and pay a factor-2 direction-split correction; the two-sided combiner does not.
gene_p <- function(T) {
  pp <- pt(T, df_t, lower.tail = FALSE)
  pn <- pt(T, df_t, lower.tail = TRUE)
  p2 <- pmin(2 * pmin(pp, pn), 1)
  R  <- diag(k)
  data.frame(
    `Brown (1-sided)`  = pmin(2 * pmin(brown_combine(pp, R),
                                       brown_combine(pn, R)), 1),
    `Cauchy (1-sided)` = pmin(2 * pmin(cauchy_combine(pp),
                                       cauchy_combine(pn)), 1),
    `Cauchy (2-sided)` = cauchy_combine(p2),
    check.names = FALSE
  )
}
draw <- function(delta, bidirectional) {
  T <- matrix(rt(nsim_s * k, df_t), nsim_s, k)
  T[, 1] <- T[, 1] + delta
  T[, 2] <- T[, 2] + if (bidirectional) -delta else delta
  T
}
deltas <- c(0, 2, 3, 4, 5, 6, 8, 10, 15)
power_df <- do.call(rbind, lapply(c(TRUE, FALSE), function(bi) {
  do.call(rbind, lapply(deltas, function(d) {
    g <- gene_p(draw(d, bi))
    data.frame(delta = d,
               signal = if (bi) "bidirectional (one up, one down)"
                        else "unidirectional (both up)",
               method = rep(names(g), each = nsim_s),
               detected = unlist(lapply(g, function(p) p < 0.05)))
  }))
}))
power_sum <- aggregate(detected ~ delta + signal + method, power_df, mean)
```

```{r sided-power, fig.height = 4.2, fig.cap = "Detection rate at alpha = 0.05 versus effect size, for m = 5 coefficients. With unidirectional signal all three agree. With opposing signal, one-sided Cauchy collapses toward zero as the effect grows, while Brown and two-sided Cauchy both rise monotonically."}
ggplot(power_sum, aes(delta, detected, colour = method)) +
  geom_hline(yintercept = 0.05, linetype = 2, colour = spide_pal$ref) +
  geom_line(linewidth = 0.8) + geom_point(size = 1.7) +
  facet_wrap(~ signal, nrow = 1) +
  scale_colour_spide("sided") + ylim(0, 1) +
  labs(x = expression(paste("effect size ", delta)),
       y = expression(paste("detection rate (", alpha, " = 0.05)")),
       colour = NULL, title = "Sidedness matters only for opposing effects")
```

The left panel is the diagnostic one. One-sided Cauchy is **non-monotone in the
effect size**: detection rises, peaks, and then falls back towards zero as the
opposing coefficient's p-value approaches 1. The clearest possible biology is the
most likely to be discarded, and silently. Brown is immune, and two-sided Cauchy
is immune while keeping the correlation-agnostic property that made Cauchy the
default.

Two-sided input costs nothing where the one-sided convention was already safe:
in the right-hand panel the three curves are indistinguishable, because the
two-sided p-value's factor of 2 is exactly offset by dropping the `fdr/2`
direction split.

```{r sided-null, fig.height = 3.6, fig.cap = "Null QQ plot for the three gene-level constructions. All are calibrated; two-sided Cauchy is not anti-conservative."}
null_sided <- gene_p(matrix(rt(nsim_s * k, df_t), nsim_s, k))
qq_s <- do.call(rbind, lapply(names(null_sided), function(m) {
  p <- sort(null_sided[[m]])
  data.frame(method = m, p = p, expected = (seq_along(p) - 0.5) / length(p))
}))
ggplot(qq_s, aes(-log10(expected), -log10(p), colour = method)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = spide_pal$ref) +
  geom_line(linewidth = 0.8) +
  scale_colour_spide("sided") +
  labs(x = expression(-log[10]("expected")),
       y = expression(-log[10]("observed")), colour = NULL,
       title = "Null calibration of the gene-level p-value")
```

```{r sided-typeI}
data.frame(
  method = names(null_sided),
  `type-I at 0.05` = signif(vapply(null_sided, function(p) mean(p < 0.05),
                                   numeric(1)), 3),
  check.names = FALSE, row.names = NULL
)
```

## What `spiDE` does

`combine = "cauchy"` combines **two-sided** p-values; `combine = "brown"` keeps
**one-sided** p-values. When the two-sided path is used the `fdr/2` direction
split is not applied (there is only one test), and Up / Down / Both are resolved
from the signs of the significant niche coefficients — which is where direction
is well defined in any case. Under the one-sided Cauchy convention the `Both`
label was effectively unreachable, because genes with opposing effects were
exactly the ones cancelled at the gene-level gate.

# Summary

On both the controlled experiment and the full pipeline, **Cauchy combination
matches or beats Brown's method**: it controls type-I error at every correlation
level, loses no meaningful power, and does so without estimating a correlation
matrix or depending on `poolr`. Its only cost is mild conservativeness in the
body of the null distribution under very strong dependence — irrelevant for the
extreme tail that FDR control depends on. For these reasons Cauchy combination is
the **default** (`combine = "cauchy"`); Brown's method remains available
(`combine = "brown"`) for settings where an explicit, correlation-aware combiner
is preferred.

That verdict holds **provided the Cauchy combiner is given two-sided p-values**.
Its $\tan$ transform diverges to $-\infty$ as $p \to 1$, so with one-sided
input a gene that is up near one niche cell type and down near another cancels
itself out, and does so more completely the stronger the effect. Brown's
$-2\log p$ is bounded at zero and has no such failure mode, which is why the
directional convention inherited from the Brown era is safe there and unsafe
here. `spiDE` therefore combines two-sided p-values under `combine = "cauchy"`
and one-sided under `combine = "brown"`.

```{r session}
sessionInfo()
```
