---
title: "Neighbourhood-dependent differential expression with spiDE"
author: "Dharmesh D. Bhuva"
package: spiDE
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Neighbourhood-dependent differential expression with spiDE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

`spiDE` identifies **context-specific, neighbourhood-dependent differential
expression** in spatial transcriptomics data. Within an *index* cell type, it
tests how gene expression changes with an experimental *condition* as a function
of the local density (the *niche*) of surrounding cell types.

The method proceeds in three steps:

1. **Niche construction** (`buildNiches()`): for each sample and each spatial
   bandwidth, a Gaussian kernel density estimate of every cell type is evaluated
   at each cell, giving per-cell "effective niche" densities.
2. **Model fitting** (`fitSpiDE()`): a per-gene negative binomial GLM is fit over
   a design containing the three-way `cell type : condition : niche`
   interactions, using the `r BiocStyle::Biocpkg("SpaNorm")` `fitNB()` engine.
3. **Testing** (`testSpiDE()`): per-gene Wald statistics for the neighbourhood
   interactions are combined across correlated covariates with Brown's method and
   across bandwidths with a Cauchy combination, then subjected to a hierarchical
   (gene &rarr; index cell type &rarr; niche cell type) FDR correction.

`spiDE()` wraps the three steps.

```{r load}
library(spiDE)
library(SpatialExperiment)
```

# Example data

The package ships a small synthetic `SpatialExperiment`, `toySpiDE`, with 20
genes across 480 cells from 6 samples (3 Responders, 3 Non-responders) and three
cell types (A, B, C). A neighbourhood signal is planted: gene `G1` is
up-regulated in index cell type **A**, in Responders, in proportion to the local
density of niche cell type **B**.

```{r data}
data(toySpiDE)
toySpiDE
table(toySpiDE$cell_type, toySpiDE$condition)
```

# The full workflow in one call

```{r full}
res <- spiDE(
  toySpiDE,
  condition = "condition",
  sigma = c(10, 30, 50),
  covariates = "Age",
  verbose = FALSE
)
res
```

The tidy results table is returned by `results()`, keyed by
(gene, index cell type, niche cell type, bandwidth):

```{r results}
tab <- results(res)
head(tab[order(tab$fdr.niche), ], 10)
```

The planted `G1` / index A / niche B effect is recovered with a positive
(up-regulated) direction:

```{r recovered}
tab[tab$gene == "G1" & tab$ct_index == "A" & tab$ct_niche == "B", ]
```

# Step by step

The wrapper is equivalent to running the three stages explicitly, which is
useful when you want to inspect intermediate objects, reuse a fit, or change
parameters at a single stage.

## 1. Build niches

`buildNiches()` adds one `reducedDim` per bandwidth. Optionally, fine cell types
can be merged into coarser niches with `mergeNiches()`, and per-sample
library-size offsets computed with `computeSizeFactors()`.

```{r build}
spe <- buildNiches(toySpiDE, sigma = c(10, 30, 50))
reducedDimNames(spe)
```

## 2. Fit the model

```{r fit}
fit <- fitSpiDE(spe, condition = "condition", covariates = "Age", verbose = FALSE)
fit
```

Each bandwidth's fit is a `SpiDEFit` object exposing the coefficient matrix
`alpha`, dispersions `psi`, the design `W`, and the covariate-type tags:

```{r fitobj}
f1 <- fits(fit)[[1]]
table(f1$covtype)
```

## 3. Test

```{r test}
res2 <- testSpiDE(fit, spe = spe, fdr = 0.05)
nrow(results(res2))
```

# Scaling

Fitting is performed over all genes at once (the negative binomial dispersion is
moderated across genes). The per-gene inference is independent across genes and
is processed in gene-blocks, dispatched with
`r BiocStyle::Biocpkg("BiocParallel")`; the counts matrix may be a
`r BiocStyle::Biocpkg("DelayedArray")` and is realised one block at a time.
Results are invariant to the block size and the parallel backend.

```{r parallel, eval = FALSE}
library(BiocParallel)
res <- spiDE(
  toySpiDE,
  condition = "condition",
  covariates = "Age",
  block.size = 500,
  BPPARAM = MulticoreParam(4)
)
```

The `backend` argument (`"auto"`, `"cpu"`, `"gpu"`) is forwarded to
`SpaNorm::fitNB()` and selects the compute backend for the model fit.

# Session info

```{r sessioninfo}
sessionInfo()
```
