---
title: "The spiDE model: neighbourhood-dependent DE and mixed-effects inference"
author: "Dharmesh D. Bhuva"
package: spiDE
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{The spiDE model and mixed-effects inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
```

# Overview

This vignette documents the statistical model behind `spiDE` (Spatial Influence DE) and a **mixed-effects extension** that corrects the anti-conservative inference caused by cell-level pseudo-replication. It has three parts:

1.  the full `spiDE` model — how the effective niche is built and how gene expression is modelled as a function of cell type, neighbourhood, and clinical response;
2.  why the default (fixed-effects) Wald test is anti-conservative when many cells come from few patients; and
3.  the mixed-effects correction, implemented through the **ridge = random effects** equivalence so it reuses the same `SpaNorm::fitNB` engine.

For a task-oriented quickstart, see the companion vignette *Neighbourhood-dependent differential expression with spiDE*.

```{r load, message = FALSE}
library(spiDE)
library(SpatialExperiment)
```

# The full spiDE model

## Notation

Let $Y_{c,g}$ be the (SpaNorm-normalised) expression of gene $g \in \{1,\dots,G\}$ in cell $c \in \{1,\dots,C\}$. Each cell has a type $t_c \in \{1,\dots,T\}$ and belongs to a patient tissue sample $s_c \in \{1,\dots,S\}$. The clinical response of a sample is $r_{s} \in \{0 = \text{non-responder},\, 1 = \text{responder}\}$; because it is a property of the *patient*, $r_{s}$ is **constant across all cells of a sample** — a fact that drives everything below.

## Effective niche

For each cell we summarise its local neighbourhood by a kernel-smoothed intensity of every surrounding cell type. With a Gaussian kernel $K_\sigma$ of bandwidth $\sigma$, the intensity of type $t'$ around cell $c$ (within its own sample $s$) is

$$
\Lambda_{c,t',s}(\sigma) \;=\; e(c) \sum_{c'=1}^{C}
  K_\sigma\!\left(d_{c,c'}\right)\,
  \mathbb{1}(t_{c'} = t')\,\mathbb{1}(s_{c'} = s),
$$

where $d_{c,c'}$ is the distance between cells and $e(c)$ is Diggle's edge correction. The expected log-count of type $t'$ in a disc of radius $\sigma$ is

$$
N_{c,t',s}(\sigma) \;=\; \log\!\left(\pi\sigma^2\,\Lambda_{c,t',s}(\sigma) + 1\right).
$$

Stacking over cell types gives the **effective niche composition** $N_{c,s}(\sigma) = \big(N_{c,1,s}(\sigma),\dots,N_{c,T,s}(\sigma)\big)$. Niches are computed per sample but then concatenated across all cells into a $C \times T$ matrix $N(\sigma)$. In the package this is `buildNiches()`, which stores $N(\sigma)$ as `reducedDim(spe, "Niche<sigma>")`.

## The negative binomial GLM

For a fixed bandwidth $\sigma$, gene $g$ is modelled with a negative binomial GLM with dispersion $\phi_g$ and log-mean

$$
\log \mathbb{E}\!\left(Y_{c,g}\mid s_c=s, t_c=t, r_s, N_{c,s}(\sigma)\right)
= \gamma_g L_s + \zeta_{g,t} + \eta_g\, r_s
 + \sum_{t'=1}^{T} N_{c,t',s}(\sigma)\,\alpha_{g,t,t'}
 + r_s \sum_{t'=1}^{T} N_{c,t',s}(\sigma)\,\beta_{g,t,t'} .
$$

The coefficients are:

| term | meaning |
|-----------------------------|-------------------------------------------|
| $\gamma_g L_s$ | sample-specific library-size effect (`computeSizeFactors()` offset) |
| $\zeta_{g,t}$ | cell-type-specific expression |
| $\eta_g$ | global response effect |
| $\alpha_{g,t,t'}$ | average effect of niche type $t'$ on gene $g$ in cell type $t$ |
| $\beta_{g,t,t'}$ | **response-specific** effect of niche type $t'$ in cell type $t$ |

The scientific quantity of interest is $\beta_{g,t,t'}$. We test

$$
H_0:\ \beta_{g,t,t'} = 0 \qquad\text{vs}\qquad H_1:\ \beta_{g,t,t'} \neq 0 .
$$

When $\beta_{g,t,t'} \neq 0$, niche cell type $t'$ has a response-specific association with the expression of gene $g$ in index cell type $t$.

In the package the design is assembled by `nicheDesign()` / `.buildNicheDesign()` as `~ 0 + <covariates> + CellType + condition*CellType:(niches) + niches`, and each column is tagged: the $\eta_g$ column is `"Response"` and the $\beta_{g,t,t'}$ (three-way `condition:CellType:niche`) columns are `"ResponseNiche"` — the ones carried through testing. The whole gene set is fit in one `SpaNorm::fitNB(Y, W, ...)` call (the dispersion $\phi_g$ is moderated across genes), and per-gene Wald statistics for $\eta_g$ / $\beta_{g,t,t'}$ are combined across correlated niches (Brown's method), across bandwidths (Cauchy combination), and corrected with a hierarchical FDR.

# The pseudo-replication problem

The GLM above treats each of the $C$ cells as an independent observation. That is fine for the within-sample coefficients ($\zeta$, $\alpha$), but the response-related coefficients $\eta_g$ and $\beta_{g,t,t'}$ are **contrasts on the patient-level label** $r_s$. There are only $S$ independent patients, yet the Wald information matrix

$$
\widehat{\mathrm{Var}}(\hat\theta) \;=\; \phi_g\left(W_\text{sub}^\top D_g\, W_\text{sub}\right)^{-1},
\qquad D_g = \mathrm{diag}\big(w_{c,g}\big),\ \ w_{c,g} = \frac{\mu_{c,g}}{1+\phi_g\mu_{c,g}},
$$

sums over all $C$ cells. The standard error of a response contrast therefore shrinks like $1/\sqrt{C}$ instead of $1/\sqrt{S}$, and with $C \gg S$ the Wald $z$-statistics are badly inflated. This reproduces the known behaviour of the original flat-script implementation and is documented as a property of the default fit.

We can see it directly. Below we simulate a **null** data set: every cell of a patient shares a per-patient random shift, but there is *no* response effect at all — so every response test should be null.

```{r sim}
sim_clustered <- function(n_samples = 10, n_per = 90, n_genes = 30,
                          field = 500, sd_patient = 0.7, seed = 1) {
  set.seed(seed)
  cts <- c("A", "B", "C"); gn <- sprintf("G%d", seq_len(n_genes))
  sids <- sprintf("S%d", seq_len(n_samples))
  cond <- rep(c("Responder", "Non-responder"), length.out = n_samples)
  names(cond) <- sids
  cells <- lapply(sids, function(sid) {
    x <- runif(n_per, 0, field); y <- runif(n_per, 0, field)
    ct <- ifelse(x > 0.6 * field & runif(n_per) < 0.7, "B",
                 sample(c("A", "C"), n_per, replace = TRUE))
    data.frame(sample_id = sid, condition = cond[[sid]],
               x = x, y = y, cell_type = ct)
  })
  cd <- do.call(rbind, cells); n <- nrow(cd)
  cd$cell_id <- sprintf("c%d", seq_len(n))
  base <- matrix(rnorm(n_genes * 3, 1.5, 0.6), n_genes, dimnames = list(gn, cts))
  lmu <- base[, cd$cell_type, drop = FALSE]
  # per-(gene, patient) random intercept: real clustering, NO response effect
  u <- matrix(rnorm(n_genes * n_samples, 0, sd_patient), n_genes,
              dimnames = list(gn, sids))
  lmu <- lmu + u[, cd$sample_id, drop = FALSE]
  counts <- matrix(rnbinom(length(lmu), mu = as.vector(exp(lmu)), size = 5),
                   n_genes, dimnames = list(gn, cd$cell_id))
  SpatialExperiment(assays = list(counts = counts),
                    colData = S4Vectors::DataFrame(cd),
                    spatialCoords = as.matrix(cd[, c("x", "y")]))
}

spe <- buildNiches(sim_clustered(), sigma = 30)
```

A helper to pull the two-sided p-values of the (null) response effect:

```{r respp}
response_p <- function(random) {
  f <- fitSpiDE(spe, "condition", sigma = 30, random = random, verbose = FALSE)
  res <- testSpiDE(f, spe = spe)          # populates the Wald inference slots
  ff <- fits(res)[[1]]
  rc <- ff$coefmap$covariate[as.character(ff$covtype) == "Response"]
  df <- if (is.null(ff$df)) Inf else ff$df
  2 * pt(-abs(ff$t_stat[, rc]), df)
}
p_fixed <- response_p("none")
mean(p_fixed < 0.05)  # should be ~0.05 under the null
```

The fixed-effects fit rejects a large fraction of these null tests at $\alpha = 0.05$ — grossly anti-conservative.

# Mixed-effects correction via ridge

## The model

We add **patient-level random effects** that soak up the between-patient variability the fixed-effects model wrongly attributes to residual (cell-level) noise. Crucially, we add them only for the *response-related* effects — the only ones we do inference on — by giving each response coefficient its patient-level random counterpart:

$$
\log \mathbb{E}(Y_{c,g}\mid\cdots)
= \underbrace{\gamma_g L_s + \zeta_{g,t}
 + \sum_{t'} N_{c,t',s}\,\alpha_{g,t,t'}}_{\text{fixed, within-sample}}
 + \underbrace{\eta_g\, r_s + r_s\!\sum_{t'} N_{c,t',s}\,\beta_{g,t,t'}}_{\text{fixed response effects (tested)}}
 + \underbrace{u_{g,s} + \sum_{t'} N_{c,t',s}\, v_{g,s,t,t'}}_{\text{patient random effects}},
$$

with

$$
u_{g,s} \sim \mathcal N(0,\tau^2_{\text{int}}),\qquad
v_{g,s,t,t'} \sim \mathcal N(0,\tau^2_{\text{slope}}).
$$

-   The random **intercept** $u_{g,s}$ is the patient-level counterpart of the response main effect $\eta_g$: it lets each patient's baseline float, so $\eta_g$ is identified from between-patient contrasts.
-   The random **slopes** $v_{g,s,t,t'}$ are the patient-level counterparts of the $\beta_{g,t,t'}$ terms: they let each patient's niche→expression slope float, so $\beta_{g,t,t'}$ is tested against between-patient *slope* variability. A random intercept alone would not protect $\beta$, because $\beta$ multiplies a covariate that varies *within* a patient.

`random = "intercept"` adds $u_{g,s}$ only; `random = "slope"` adds both. The
slope model gives the fullest protection of $\beta$, but note the trade-off: the
random slope $v_{g,s,t,t'}$ and the fixed effect $\beta_{g,t,t'}$ both describe
how the niche→expression slope differs across patients, so with **few patients
per response group** the random slope can absorb a genuine $\beta$ (a real effect
seen in only a handful of responders is, statistically, hard to distinguish from
patient-to-patient slope noise). With an adequate number of samples prefer
`"slope"`; with only a few, `"intercept"` is a more powerful default and still
corrects the dominant (baseline) source of pseudo-replication.

## Ridge $\equiv$ random effects

Write the expanded design as $\tilde W = [\,W \mid Z\,]$, where $Z$ holds the random-effect columns (one indicator per sample for $u$, and sample-indicator $\times$ `CellType:niche` columns for $v$). On the IRLS working scale a Gaussian random effect is *exactly* an $L_2$ (ridge) penalty applied only to its columns: maximising the penalised working likelihood

$$
\big\lVert z - \tilde W\theta \big\rVert^2_{D_g} \;+\; \theta^\top \Lambda\, \theta,
\qquad
\Lambda = \mathrm{blockdiag}\!\left(0_{\text{fixed}},\; \lambda_{\text{int}} I,\; \lambda_{\text{slope}} I\right),
$$

gives the BLUP/penalised estimate $\hat\theta = (\tilde W^\top D_g \tilde W +
\Lambda)^{-1}\tilde W^\top D_g z$, with the variance-component correspondence

$$
\boxed{\ \lambda_k \;=\; 1/\tau^2_k\ }.
$$

This is the standard penalised quasi-likelihood (PQL) construction used by `lme4` / `mgcv`. It fits `spiDE` perfectly: `SpaNorm::fitNB` already accepts a **per-column** ridge penalty `lambda.a`, so we set it to $0$ on the fixed columns and $1/\tau^2_k$ on the random ones — no change to the fitting engine.

## Estimating the variance components

`fitNB` treats `lambda.a` as fixed, so the variance components are estimated by an outer **Schall / PQL** loop (shared across genes, matching the dispersion moderation): alternate a penalised fit with a moment update

$$
\hat\tau^2_k \;=\; \frac{\sum_g \lVert \hat b_{g,k}\rVert^2}{G\,\big(q_k - \lambda_k\,\mathrm{tr}\, C_{kk}\big)},
\qquad C = (\tilde W^\top \bar D\, \tilde W + \Lambda)^{-1},
$$

where $\hat b_{g,k}$ are the BLUPs of block $k$, $q_k$ its size and $C_{kk}$ its sub-block, until $\tau^2$ converges. This is `.fitNBmixed()` in `R/fitSpiDE.R`.

## Corrected inference

With the random effects in place, three things change relative to the fixed-effects Wald test:

1.  **Covariance** uses the full penalised information and extracts the fixed response block, $\widehat{\mathrm{Var}}(\hat\theta_\text{resp}) = \varphi_g \big[(\tilde
    W^\top D_g \tilde W + \Lambda)^{-1}\big]_\text{resp}$. Because $Z$ absorbs the within-patient correlation, this block is inflated to a between-patient scale.
2.  **Dispersion** $\varphi_g$ is the working (Pearson) dispersion of the *random-effects* model, $\varphi_g = \frac{1}{C - p}\sum_c (Y_{c,g} -
    \mu_{c,g})^2 / (\mu_{c,g} + \phi_g\mu_{c,g}^2)$, rather than the raw NB dispersion.
3.  **Reference distribution** is a $t$ with a **between-patient** degrees of freedom $S - 2$ (the response has two levels), not a normal.

Together these test each response contrast against patient-level variability. Let us re-run the null simulation with the mixed-effects fit:

```{r mixed}
p_mixed <- response_p("intercept")
c(fixed = mean(p_fixed < 0.05), mixed = mean(p_mixed < 0.05))
```

```{r qq, fig.small = TRUE, fig.cap = "Null response-effect p-values. The fixed-effects fit is grossly anti-conservative; the mixed-effects fit is close to the uniform (dashed) reference."}
qqp <- function(p, ...) {
  points(sort(-log10(ppoints(length(p)))), sort(-log10(p)), ...)
}
plot(NA, xlim = c(0, 3), ylim = c(0, 6), xlab = "expected -log10 p",
     ylab = "observed -log10 p")
abline(0, 1, lty = 2)
qqp(p_fixed, pch = 16, col = "firebrick")
qqp(p_mixed, pch = 16, col = "steelblue")
legend("topleft", c("fixed", "mixed"), pch = 16,
       col = c("firebrick", "steelblue"), bty = "n")
```

The mixed-effects fit also **recovers the planted between-patient variance** ($0.7^2 = 0.49$):

```{r tau2}
fit <- fitSpiDE(spe, "condition", sigma = 30, random = "intercept",
                verbose = FALSE)
fits(fit)[[1]]@tau2
```

## Usage and caveats

```{r usage, eval = FALSE}
# random intercept + random niche slopes (fullest correction; use with an
# adequate number of samples). For few samples, random = "intercept" is safer.
res <- spiDE(spe, condition = "condition", sigma = c(10, 30, 50),
             random = "slope")
```

-   **Sample-level covariates.** A per-sample random intercept already adjusts for *all* between-sample nuisance variation, so it is confounded with measured sample-level covariates (Age, Sex, Stage). `spiDE` therefore errors if such a covariate is passed together with `random != "none"` — drop it; the random intercept subsumes it.
-   **Number of samples.** The correction is only as good as the between-sample information: with very few samples the variance components (and the $S-2$ reference df) are unstable, and a warning is issued.
-   **Correct, not merely "fixed with extras".** The mixed fit does more than add
    a between-patient variance component: it also replaces the default's NB
    dispersion scaling and sub-design covariance with the correct **working
    (quasi-NB) dispersion** and **full penalised covariance**. So even when
    $\tau^2 \to 0$ (no genuine clustering) the mixed fit does *not* reproduce the
    default fit — it gives the correctly-scaled, and hence more conservative,
    standard errors. This is deliberate: the default fit is known to be
    anti-conservative, so its larger apparent "power" is partly spurious.
-   **Power needs samples and effect size.** Because the test is now honest, a
    real response-niche effect must be visible *between patients* to be declared.
    With few patients, or an effect seen in only a handful of responders, the
    between-patient df ($S-2$) and the inflated SEs will (correctly) leave it
    non-significant. Larger sample sizes restore power.

# Session info

```{r sessioninfo}
sessionInfo()
```