---
title: "Neighbourhood-dependent differential expression with spiDE"
author: "Dharmesh D. Bhuva"
package: spiDE
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Neighbourhood-dependent differential expression with spiDE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

`spiDE` finds **neighbourhood-dependent differential expression** in spatial
transcriptomics. Within an *index* cell type, it asks whether a gene's
expression changes with the local density (the *niche*) of another cell type.
One hypothesis is one (gene, index type, niche type) *triplet*, and spiDE asks
two questions of it:

- **the pooled test**: within patients, does the gene's expression in the
  index type change with the density of the niche type, consistently across
  patients?
- **the condition-specific test**: does that dependence differ between two
  conditions (responders and non-responders, early and late stage)?

Both are questions about patients. The condition is a property of a patient,
not of a cell, so spiDE takes the patient as its unit of replication: a
triplet's evidence is the variation of its niche slope *between* patients,
never the number of cells.

Three steps:

1. **Niches**: `buildNiches()` evaluates, per sample, a Gaussian kernel
   density of every cell type at every cell.
2. **Fit**: `fitSpiDE()` fits, for every index cell type, one negative
   binomial GLM per gene of its expression on the niche densities, with an
   intercept per patient, so each niche coefficient is a *within-patient*
   slope.
3. **Test**: `testSpiDE()` tests those slopes across patients. It always runs
   the pooled test, and runs the condition-specific test when it is given a
   condition.

`spiDE()` runs all three. The model is stated in full in the *model*
vignette (`vignette("spiDE-model")`), and what its calibration rests on in
the *calibration* vignette (`vignette("spiDE-calibration")`).

# Data

spiDE needs a `SpatialExperiment` with

- **raw integer counts** (the `counts` assay), not normalised or
  back-transformed values;
- a **cell type** per cell (`cell_type`);
- a **patient** per cell (`sample_id`). One sample per patient: if a patient
  has several sections, give the patient in `sample_id` and the section in
  `section`, so the spatial variance is estimated per section;
- for the condition-specific test, a **two-level condition that is constant
  within a patient**.

The shipped example has 16 patients (eight responders, eight
non-responders), three cell types and 20 genes. It plants one effect: gene
`G1` is up-regulated in index type `A`, in responders only, in proportion to
the local density of niche type `B`.

```{r data, message = FALSE}
library(spiDE)
library(SpatialExperiment)
data(toySpiDE)
toySpiDE
patients <- unique(as.data.frame(colData(toySpiDE))[, c("sample_id", "condition")])
table(patients$condition)
table(toySpiDE$cell_type)
```

# Niches

`buildNiches()` computes the niche covariates at one or more bandwidths
`sigma` (in the units of the spatial coordinates) and stores each as a
`reducedDim` named `Niche<sigma>`:

```{r niches}
spe <- buildNiches(toySpiDE, sigma = 30, verbose = FALSE)
head(reducedDim(spe, "Niche30"), 3)
```

The bandwidth sets the scale of a neighbourhood. Fit one bandwidth at a time;
comparing results across bandwidths is a sensitivity analysis, not a
multiplicity-free search. `mergeNiches()` pools niche columns into groups
(e.g. all lymphocytes) before fitting, and spiDE never tests an index type
against its own niche, merged or not.

# Fit

```{r fit}
fit <- fitSpiDE(spe, condition = "condition", sigma = 30, verbose = FALSE)
fit
```

`fitSpiDE()` fits every index type with at least `min.cells` cells in at
least `min.patients` patients (`index` restricts it), and tests every other
cell type as a niche (`niche` restricts it; the others stay in the model as
adjustments). Genes detected in fewer than `min.detect` of an index type's
cells are not tested in it. Sequencing depth enters as the centred log
library size of each cell (`depth = "loglib"`); cell-level covariates are
added with `covariates`.

## The two engines

`engine` chooses how the between-patient error is estimated:

- **`"slopes"`** (the default) fits the model once, without the condition,
  then estimates every patient's own niche slopes (a one-step negative
  binomial estimate from the shared fit) with a within-patient spatial
  variance. `testSpiDE()` combines them across patients with weighted
  `limma`, allowing the slope to vary between patients. Because the fit
  never sees the condition, any patient-level condition can be tested
  afterwards, and a permutation of the condition reruns only the test.
- **`"sandwich"`** fits the condition-specific model directly and tests its
  coefficients with a patient-clustered CR2 sandwich and Bell-McCaffrey
  degrees of freedom. It needs the condition at fit time.

The slopes engine passes every calibration null of five comparisons in four
spatial cohorts, and the sandwich engine all but one borderline tail. The
slopes engine is the more powerful when effects vary between patients, the
sandwich engine competitive when every patient shares the effect; the
calibration vignette gives the numbers.

## Each patient's slopes

The slopes engine keeps every patient's slope, which is the evidence the
tests weigh:

```{r slopes, fig.width = 6, fig.height = 3.5}
ps <- patientSlopes(fit, gene = "G1", index = "A", niche = "B")
head(ps[, c("patient", "condition", "slope", "var_spatial", "ncells")])
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  ggplot(ps, aes(condition, slope)) +
    geom_hline(yintercept = 0, colour = "grey60") +
    geom_pointrange(aes(ymin = slope - 2 * sqrt(var_spatial),
                        ymax = slope + 2 * sqrt(var_spatial)),
                    position = position_jitter(width = 0.15, seed = 1)) +
    labs(x = NULL, y = "G1 slope on B density in A cells\n(log scale per unit log1p density)") +
    theme_minimal()
}
```

# Test

```{r test}
res <- testSpiDE(fit)
res
head(results(res, test = "pooled"))
```

With a condition (here recorded by `fitSpiDE()`; it can also be given to
`testSpiDE()`), the condition-specific test runs as well. The contrast is the
second level minus the first:

```{r condition}
res$contrast
cond <- results(res, test = "condition")
cond[cond$in_family, ]
```

By default (`procedure = "filtered"`) the condition-specific test is carried
out only among the triplets whose pooled test passes at `fdr`: the pooled
test never looks at the condition, so the filter is exact under any
relabelling of patients, and the multiplicity correction is spent on
triplets that have a niche dependence to modify. Rows outside that family
have `q = NA`. `procedure = "all"` tests every triplet.

The planted triplet, G1 in A against B, is the strongest pooled association
(`t` = `r round(results(res, test = "pooled")$t[1], 2)`), and in the
condition-specific test its slope is larger in responders by
`r round(cond$estimate[cond$gene == "G1" & cond$index == "A" & cond$niche == "B"], 2)` log units
per unit `log1p` density (`q` =
`r signif(cond$q[cond$gene == "G1" & cond$index == "A" & cond$niche == "B"], 2)`). `spiDE()` runs the three steps in one call:

```{r spiDE, eval = FALSE}
res <- spiDE(toySpiDE, condition = "condition", sigma = 30)
```

# A different question: niche abundance

spiDE's slopes are within-patient. A patient whose index cells sit among many
niche cells may also express a gene differently, as a whole, from a patient
whose index cells do not; that between-patient association is absorbed by
the patient intercepts and is invisible to `testSpiDE()`. It is real, and it
is a different claim: that patients with more of a niche type differ, not
that a cell responds to its neighbours. `testNicheAbundance()` asks it on
pseudobulk profiles, one per patient and index type, against the patient's
mean niche density:

```{r abundance}
ab <- testNicheAbundance(spe, condition = "condition", sigma = 30, verbose = FALSE)
head(ab)
```

Never report a result of one as a result of the other.

# Practicalities

- **Parallelism.** Genes are fitted independently given an index type, so
  `BPPARAM` parallelises over genes. Forked workers run one BLAS thread each
  when `RhpcBLASctl` is installed; set `OMP_NUM_THREADS=1` in the job
  otherwise.
- **Counts must be integers.** A back-transformed normalised assay is not a
  count and is refused.
- **Objects from spiDE 0.99.22 and earlier** came from a different model (a
  joint mixed-effects GLM). They cannot be read by this version; the engine
  is archived as the research package `spiDEmixed` (`research/mixed` in the
  spiDE repository), whose `readSpiDE()` reads them.

# Session information

```{r session}
sessionInfo()
```
