---
title: "Neighbourhood-dependent differential expression with spiDE"
author: "Dharmesh D. Bhuva"
package: spiDE
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Neighbourhood-dependent differential expression with spiDE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

`spiDE` finds **neighbourhood-dependent differential expression** in spatial
transcriptomics. Within an *index* cell type, it tests whether a gene's
response to an experimental *condition* changes with the local density (the
*niche*) of a surrounding cell type. One hypothesis is one (gene, index type,
niche type) *triplet*.

Four stages:

1. **Niches** — `buildNiches()`: per sample and per bandwidth, a Gaussian
   kernel density of every cell type evaluated at every cell.
2. **Fit** — `fitSpiDE()`: one negative binomial GLM per gene over a design
   with `cell type : condition : niche` interactions and patient-level random
   effects, all genes in one shared fit.
3. **Polish** — `polishSpiDE()`: converge each gene to its own optimum, set
   its dispersion there, and re-estimate the variance components from the
   converged fit. A stage the user runs, or skips, according to their data.
4. **Test** — `testSpiDE()`: per-gene Wald statistics, combined across a
   gene's correlated niches and across bandwidths, then a hierarchical FDR.
   `spiGSEA()` then tests gene sets on the same statistics.

`spiDE()` runs the first four (`polish = TRUE` by default). The model is stated in full in
`vignette("spiDE-model")`; every benchmark number quoted about it is in
`vignette("spiDE-calibration")`.

```{r load, message = FALSE}
library(spiDE)
library(SpatialExperiment)
```

# Example data

The package ships a small synthetic `SpatialExperiment`, `toySpiDE`: 20 genes,
480 cells, 6 samples (3 Responders, 3 Non-responders), three cell types. Gene
`G1` is planted up-regulated in index type **A**, in Responders, in proportion
to the local density of niche type **B**.

```{r data}
data(toySpiDE)
toySpiDE
table(toySpiDE$cell_type, toySpiDE$condition)
```

# One call

```{r full}
res <- spiDE(toySpiDE, condition = "condition", sigma = c(10, 30, 50),
             verbose = FALSE)
res
```

`results()` returns the tidy table, keyed by (gene, index type, niche type)
with the best bandwidth:

```{r results}
tab <- results(res)
head(tab[order(tab$fdr.niche), ], 10)
```

The planted effect is recovered, with a positive direction:

```{r recovered}
tab[tab$gene == "G1" & tab$ct_index == "A" & tab$ct_niche == "B", ]
```

## The result layers, and what each estimates

One fit answers three questions, each a different quantity:

```{r result-types}
vapply(c("niche", "celltype", "patient"),
       function(ty) nrow(results(res, type = ty)), integer(1))
```

- **`"niche"`** (default): does the neighbourhood modulate the response
  *within* an index cell type? The estimate is the gradient of the condition
  effect along the niche density, within (sample, cell type). Keyed by
  (gene, `ct_index`, `ct_niche`).
- **`"celltype"`**: does the response differ *by cell type*, regardless of
  neighbourhood? The flat, cell-type-wide shift, keyed by (gene, `ct_index`),
  with a two-level FDR.
- **`"patient"`**: does the gene respond at the patient level at all? One
  test per gene.

```{r result-celltype}
ct <- results(res, type = "celltype")
if (nrow(ct)) head(ct[order(ct$fdr.celltype), ], 5) else "no calls on this toy data"
```

The layers are not redundant: a flat response is not a neighbourhood effect
and is reported separately rather than allowed to load onto the niche slope.
A fourth question — does a *patient's* composition, the mean niche density
around their index cells, associate with expression in those cells — is
patient-level and is asked by `compositionTest()`, below.

# Step by step

The wrapper equals the three stages run explicitly, which is useful to inspect
intermediate objects, reuse a fit, or change one stage.

## Build niches

`buildNiches()` adds one `reducedDim` per bandwidth. `mergeNiches()` coarsens
cell types into niche groups; `computeSizeFactors()` derives a per-sample
library-size offset column.

```{r build}
spe <- buildNiches(toySpiDE, sigma = c(10, 30, 50))
reducedDimNames(spe)
```

## Fit

```{r fit}
fit <- fitSpiDE(spe, condition = "condition", verbose = FALSE)
fit
```

Each bandwidth's fit is a `SpiDEFit` with the coefficients `alpha`,
dispersions `psi`, the design `W`, a tag per design column, and — for the
default mixed fit — the variance components `tau2`:

```{r fitobj}
f1 <- fits(fit)[[1]]
table(f1$covtype)
f1$tau2
```

## Polish

The shared fit is one IRLS loop over all genes with one gene-averaged cell
weight vector, which is what makes a whole-transcriptome fit affordable; for
bright, cell-type-restricted genes it can stop short of that gene's own
optimum. `polishSpiDE()` converges each gene by damped Newton on its own
penalised likelihood, re-estimates its dispersion there (`psi = "profile"`),
and, for a mixed fit, re-estimates the variance components from the
converged coefficients (`tau2 = TRUE`) and refreshes the reference df. It
returns the object with inference cleared, so `testSpiDE()` runs on it:

```{r polish}
fit <- polishSpiDE(fit, spe, verbose = FALSE)
f1 <- fits(fit)[[1]]
head(f1$polish, 3)
f1$tau2
```

Skip the stage for a quick look at the shared fit, or when the counts are not
integers (the negative binomial likelihood is undefined off the integers, so
the stage refuses a back-transformed assay).

## Test

```{r test}
res2 <- testSpiDE(fit, spe = spe, fdr = 0.05)
nrow(results(res2))
```

# The patient-level question

`compositionTest()` asks the composition question on patients: per (sample,
index type) a pseudobulk profile of the index cells and the mean niche density
around them, then a `limma` moderated $t$ across samples. Two terms are
reported per (gene, index, niche): `"niche"`, the association pooled across
conditions, and `"condition:niche"`, its difference between conditions.

```{r composition}
ct <- compositionTest(spe, condition = "condition", sigma = 30, verbose = FALSE)
head(ct[order(ct$p), c("gene", "ct_index", "ct_niche", "term", "t", "fdr")])
```

On a fixture this small nothing is expected to pass. The point is that the
within-sample niche question and the between-patient composition question
have two tests, and neither is reported as the other.

# Niche-only analysis

Not every question has a condition. With `condition = NULL` the same
framework tests how expression within an index type changes with the density
of a niche type, in every sample alike:

```{r niche-only}
res_niche <- spiDE(toySpiDE, condition = NULL, sigma = c(10, 30),
                   verbose = FALSE)
head(results(res_niche))
```

The tested quantity is then the `CellType:niche` slope. The `"celltype"` and
`"patient"` layers are empty in this mode, and `spiGSEA(type = "celltype")`
errors, there being no condition to contrast. The mode is mildly
anti-conservative because the niche covariate is spatially autocorrelated;
see the model vignette.

# Scaling

The fit sees all genes at once (the dispersion is moderated across genes
before the per-gene convergence step). Inference and convergence are
independent across genes, processed in gene blocks and dispatched with
`r BiocStyle::Biocpkg("BiocParallel")`; the counts may be a
`r BiocStyle::Biocpkg("DelayedArray")`, realised one block at a time. On the
CPU backend, results are exactly invariant to the block size and the number
of workers.

```{r parallel, eval = FALSE}
library(BiocParallel)
res <- spiDE(toySpiDE, condition = "condition",
             block.size = 500, BPPARAM = MulticoreParam(4))
```

`backend = c("auto", "cpu", "gpu")` selects the compute backend for both
stages. On the GPU the per-gene Wald covariance and working weights are
batched across each gene block on the device, the block size is chosen to fit
the detected memory (override with `gpu.mem.budget`, in bytes), and a serial
`BPPARAM` is used since execution is already batched. The GPU path uses single
precision on Metal, so it matches the CPU path to a tolerance rather than
exactly.

# The defaults

What `spiDE()` and `fitSpiDE()` do unless told otherwise, one line each. The
reasons are in `vignette("spiDE-model")` and the measurements in
`vignette("spiDE-calibration")`.

| argument | default | what it does |
|---|---|---|
| `random` | `"intercept"` | a random intercept per sample, so response effects are tested against patient-level variation |
| `re.celltype` | `TRUE` | a random intercept per (sample, cell type), so the niche slope is a within-group slope and does not carry a patient's composition |
| `polish` | `TRUE` | run `polishSpiDE()` between the fit and the test: converge each gene, set its dispersion by profile likelihood, and re-estimate the variance components from the converged fit; needs integer counts |
| `dispersion` | `"ql"` | scale the standard errors by the gene's quasi-likelihood dispersion, moderated across genes; the only scale that holds the nominal level at four samples |
| `df.method` | `"satterthwaite"` | a reference df per tested coefficient |
| `combine` | `"cauchy"` | correlation-agnostic combination of a gene's niche p-values |

A sample-level covariate (Age, Stage) is collinear with the per-sample random
intercept and is rejected; adjust for cell-level covariates, including per-cell
library size, through `covariates`.

# Session info

```{r sessioninfo}
sessionInfo()
```
