---
title: "Neighbourhood-dependent differential expression with spiDE"
author: "Dharmesh D. Bhuva"
package: spiDE
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Neighbourhood-dependent differential expression with spiDE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

`spiDE` identifies **context-specific, neighbourhood-dependent differential expression** in spatial transcriptomics data. Within an *index* cell type, it tests how gene expression changes with an experimental *condition* as a function of the local density (the *niche*) of surrounding cell types.

The method proceeds in three steps:

1.  **Niche construction** (`buildNiches()`): for each sample and each spatial bandwidth, a Gaussian kernel density estimate of every cell type is evaluated at each cell, giving per-cell "effective niche" densities.
2.  **Model fitting** (`fitSpiDE()`): a per-gene negative binomial GLM is fit over a design containing the three-way `cell type : condition : niche` interactions, using the `r BiocStyle::Biocpkg("SpaNorm")` `fitNB()` engine.
3.  **Testing** (`testSpiDE()`): per-gene Wald statistics for the neighbourhood interactions are combined across correlated covariates (a Cauchy combination by default; Brown's method via `combine = "brown"`) and across bandwidths with a Cauchy combination, then subjected to a hierarchical (gene → index cell type → niche cell type) FDR correction.

`spiDE()` wraps the three steps.

```{r load}
library(spiDE)
library(SpatialExperiment)
```

# Example data

The package ships a small synthetic `SpatialExperiment`, `toySpiDE`, with 20 genes across 480 cells from 6 samples (3 Responders, 3 Non-responders) and three cell types (A, B, C). A neighbourhood signal is planted: gene `G1` is up-regulated in index cell type **A**, in Responders, in proportion to the local density of niche cell type **B**.

```{r data}
data(toySpiDE)
toySpiDE
table(toySpiDE$cell_type, toySpiDE$condition)
```

# The full workflow in one call

`spiDE()` defaults to `random = "intercept"`, a per-sample random intercept that
corrects the cell-level pseudo-replication described below. Note what that
implies for covariates: a per-sample random intercept already absorbs *all*
between-sample variation, so a sample-level covariate such as `Age` is
collinear with it and is rejected. Adjust for cell-level covariates only.

```{r full}
res <- spiDE(
  toySpiDE,
  condition = "condition",
  sigma = c(10, 30, 50),
  verbose = FALSE
)
res
```

The tidy results table is returned by `results()`, keyed by (gene, index cell type, niche cell type, bandwidth):

```{r results}
tab <- results(res)
head(tab[order(tab$fdr.niche), ], 10)
```

## Three kinds of response result

`results()` returns the niche-associated calls by default, but the design also
carries a `CellType:condition` term, which makes two further questions
answerable from the same fit --- no refitting, no second model.

```{r result-types}
vapply(c("niche", "celltype", "patient"),
       function(ty) nrow(results(res, type = ty)), integer(1))
```

- **`"niche"`** (default) --- does the *neighbourhood* modulate the response,
  within an index cell type? Keyed by `(gene, ct_index, ct_niche)`.
- **`"celltype"`** --- does the response differ *by cell type*, irrespective of
  neighbourhood? Keyed by `(gene, ct_index)`, with a two-step hierarchical FDR
  (gene, then cell type within gene).
- **`"patient"`** --- does the gene respond at the *patient* level at all? One
  test per gene, and the layer least exposed to the per-cell-type calibration
  differences discussed in `vignette("spiDE-calibration")`.

```{r result-celltype}
ct <- results(res, type = "celltype")
if (nrow(ct)) head(ct[order(ct$fdr.celltype), ], 5) else "no calls on this toy data"
```

On a dataset this small the two new layers are usually empty --- they are shown
here for the interface. On a real cohort they are not redundant with the niche
layer: they routinely recover genes the niche layer never reports, because a
cell-type-wide response shift is not a neighbourhood effect and has nowhere to
go in the three-way term.

The planted `G1` / index A / niche B effect is recovered with a positive (up-regulated) direction:

```{r recovered}
tab[tab$gene == "G1" & tab$ct_index == "A" & tab$ct_niche == "B", ]
```

# Step by step

The wrapper is equivalent to running the three stages explicitly, which is useful when you want to inspect intermediate objects, reuse a fit, or change parameters at a single stage.

## 1. Build niches

`buildNiches()` adds one `reducedDim` per bandwidth. Optionally, fine cell types can be merged into coarser niches with `mergeNiches()`, and per-sample library-size offsets computed with `computeSizeFactors()`.

```{r build}
spe <- buildNiches(toySpiDE, sigma = c(10, 30, 50))
reducedDimNames(spe)
```

## 2. Fit the model

```{r fit}
fit <- fitSpiDE(spe, condition = "condition", verbose = FALSE)
fit
```

Each bandwidth's fit is a `SpiDEFit` object exposing the coefficient matrix `alpha`, dispersions `psi`, the design `W`, and the covariate-type tags:

```{r fitobj}
f1 <- fits(fit)[[1]]
table(f1$covtype)
```

## 3. Test

```{r test}
res2 <- testSpiDE(fit, spe = spe, fdr = 0.05)
nrow(results(res2))
```

# Niche-only analysis: no condition

Not every question needs an experimental condition. The same framework tests how expression within an *index* cell type changes with the local density of a *niche* cell type, with no condition involved at all — the two-way `CellType:niche` interaction. Pass `condition = NULL` to get it:

```{r niche-only}
res_niche <- spiDE(toySpiDE, condition = NULL, sigma = c(10, 30),
                   verbose = FALSE)
res_niche
head(results(res_niche))
```

The design drops the condition terms *and* the bare niche main effects,

```
~ 0 + covariates + CellType + CellType:(niches)
```

leaving the `CellType:niche` columns — nuisance terms in the conditioned model — as the tested effects, cell-means coded. Dropping the niche main effects matters: with them in, `niche_n` equals the sum of `CellType_c:niche_n` over cell types, so one interaction per niche would be aliased away and the affected index cell type could never be tested. Without them, every index cell type keeps a slope against every non-self niche and each coefficient reads directly as a within-cell-type slope of expression on that niche's log density.

Everything downstream is shared: the same Wald statistics, the same within-gene and cross-bandwidth Cauchy combination, and the same three-level gene → index cell type → niche cell type FDR cascade. The `results(type = "celltype")` and `results(type = "patient")` tables are empty in this mode, there being no condition to contrast, and `spiGSEA(type = "celltype")` errors for the same reason.

The mixed-effects correction applies here too. Note one caveat specific to this mode: the tested slope is a *within-sample* contrast on a *spatially autocorrelated* covariate, and spiDE does not model spatial autocorrelation, so neighbouring cells are not independent replicates of that slope. Random effects remove most of the resulting inflation — on a null fixture, 37 calls with `random = "none"` against 5 with `random = "intercept"` — but niche mode remains mildly anti-conservative. Treat borderline calls with corresponding caution.

# Scaling

Fitting is performed over all genes at once (the negative binomial dispersion is moderated across genes). The per-gene inference is independent across genes and is processed in gene-blocks, dispatched with `r BiocStyle::Biocpkg("BiocParallel")`; the counts matrix may be a `r BiocStyle::Biocpkg("DelayedArray")` and is realised one block at a time. On the CPU backend, results are exactly invariant to the block size and the parallel backend; see below for the GPU backend's tolerance-level invariance.

```{r parallel, eval = FALSE}
library(BiocParallel)
res <- spiDE(
  toySpiDE,
  condition = "condition",
  block.size = 500,
  BPPARAM = MulticoreParam(4)
)
```

The `backend` argument (`"auto"`, `"cpu"`, `"gpu"`) selects the compute backend for **both** stages. For the model fit it is forwarded to `SpaNorm::fitNB()`. For inference, the GPU backend batches the per-gene Wald covariance and negative-binomial working weights across each gene-block on the accelerator (via `r BiocStyle::Biocpkg("SpaNorm")`'s tensor engine), rather than computing them one gene at a time. On the GPU backend the block size is auto-selected to keep peak device memory within the detected budget (override with `gpu.mem.budget`, in bytes), and a serial `BPPARAM` is used automatically — GPU execution is already batched across genes, so multiple CPU workers would only contend for the one device. The GPU path uses single-precision on Metal/MPS, so its results match the CPU path to a small tolerance rather than exactly (the CPU path remains exactly reproducible and block-size invariant).

# Mixed-effects inference for pseudo-replication

By default each cell is treated as an independent observation. When many cells come from few patients, the response-related tests become anti-conservative (cell-level pseudo-replication), because the clinical response is a patient-level label. Passing `random = "slope"` (or `"intercept"`) to `spiDE()` / `fitSpiDE()` adds patient-level random effects — implemented as ridge-penalised design columns — that test the response effects against between-patient variability:

```{r random, eval = FALSE}
res <- spiDE(toySpiDE, condition = "condition", sigma = c(10, 30, 50),
             random = "slope")
```

The model, the ridge = random-effects equivalence, and a worked calibration example are described in the companion vignette *The spiDE model, mixed-effects inference, and two-stage estimation*. Note that a per-sample random intercept absorbs all between-sample effects, so sample-level covariates (e.g. `Age`) must be dropped when random effects are used.

An alternative, structurally different route to the same question is `twoStageSpiDE()`, which estimates each patient's niche slope first and then contrasts the patient slopes with a moderated *t*-test — removing the pseudo-replication by construction rather than by correction. It trades power for the best-calibrated error control measured in the package's benchmark suite and is intended for pre-specified (restricted) hypotheses; the model vignette documents it in full, and the *Two-stage estimation* benchmark report on the spiDE-research site (https://bhuvalab.github.io/spiDE-research/) reports its operating characteristics.

# Session info

```{r sessioninfo}
sessionInfo()
```
