---
title: "Speeding up the mixed-effects fit: the PQL subsampling trade-off"
author: "Dharmesh D. Bhuva"
package: spiDE
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Speeding up the mixed-effects fit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE, collapse = TRUE, comment = "#>", message = FALSE,
  warning = FALSE, fig.width = 7, fig.height = 4, dpi = 150, fig.retina = 2
)
```

```{r libs}
library(spiDE)
library(ggplot2)
source(system.file("spide_palette.R", package = "spiDE"))

bench <- function(f) {
  path <- system.file("extdata", "benchmark", f, package = "spiDE")
  if (file.exists(path)) utils::read.csv(path, check.names = FALSE) else NULL
}
```

# Overview

The mixed-effects fit (`fitSpiDE(..., random = "intercept" | "slope")`) corrects
the anti-conservative inference that cell-level pseudo-replication causes (see
*The spiDE model* vignette). It does so with a Schall/PQL loop that re-estimates
the shared random-effect variance components `tau2`, re-fitting the
negative-binomial GLM once per iteration — the expensive part of the fit.

Those variance components are a *single shared scalar per random-effect group* — a
population quantity that does not need every cell to estimate, exactly as
`SpaNorm::fitNB` already subsamples cells for its dispersion step. `spiDE` exposes
three knobs that trade a little `tau2` accuracy for a large speed-up:

| argument       | default | effect |
|----------------|---------|--------|
| `re.prop`      | `0.2`   | fraction of cells sampled per cell type within each sample for the PQL loop |
| `re.maxit.psi` | `1`     | dispersion iterations inside each PQL step |
| `re.min.cells` | `100`   | per-stratum floor for `re.prop` |

Crucially the loop only supplies `tau2`; a single **final fit on all cells with
full dispersion** produces the coefficients and dispersion that inference uses. So
subsampling perturbs only the shared variance estimate, not the per-gene effect
sizes. This vignette quantifies the time saved and the accuracy retained, using
the benchmark study in the package's `research/` directory.

# Time versus accuracy across `re.prop`

The study fits the same realistic dataset at several values of `re.prop`, timing
each fit and correlating its response-niche t-statistics against the fully
accurate reference (`re.prop = 1`, all cells). Two figures tell the story:
subsampling makes the fit markedly faster while the t-statistics it feeds into
inference stay highly correlated with the full-data fit.

```{r pql, fig.height = 4, fig.cap = "Left: the mixed-effects fit is markedly faster once cells are subsampled for the PQL loop (about 1.8x versus the all-cell fit), after which it plateaus. Right: response-niche t-statistics stay highly correlated with the full-data reference (~0.96) across the sampling range."}
timing <- bench("pql_timing.csv")
acc <- bench("pql_accuracy.csv")

if (!is.null(timing)) {
  p_time <- ggplot(timing, aes(re.prop, elapsed)) +
    geom_line(colour = spide_pal$ref) +
    geom_point(size = 3, colour = spide_pal$accent) +
    scale_x_continuous(trans = "reverse") +
    labs(x = "re.prop (cell fraction, reversed)", y = "fit time (s)",
         title = "Fit time versus subsampling")

  if (!is.null(acc)) {
    # anchor the reference (re.prop = 1) at correlation 1
    acc <- rbind(acc, data.frame(re.prop = 1, cor = 1))
    p_acc <- ggplot(acc, aes(re.prop, cor)) +
      geom_line(colour = spide_pal$ref) +
      geom_point(size = 3, colour = spide_pal$accent) +
      scale_x_continuous(trans = "reverse") +
      coord_cartesian(ylim = c(min(0.9, min(acc$cor)), 1.001)) +
      labs(x = "re.prop (cell fraction, reversed)",
           y = "t-stat correlation vs full-data fit",
           title = "Accuracy versus subsampling")
    print(patchwork::wrap_plots(p_time, p_acc))
  } else {
    print(p_time)
  }
}
```

Accuracy is largely retained because the coefficients come from the full final
fit; `re.prop` only moves the shared `tau2`. The subsample does modestly
underestimate `tau2` (here about 0.19 versus 0.25 at full data), and this feeds
through to a small drop in the t-statistic correlation — to roughly 0.96 rather
than 1. The default `re.prop = 0.2` therefore delivers the bulk of the speed-up
at a small, well-characterised cost to the inference; use `re.prop = 1` when
exact reproducibility matters more than speed.

```{r tau2-table, echo = FALSE}
if (!is.null(timing)) {
  knitr::kable(timing[order(-timing$re.prop),
                      c("re.prop", "elapsed", "tau2")],
               digits = 3, row.names = FALSE,
               col.names = c("re.prop", "fit time (s)",
                             "tau2 (intercept)"),
               caption = "Fit time and the estimated intercept variance component across re.prop.")
}
```

# Scaling to realistic gene counts

A realistic dataset has ten-thousand-plus genes. Because the whole gene set is
fit in a single call (dispersion is moderated across all genes together), the
per-fit cost grows with both the number of genes and the number of cells. The
study records the fit and inference time as the number of cells per sample grows
at fixed gene count.

```{r timing, fig.height = 4, fig.cap = "Fit and inference wall-clock time as the number of cells grows, at a realistic gene count. Inference is the blockable, parallelisable stage."}
tm <- bench("timing.csv")
if (!is.null(tm) && nrow(tm) > 0) {
  long <- rbind(
    data.frame(n_cells = tm$n_cells, stage = "fit", seconds = tm$fit_seconds),
    data.frame(n_cells = tm$n_cells, stage = "inference",
               seconds = tm$test_seconds))
  ggplot(long, aes(n_cells, seconds, colour = stage)) +
    geom_line(linewidth = 0.8) + geom_point(size = 2) +
    scale_colour_manual(values = unname(spide_pal$method[c(1, 2)])) +
    labs(x = "total cells", y = "wall-clock time (s)", colour = NULL,
         title = sprintf("Scaling at ~%s genes",
                         format(round(mean(tm$n_genes), -2), big.mark = ",")))
} else {
  cat("*(Timing summary not found; run the benchmark study in `research/`.)*")
}
```

# Summary

* Subsampling cells for the PQL loop (`re.prop`) plus a single inner dispersion
  iteration (`re.maxit.psi = 1`) cuts the mixed-effects fit time by roughly 1.8x,
  while the final all-cell fit keeps the coefficients exact.
* Response-niche t-statistics stay highly correlated (~0.96) with the full-data
  fit across `re.prop`, so the default `re.prop = 0.2` is a fast choice at a
  small, well-characterised accuracy cost.
* Set `re.prop = 1` for a fully reproducible (seed-independent) fit, at the cost
  of speed; otherwise set an external `set.seed()` to make a subsampled fit
  reproducible.

Timing numbers reflect the machine that produced the benchmark tables; regenerate
them for your own hardware with the study in `research/`.

```{r session}
sessionInfo()
```
