---
title: "The spiDE model"
author: "Dharmesh D. Bhuva"
package: spiDE
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{The spiDE model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# echo = TRUE: this vignette is a code-along walkthrough -- its prose narrates
# the chunks. Pure display tables opt out per chunk with echo = FALSE.
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
set.seed(1)
```

# Overview

This vignette states the model behind `spiDE` once, from the top down: the
quantity it estimates, the design that isolates that quantity, why cells are
not replicates and what the random effects do about it, how each gene's fit is
converged, and how a Wald statistic becomes a row of `results()`. It is the
reference for the other vignettes. The quickstart (`vignette("spiDE")`) shows
the calls; the calibration vignette (`vignette("spiDE-calibration")`) holds
every benchmark number quoted in this package; the two-stage estimator, a
different route to the same question, has its own vignette
(`vignette("spiDE-twostage")`).

One idea organises the whole design and is used throughout: the
**Frisch–Waugh–Lovell theorem**. Every nuisance block in the design removes one
component from the tested slope, and the model is best read as a sequence of
such removals. It is stated in full in the section *What each block does to
the tested slope*.

```{r load, message = FALSE}
library(spiDE)
library(SpatialExperiment)
library(ggplot2)
source(system.file("spide_palette.R", package = "spiDE"))
```

# The question, and the quantity estimated

Within an **index** cell type, does a gene's response to an experimental
**condition** change with the local density of a **niche** cell type? That is
one hypothesis per (gene, index type, niche type) *triplet*, and the quantity
behind it is a slope: how much the condition effect on log expression changes
per unit of niche density, among cells of the index type.

Three quantities have to be kept apart, and the design gives each its own
column so that the tested slope carries only the third:

1. **A flat cell-type response.** The gene is up in every index cell of every
   Responder, regardless of neighbourhood. This is a between-patient
   contrast on cells of one type; `results(type = "celltype")` reports it.
2. **A condition-independent niche slope.** Expression in the index type
   tracks the niche's density in every patient, Responder or not. This is a
   nuisance in the conditioned analysis and the tested quantity in
   niche-only mode (`condition = NULL`).
3. **The niche-dependent response.** The *difference* in that slope between
   conditions — the three-way `cell type : condition : niche` term.

Two more things are deliberately *not* part of the third quantity, and the
model's random effects exist to keep them out. Patients differ in baseline
expression of every gene, and patients differ in composition: a patient whose
index cells sit, on average, in a denser niche of some type may also express
the gene differently in that type for reasons that have nothing to do with
the neighbourhood of any single cell. Both are patient-level associations with
one observation per patient. The second is a legitimate question of its own
and `compositionTest()` asks it at the patient level; it is not
neighbourhood-dependent expression and the niche slope must not carry it.

# The effective niche

Every quantity in this section is computed **within one sample at a time**,
entirely independently of every other sample. Let $\mathcal C_s$ be the cells
of sample $s$. For each cell $c \in \mathcal C_s$ the local intensity of cell
type $t'$ is a kernel-smoothed count of that type's cells around $c$, with a
Gaussian kernel $K_\sigma$ of bandwidth $\sigma$:

$$
\Lambda_{c,t',s}(\sigma) \;=\; e_s(c) \sum_{c' \in \mathcal C_s}
  K_\sigma\!\left(d_{c,c'}\right)\,\mathbb{1}(t_{c'} = t'),
$$

where $d_{c,c'}$ is the planar distance between two cells of the same sample
and $e_s(c)$ is Diggle's edge correction for that sample's own spatial window.
`buildNiches()` never forms a distance matrix across samples: it splits the
`SpatialExperiment` per sample and builds a fresh `spatstat` point pattern,
window and density surface for each subset, so two cells from different
samples that happen to share coordinates never interact. The expected
log-count of type $t'$ in a disc of radius $\sigma$ is then

$$
N_{c,t',s}(\sigma) \;=\; \log\!\left(\pi\sigma^2\,\Lambda_{c,t',s}(\sigma) + 1\right),
$$

and stacking over types gives the **effective niche** of the cell,
$N_{c,s}(\sigma)$, a vector with one entry per cell type. Per-sample results
are row-bound into one cells × types matrix per bandwidth, stored as
`reducedDim(spe, "Niche<sigma>")`. Several bandwidths can be built at once;
inference combines them later.

# The model

## Notation

$Y_{c,g}$ is the count of gene $g$ in cell $c$; $t_c$ its cell type and $s_c$
its sample. The condition $r_s \in \{0, 1\}$ is a property of the sample —
**constant across all cells of a sample**, which drives everything below. $k$
indexes the (sample, cell type) group a cell belongs to. $S$ is the number of
samples, $C$ the number of cells, $G$ the number of genes.

## The negative binomial GLM, stated once

For one bandwidth $\sigma$, gene $g$ has a negative binomial distribution with
dispersion $\psi_g$, $\mathrm{Var}(Y_{c,g}) = \mu_{c,g} + \psi_g\mu_{c,g}^2$,
and log-mean

$$
\begin{aligned}
\log \mu_{c,g} \;=\;&
 \underbrace{\gamma_g^\top x_c \;+\; \zeta_{g,t_c}
 \;+\; \sum_{t'} N_{c,t'}\,\alpha_{g,t_c,t'}}_{\text{within-sample, fixed}}
 \;+\; \underbrace{\eta_{g,t_c}\, r_{s_c}
 \;+\; r_{s_c}\sum_{t'} N_{c,t'}\,\beta_{g,t_c,t'}}_{\text{response effects, fixed, tested}} \\[4pt]
 &+\; \underbrace{u_{g,s_c} \;+\; v_{g,s_c t_c}
 \;+\; \sum_{t'} N_{c,t'}\, w_{g,s_c,t_c,t'}}_{\text{patient-level random effects}} ,
\end{aligned}
$$

with

$$
u_{g,s} \sim \mathcal N(0, \tau^2_{\text{sample}}),\qquad
v_{g,sk} \sim \mathcal N(0, \tau^2_{\text{sample}\times\text{type}}),\qquad
w_{g,s,t,t'} \sim \mathcal N(0, \tau^2_{\text{slope}}).
$$

| term | meaning | design tag |
|---|---|---|
| $\gamma_g^\top x_c$ | cell-level covariates (`covariates =`), e.g. log library size | `Other` |
| $\zeta_{g,t}$ | baseline expression of the gene in cell type $t$ | `CellType` |
| $\alpha_{g,t,t'}$ | condition-independent slope of expression in type $t$ on the density of type $t'$ | `Niche` |
| $\eta_{g,t}$ | the flat response of type $t$: quantity 1 above | `ResponseCellType` |
| $\beta_{g,t,t'}$ | **the niche-dependent response**: quantity 3, the tested slope | `ResponseNiche` |
| $u_{g,s}$ | random intercept per sample | `Random` (`SampleInt`) |
| $v_{g,sk}$ | random intercept per (sample, cell type) — `re.celltype = TRUE`, the default | `Random` (`SampleCellTypeInt`) |
| $w_{g,s,t,t'}$ | random niche slopes per sample — `random = "slope"` only | `Random` (`SampleSlope`) |

The variance components $\tau^2$ carry no gene index: one value of each is
shared by every gene, estimated from all of them. The realised deviations
$u_{g,s}$, $v_{g,sk}$ vary by gene. `random = "intercept"` (the default) adds
$u$ and, with `re.celltype = TRUE`, $v$; `random = "slope"` adds $w$ as well;
`random = "none"` adds nothing and should not be used for inference (see *Why
cells are not replicates*).

We test, per triplet,

$$
H_0:\ \beta_{g,t,t'} = 0 \qquad\text{against}\qquad H_1:\ \beta_{g,t,t'} \ne 0 .
$$

## Cell-means coding

`nicheDesign()` assembles the fixed part as

```
~ 0 + <covariates> + CellType + CellType:condition
    + CellType:(niches) + CellType:condition:(niches) + niches
```

There is no bare `condition` main effect: the response contrast is carried by
one `CellType:condition` column per cell type, each readable directly as that
type's flat response, and the three-way columns are one per (index type, niche
type) with the index type's own density dropped (a type's density of itself is
not a niche). The random-effect columns are ordinary design columns — an
indicator per sample, an indicator per (sample, cell type), and, for slopes,
sample indicators times the `CellType:niche` bases — tagged `Random` and
identified by the ridge penalty described under *Random effects as ridge*.
Every column carries a tag (`fits(res)[[1]]$covtype`), and the tested tag is
`ResponseNiche` (or `Niche` in niche-only mode).

## What each block does to the tested slope: Frisch–Waugh–Lovell

Consider a weighted least-squares fit of $y$ on two blocks of columns,
$[X_1 \mid X_2]$, with weights $D$. The **Frisch–Waugh–Lovell theorem** says
the coefficient on $X_2$ equals the coefficient from a regression of
$M_1 y$ on $M_1 X_2$, where

$$
M_1 \;=\; I - X_1\big(X_1^\top D X_1\big)^{-1} X_1^\top D
$$

removes from any vector its weighted projection on $X_1$. In words: **the
coefficient on a column is estimated only from the part of that column, and
of the response, that the other columns cannot explain.** Every nuisance block
in the design above is a choice of $X_1$, and the theorem says exactly what it
removes from $\beta$.

**The (sample × cell type) indicators.** When $X_1$ is a set of group
indicators, $M_1$ subtracts each group's weighted mean:
$(M_1 x)_c = x_c - \bar x_{k(c)}$. With the nested block in the design, the
tested slope is therefore estimated from the *within-(sample, cell type)*
deviations of niche density and of expression, and the between-group part —
patients whose index cells sit in denser niches also expressing the gene
differently in that type — is removed exactly and assigned to the intercepts,
where it belongs. Without the block, only a per-sample intercept is present,
shared across cell types; nothing removes the between-sample covariance
*within* an index type, so it loads onto $\beta$ and is reported with a
standard error computed from $C$ cells rather than $S$ patients. A
permutation of the niche rows within (sample, cell type) cannot detect this,
because it preserves every group's mean niche density and so preserves the
confound. That is what the nested block was added for, and it is why
`random = "slope"` is not a substitute: per-sample slopes on the niche bases
leave the group means untouched.

**The `CellType:condition` column.** Now let $X_1$ be the flat response of
the index type. $M_1$ removes the condition's mean effect in that type from
both the response and the niche column, so $\beta$ is the *gradient* of the
response along the niche and not its level. Without the column the flat
response has nowhere to go but the uncentred slope, and the slope's standard
error is that of a regression through the origin — smaller, because it is
bought by assuming there is no level to estimate. A design without this
column reports more niche calls; it is estimating a broader quantity under the
niche label, not testing the same one better. The measurement is in the
calibration vignette.

**The IRLS fit.** The theorem is about least squares, and the negative
binomial fit is not least squares — but each step of the iteratively
reweighted least-squares (and, later, Newton) iteration *is* a weighted
least-squares problem in the working response, so the theorem holds step by
step and hence at the converged fit, with $D$ the working weights
$w_{c,g} = \mu_{c,g}/(1+\psi_g\mu_{c,g})$.

**One caveat, because it is honest.** The nested block is not free: it is
ridge-penalised (next section), and a penalised block does not centre
completely. With penalty $\lambda$ on the group indicators, group $k$'s mean
is removed by the factor $n_k \bar w_k / (n_k \bar w_k + \lambda)$, where
$n_k \bar w_k$ is the group's total working weight — exact centring only as
$\lambda = 1/\tau^2 \to 0$ (large estimated between-group variance) or as
groups become large. On the real cohort that this was built for, the shrunken
centring reproduced the exact-centring null within measurement; the numbers
are in the calibration vignette.

# Why cells are not replicates

The design has one row per cell. That is fine for the within-sample
coefficients ($\zeta$, $\alpha$), but $\eta$ and $\beta$ are contrasts on the
patient-level label $r_s$: there are $S$ independent patients, yet a Wald
information matrix that sums over $C$ cells makes their standard errors
shrink like $1/\sqrt C$ instead of $1/\sqrt S$. With $C \gg S$ the statistics
are badly inflated. This is what `random = "none"` does, and the reason it is
no longer the default.

The demonstration below simulates a **null**: every cell of a patient shares
a per-patient shift, and there is no response effect at all, so every response
test should be null.

```{r sim}
sim_clustered <- function(n_samples = 10, n_per = 90, n_genes = 300,
                          field = 500, sd_patient = 0.7, seed = 1) {
  set.seed(seed)
  cts <- c("A", "B", "C"); gn <- sprintf("G%d", seq_len(n_genes))
  sids <- sprintf("S%d", seq_len(n_samples))
  cond <- rep(c("Responder", "Non-responder"), length.out = n_samples)
  names(cond) <- sids
  cells <- lapply(sids, function(sid) {
    x <- runif(n_per, 0, field); y <- runif(n_per, 0, field)
    ct <- ifelse(x > 0.6 * field & runif(n_per) < 0.7, "B",
                 sample(c("A", "C"), n_per, replace = TRUE))
    data.frame(sample_id = sid, condition = cond[[sid]],
               x = x, y = y, cell_type = ct)
  })
  cd <- do.call(rbind, cells); n <- nrow(cd)
  cd$cell_id <- sprintf("c%d", seq_len(n))
  base <- matrix(rnorm(n_genes * 3, 1.5, 0.6), n_genes, dimnames = list(gn, cts))
  lmu <- base[, cd$cell_type, drop = FALSE]
  # per-(gene, patient) random intercept: real clustering, NO response effect
  u <- matrix(rnorm(n_genes * n_samples, 0, sd_patient), n_genes,
              dimnames = list(gn, sids))
  lmu <- lmu + u[, cd$sample_id, drop = FALSE]
  counts <- matrix(rnbinom(length(lmu), mu = as.vector(exp(lmu)), size = 5),
                   n_genes, dimnames = list(gn, cd$cell_id))
  list(spe = SpatialExperiment(assays = list(counts = counts),
                               colData = S4Vectors::DataFrame(cd),
                               spatialCoords = as.matrix(cd[, c("x", "y")])),
       u = u)
}

sim_null <- sim_clustered()
spe <- buildNiches(sim_null$spe, sigma = 30)
```

300 genes rather than a handful, because $\tau^2$ is a single value shared
across every gene: estimating a variance from a few genes is noisy in the same
way as estimating it from a few observations.

A helper that fits one random mode and returns the response p-values and the
fitted object:

```{r respp}
fit_mode <- function(random) {
  # df.method = "between" (the scalar S - 2) is pinned here: this section is
  # about the correction itself, and the between reference keeps the demo
  # cheap. The default "satterthwaite" per-coefficient df has its own section.
  f <- fitSpiDE(spe, "condition", sigma = 30, random = random,
                df.method = "between", verbose = FALSE)
  res <- testSpiDE(f, spe = spe)          # populates the Wald inference slots
  ff <- fits(res)[[1]]
  # the response contrast is carried by the CellType:condition columns; match
  # on column NAMES -- `covtype` spans every design column, `t_stat` only the
  # tested ones
  ct <- as.character(ff$covtype)
  rc <- intersect(colnames(ff$t_stat),
                  ff$coefmap$covariate[ct == "ResponseCellType"])
  tt <- ff$t_stat[, rc, drop = FALSE]
  df <- if (is.null(ff$df)) Inf else if (length(ff$df) == 1L) ff$df else ff$df[rc]
  dfm <- matrix(rep(df, each = nrow(tt)), nrow = nrow(tt))
  list(fit = ff, p = 2 * pt(-abs(tt), dfm))
}
fixed <- fit_mode("none")
mean(fixed$p < 0.05)  # should be ~0.05 under the null
```

The fixed-effects fit rejects a large fraction of null tests at
$\alpha = 0.05$.

## Why not add sample as a fixed effect?

The instinct from fixed-effects modelling is to add `sample_id` as a factor,
the way one adds `batch`. It cannot work here, and the reason motivates the
random effect:

```{r whynotfixed}
set.seed(42)
S <- 12; n_per <- 12
toy_sample <- factor(rep(seq_len(S), each = n_per))
toy_resp   <- rep(rbinom(S, 1, 0.5), each = n_per)   # constant within a sample
true_tau2  <- 0.49
u_true     <- rnorm(S, 0, sqrt(true_tau2))
eta_true   <- 0.3 * toy_resp + u_true[as.integer(toy_sample)]
toy <- data.frame(y = rpois(length(eta_true), exp(eta_true)),
                  resp = toy_resp, sample_id = toy_sample)
coef(glm(y ~ sample_id + resp, family = poisson, data = toy))["resp"]
```

`resp` is `NA` because every cell of a sample shares the response label:
`resp` is an exact linear combination of the sample dummies, and a design with
a free column per sample has nothing left to estimate the contrast from.
`spiDE` enforces the mirror image — `checkSample()` errors if a sample-constant
covariate is passed with `random != "none"`, since the random intercept makes
it exactly as inestimable. A free per-sample parameter would also fit each
sample perfectly and say nothing about how much patients differ *in general*,
which is precisely the quantity the corrected test needs. A random effect
sits between the two extremes: each patient gets its own deviation $u_s$, but
the deviations are draws from a shared $\mathcal N(0, \tau^2)$, with $\tau^2$
estimated from the data and each $\hat u_s$ shrunk toward zero by an amount
that reflects how reliable that patient's own data is.

# Random effects as ridge

## The identity

One IRLS step for a GLM is a weighted least-squares regression of the working
response $z_{c,g} = \eta_{c,g} + (Y_{c,g} - \mu_{c,g})/\mu_{c,g}$ on the
design, with weights $D_g = \mathrm{diag}(w_{c,g})$. Write the expanded design
as $\tilde W = [\,W \mid Z\,]$ with $Z$ the random-effect columns. On this
working scale, **a Gaussian random effect on a block of columns is exactly an
$L_2$ (ridge) penalty on those columns** — an identity, not an approximation.
The penalised objective

$$
\lVert z - \tilde W\theta \rVert^2_{D_g} + \theta^\top \Lambda\,\theta,
\qquad
\Lambda = \mathrm{blockdiag}\big(0_{\text{fixed}},\ \lambda_{\text{sample}} I,\
\lambda_{\text{sample}\times\text{type}} I,\ \lambda_{\text{slope}} I\big),
$$

has the closed-form solution
$\hat\theta = (\tilde W^\top D_g \tilde W + \Lambda)^{-1}\tilde W^\top D_g z$
with

$$
\boxed{\ \lambda_k = 1/\tau^2_k\ } .
$$

Read it either way: a ridge penalty is the log-density of a Gaussian prior
$b \sim \mathcal N(0, 1/\lambda)$, so the penalised fit is the posterior mode
under the random-effect distribution; or, the penalty pulls every $\hat u_s$
toward zero, which is what borrowing strength across patients means
numerically. The penalised estimates of the random columns are the model's
**BLUPs**. This is the penalised quasi-likelihood construction (Breslow &
Clayton, 1993; Schall, 1991) that `lme4`, `mgcv` and `MASS::glmmPQL()` build
on, and it needs no change to the fitting engine: `SpaNorm::fitNB` accepts a
per-column penalty `lambda.a`, set to $0$ on fixed columns and $1/\tau^2_k$ on
random ones.

## Estimating the variance components

The one thing the identity does not supply is $\tau^2$. `.fitNBmixed()` gets
it with two nested loops. The **inner loop** is `fitNB` at a fixed penalty:
the penalised estimate of every gene at once, BLUPs included. The **outer
loop** re-estimates each component by a method-of-moments update and refits:

$$
\hat\tau^2_k = \frac{\sum_g \lVert \hat b_{g,k}\rVert^2}
                    {G\,\big(q_k - \lambda_k\,\mathrm{tr}\,C_{kk}\big)},
\qquad
C = \big(\tilde W^\top \bar D\,\tilde W + \Lambda\big)^{-1},
$$

where $\hat b_{g,k}$ are gene $g$'s BLUPs in block $k$, $q_k$ the block's
number of columns, $\bar D$ a gene-averaged working-weight matrix (one
representative information matrix, since $\tau^2$ is shared), $C_{kk}$ the
block's penalised covariance, and $q_k - \lambda_k\,\mathrm{tr}\,C_{kk}$ its
**effective degrees of freedom** — less than $q_k$ because shrunken BLUPs use
up less than a parameter each. That denominator is what makes $\hat\tau^2$ an
estimate of the *true* between-patient variance rather than the variance of
the shrunken BLUPs. The loop stops when $\log\hat\tau^2$ moves by less than
`re.tol`, or after `re.maxit` iterations (default 2: for the intercept model
one iteration is indistinguishable from ten, and a small cap avoids a
two-cycle the estimate can fall into; pass `re.maxit = 10` for
`random = "slope"`, whose slope component converges slowly). `re.prop < 1`
runs the inner fits on a stratified cell subsample; the coefficients inference
uses always come from a final fit on all cells, but the shared $\tau^2$ is
biased downward by subsampling, so the default is `re.prop = 1`.

## Watching it work

The same construction, on one gene of the toy data, with an independent
implementation — `MASS::glmmPQL()`:

```{r pql_demo, message = FALSE}
library(MASS)
fit_pql <- glmmPQL(y ~ resp, random = ~ 1 | sample_id, family = poisson,
                   data = toy, verbose = FALSE)
summary(fit_pql)$tTable
```

`resp` is now estimable, and its reference `DF` is $S - 2 = 10$, the
between-patient degrees of freedom. `glmmPQL()` exposes its outer-iteration
cap as `niter`, so refitting with `niter = 1, 2, …` replays the trajectory
`.fitNBmixed()`'s loop takes:

```{r tau2_trace, fig.small = TRUE, fig.cap = "The outer loop converging on the patient-level variance component for one gene. The dashed line is the simulated value."}
tau2_trace <- vapply(1:8, function(k) {
  f <- suppressWarnings(glmmPQL(y ~ resp, random = ~ 1 | sample_id,
                                family = poisson, data = toy, niter = k,
                                verbose = FALSE))
  as.numeric(nlme::VarCorr(f)[1, 1])
}, numeric(1))
ggplot(data.frame(iteration = seq_along(tau2_trace), tau2 = tau2_trace),
      aes(iteration, tau2)) +
  geom_hline(yintercept = true_tau2, linetype = 2, colour = spide_pal$ref) +
  geom_line(colour = spide_pal$accent) + geom_point(colour = spide_pal$accent) +
  labs(x = "outer iteration", y = expression(hat(tau)^2),
       title = "One gene's variance-component estimate")
```

It converges in a handful of iterations and does not land on the true value:
with $S = 12$ patients, one gene's $\hat\tau^2$ has real sampling noise. That
is why `spiDE` pools the numerator across every gene; on the 300-gene null
from earlier the pooled estimate sits close to the simulated $0.49$:

```{r tau2_pooled}
mixed <- fit_mode("intercept")
mixed$fit$tau2[["SampleInt"]]
```

**Shrinkage** is the other signature of a random effect. Compare each
sample's BLUP with the fully unpooled per-sample estimate from the saturated
model `y ~ sample_id - 1`:

```{r shrinkage, fig.width = 6.5, fig.height = 4, fig.cap = "Each sample's unpooled deviation and its BLUP, connected by an arrow. Every arrow points toward zero, and the least reliable estimate is pulled in the most."}
naive_u <- coef(glm(y ~ sample_id - 1, family = poisson, data = toy))
naive_u <- naive_u - mean(naive_u)
blup_u <- nlme::ranef(fit_pql)[[1]]
ord <- order(naive_u)
shrink_df <- data.frame(sample = factor(seq_len(S)[ord], levels = seq_len(S)[ord]),
                        naive = naive_u[ord], blup = blup_u[ord])
ggplot(shrink_df) +
  geom_segment(aes(x = sample, xend = sample, y = naive, yend = blup),
              arrow = arrow(length = unit(0.12, "cm")), colour = "grey55") +
  geom_hline(yintercept = 0, linetype = 2, colour = spide_pal$ref) +
  geom_point(aes(sample, naive, colour = "unpooled"), size = 2.2) +
  geom_point(aes(sample, blup, colour = "BLUP"), size = 2.2) +
  scale_colour_manual(
    values = c("unpooled" = unname(spide_pal$method[["fixed"]]),
              "BLUP" = unname(spide_pal$method[["intercept"]])), name = NULL) +
  labs(x = "sample (sorted by unpooled deviation)",
       y = "deviation from population mean",
       title = sprintf("var(unpooled) = %.2f, var(BLUP) = %.2f",
                       var(naive_u), var(blup_u))) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

Shrinkage is strongest where a sample's own data is least reliable relative
to $\tau^2$ — the toy uses few cells and a low mean so that it is visible. A
well-measured gene gets little help from the random effect and only a modest
standard-error correction; a sparse gene, whose per-sample counts are noisy,
is where the correction does most of its work.

# Converging each gene

## Why the shared fit is not at every gene's optimum

`SpaNorm::fitNB` fits every gene in one IRLS loop. To make a
whole-transcriptome fit affordable it shares one gene-averaged cell-weight
vector across genes, judges step-halving and convergence on the *aggregate*
log-likelihood, and clamps coefficients across genes. For most genes that is
indistinguishable from the gene's own optimum. For a bright,
cell-type-restricted gene, whose working weights look nothing like the
average, the aggregate criterion is met before that gene's own score reaches
zero, and the dispersion estimated at that point is off. `converge = TRUE`
(the default) finishes the job per gene, and `polishSpiDE()` applies the same
stage to an existing fit — made with `converge = FALSE`, or saved by an older
version — without refitting. The two routes share one implementation.

## Damped Newton on the gene's own likelihood

For gene $g$, with the design $\tilde W$ and penalty $\Lambda$ fixed, the
objective is its own penalised negative binomial log-likelihood,

$$
\ell_g(\theta;\psi_g) \;=\; \sum_c \log \mathrm{NB}\!\left(Y_{c,g};\
\mu_{c,g} = e^{\tilde w_c^\top\theta},\ \psi_g\right)
\;-\; \tfrac12\,\theta^\top\Lambda\,\theta ,
$$

whose score and (expected) information are

$$
s(\theta) = \tilde W^\top \frac{Y_g - \mu_g}{1 + \psi_g\mu_g} - \Lambda\theta,
\qquad
I(\theta) = \tilde W^\top \mathrm{diag}\!\left(\frac{\mu_g}{1+\psi_g\mu_g}\right)\tilde W + \Lambda .
$$

Each Newton step solves $I(\theta)\,\delta = s(\theta)$ and halves the step
until $\ell_g$ increases; the iteration stops when the gain falls below `tol`
relative to $|\ell_g|$. A gene whose starting fit is degenerate — a fitted
mean below $e^{-10}$ somewhere — restarts from the per-cell-type log means,
because Newton *from* a degenerate point diverges. Each polished gene depends
only on its own counts, so the stage is blocked over genes and parallelised
(`block.size`, `BPPARAM`), exactly as inference is.

## The dispersion at the converged mean

`fitNB` estimates $\psi_g$ with `edgeR`'s empirical-Bayes moderation, which
shrinks each gene's dispersion toward a cross-gene trend. The polish keeps
that value (`polish.psi = "moderated"`, the default) and converges only the
coefficients under it. The alternative, `polish.psi = "profile"`, replaces it
with the gene's own **profile maximum-likelihood** value: hold the
coefficients at their converged value $\hat\theta$, so that $\hat\mu_{c,g}$ is
fixed, view the log-likelihood as a function of the dispersion alone,

$$
\ell_g(\psi \mid \hat\mu) \;=\; \sum_c \Big[
  \log\Gamma\!\big(Y_{c,g} + \tfrac1\psi\big) - \log\Gamma\!\big(\tfrac1\psi\big)
  - \log\Gamma(Y_{c,g}+1)
  + Y_{c,g}\log\frac{\psi\hat\mu_{c,g}}{1+\psi\hat\mu_{c,g}}
  - \frac1\psi\log\big(1+\psi\hat\mu_{c,g}\big)\Big] ,
$$

maximise it over $\log\psi$ on a bounded interval (`psi.range`, by default
$[10^{-3}, 10^{3}]$) with a one-dimensional search (`stats::optimize`),
re-polish the coefficients at the new $\psi_g$, and repeat the pair once, so
the returned $(\hat\theta, \hat\psi_g)$ is a joint optimum to the stated
tolerance. A gene that carries no information about overdispersion — all
zero, or effectively Poisson — has a profile likelihood that is monotone in
$\psi$, so the search returns a *bound*, not an estimate; such a gene keeps the
moderated value and is flagged `psi_bound` in `@polish`.

Why the moderated value is the default is a measurement, not a preference.
The profile step is the expensive part of the stage (a bounded search with
two re-polishes per gene), and on both the synthetic null and the real
cohort the two rules are indistinguishable: the same type-I error at every
sample size to the third decimal, the same triplets called. What the profile
value *did* do in an earlier release was sharpen a planted effect whose
moderated dispersion was far from its own optimum — but that sharpening
came through the standard error, and the standard error is now scaled by
the quasi-likelihood dispersion (next section), which sees the residual
scatter directly and does not depend on which $\psi_g$ the mean was
converged under. Either way, $\psi_g$ is not itself the scale of the
standard error.

The polish also converges on the **unclamped** likelihood: `fitNB`'s
winsorisation is a robustness device for a fit that shares one weight vector
across genes; a per-gene converged fit does not need it, and inference
evaluates the same unclamped mean the polish converged on. With
`converge = TRUE` the `winsor` argument therefore only sets the Newton
starting point. The negative binomial likelihood is undefined off the
integers, so the stage **requires integer counts** and refuses a
back-transformed assay rather than silently collapsing every dispersion to a
bound.

## Absorbing the nested block: the Schur complement

The nested block adds one column per non-empty (sample, cell type) group —
several hundred on a cohort — and a naive Newton step would factor an
information matrix of that size for every gene at every iteration. It does
not have to. Partition the design into the dense columns $X$ and the group
indicators $Z$, so that the information is

$$
I(\theta) \;=\;
\begin{pmatrix} A & B \\ B^\top & C \end{pmatrix},
\qquad
A = X^\top D X + \Lambda_X,\quad
B = X^\top D Z,\quad
C = Z^\top D Z + \Lambda_Z ,
$$

with $D$ the working weights. Because **every cell belongs to exactly one
group**, $Z^\top D Z$ is diagonal — entry $k$ is the total working weight of
group $k$ — and so is $C$ once the penalty is added. The Newton step for the
dense columns is then the $p \times p$ system

$$
S\,\delta_X = s_X - B\,C^{-1} s_Z,
\qquad
S = A - B\,C^{-1}B^\top ,
$$

and the group steps follow by back-substitution,
$\delta_Z = C^{-1}\big(s_Z - B^\top\delta_X\big)$. $S$ is the **Schur
complement** of $C$. Forming it costs one dense gram $X^\top D X$ plus group
sums of the weighted columns, whatever the number of groups, which is what
makes the per-gene polish affordable on a cohort.

The Schur complement is the Frisch–Waugh–Lovell theorem in matrix form:
$S$ is precisely the information of the dense columns *after the group
indicators have been partialled out* — the centring described earlier, done
by algebra rather than by subtracting means. `.newtonSolver()` verifies the
partition (each row of $Z$ sums to one) before relying on it, and
`.blockedInference()` uses the same absorption to form each gene's covariance.

## What the stage reports

`@polish` is a per-gene table: `iterations`, `psi_fitnb` (the dispersion
before polishing), `restarted`, `capped`, `singular`, `psi_bound` and
`polished`. `polished = FALSE` marks a gene the Newton could not run on at
all, which keeps the multi-gene fit's estimate rather than substituting a
worse one. Both flags are expected in small numbers; a large count is worth
investigating before trusting that bandwidth. On bright genes the converged
standard errors are *larger* than the unconverged ones — the unconverged
dispersion, estimated off the optimum, was hiding inflation behind deflation
— and the larger number is the honest one.

# Inference

## Covariance and dispersion

With the random effects and the converged fit in place, the Wald statistic
for a tested coefficient uses three things.

1. **The full penalised covariance,**
   $\widehat{\mathrm{Var}}(\hat\theta) = \varphi_g\,[(\tilde W^\top D_g \tilde W + \Lambda)^{-1}]$,
   with the tested block extracted. Because $Z$ absorbs the within-patient
   correlation, this block is on a between-patient scale — larger than the
   fixed-effects standard error, which is the point.
2. **The quasi-likelihood dispersion** (`dispersion = "ql"`, the default).
   The unit deviance $d_{c,g}$ of each observation is rescaled by the first
   two moments it has under the fitted NB, matched to a scaled chi-square
   (edgeR v4's construction: if $d \sim c\,\chi^2_\nu$ then
   $w_0 = 2E[d]/\mathrm{Var}[d]$ rescales the deviance and
   $w_1 = 2E[d]^2/\mathrm{Var}[d]$ is the observation's effective degrees of
   freedom, both obtained by direct summation over the pmf), and
   $$
   \varphi_g = \frac{\sum_c w_{0,c,g}\, d_{c,g}}{(1 - p/C)\sum_c w_{1,c,g}}
   $$
   is then moderated across genes with `limma::squeezeVar()` against the
   gene's mean count. At small counts $E[d]$ is nowhere near 1 — at a mean of
   0.1 with dispersion 5 each observation carries about half a degree of
   freedom — which is why the raw deviance over $C - p$ is unusable there
   and why this construction, rather than the Pearson one, holds the nominal
   level at four samples (calibration vignette). The moments depend only on
   $(\mu, \psi)$, so they are evaluated once on a shared table and
   interpolated; the rest is elementwise over genes and cells, on the CPU
   or the accelerator (`SpaNorm::qlDispersion()`). The **Pearson working
   dispersion**,
   $\varphi_g = \frac{1}{C - p}\sum_c (Y_{c,g} - \mu_{c,g})^2 / (\mu_{c,g} + \psi_g\mu_{c,g}^2)$,
   the device `summary(glm(..., family = quasipoisson))` reports, remains
   available as `dispersion = "pearson"`; it rejects at 0.09 against a
   nominal 0.05 with four samples and fades to 0.068 at thirty.
3. **A $t$ reference distribution** with a per-coefficient degrees of
   freedom, next.

Every per-gene matrix inversion is guarded: a gene whose information matrix
is singular drops out as `NA` rather than aborting the run.

## Reference degrees of freedom

The flat response $\eta_{g,t}$ is a pure between-patient contrast, so its
reference df is close to $S - 2$, the bookkeeping of a two-sample $t$-test on
patients. The niche slopes $\beta_{g,t,t'}$ also draw on *within*-patient
variation in the niche covariate, so they carry a larger effective sample
size, and one $S - 2$ cannot be right for both. `df.method = "satterthwaite"`
(the default) gives each tested coefficient its own df,

$$
\nu_j \;=\; \frac{2\,\big(\widehat{\mathrm{Var}}\,\hat\theta_j\big)^2}
                 {\mathrm{Var}\!\big(\widehat{\mathrm{Var}}\,\hat\theta_j\big)},
$$

with the denominator propagating the uncertainty of the variance parameters
$(\varphi, \tau^2_{\text{sample}}, \tau^2_{\text{sample}\times\text{type}},
\tau^2_{\text{slope}})$ through the standard error, from the reduced-form
REML information of the working model — a small matrix built from the
$p \times p$ penalised inverse, never an $n \times n$ one. The ratio is
invariant to $\varphi_g$, so one df vector per bandwidth serves every gene.
`df.method = "between"` uses the scalar $S - 2$ for every coefficient and is
severely conservative for the slopes when patients are few.

```{r satterthwaite, message = FALSE}
fs <- fitSpiDE(spe, "condition", sigma = 30, random = "intercept",
               df.method = "satterthwaite", verbose = FALSE)
ff <- fits(fs)[[1]]
ct <- as.character(ff$covtype)
resp  <- ff$coefmap$covariate[ct == "ResponseCellType"]
niche <- ff$coefmap$covariate[ct == "ResponseNiche"]
c(ResponseCellType.median = round(median(ff$df[resp]), 1),
  ResponseNiche.median = round(median(ff$df[niche]), 1))
```

The flat-response df sits near $S - 2 = 8$ and the slope df is much larger.
`@df` is therefore a **named vector** aligned to the columns of `@t_stat`;
index the column you are testing rather than passing the vector to `pt()`.

The null calibration of the corrected fit, against the fixed-effects one from
earlier:

```{r mixed}
c(fixed = mean(fixed$p < 0.05), mixed = mean(mixed$p < 0.05))
```

```{r qq, fig.small = TRUE, fig.cap = "Null response-effect p-values: the fixed-effects fit against the mixed fit."}
qq_points <- function(p, mode) {
  data.frame(mode = mode, expected = sort(-log10(ppoints(length(p)))),
             observed = sort(-log10(p)))
}
qq_df <- rbind(qq_points(fixed$p, "fixed"), qq_points(mixed$p, "mixed"))
ggplot(qq_df, aes(expected, observed, colour = mode)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = spide_pal$ref) +
  geom_point(size = 1.5) +
  scale_colour_manual(values = unname(spide_pal$method[c("fixed", "intercept")]),
                      name = NULL) +
  labs(x = expression(-log[10]("expected p")), y = expression(-log[10]("observed p")))
```

## From coefficients to results

A gene has one tested slope per (index type, niche type), and a gene's niche
columns are correlated because niche densities are spatially smooth. Three
steps turn the per-coefficient statistics into the `results()` table.

**Within a gene, across niches.** The two-sided p-values of a gene's slopes
are combined by the **Cauchy combination** (`combine = "cauchy"`, the
default): $T = \sum_j w_j \tan\{(\tfrac12 - p_j)\pi\}$, whose null
distribution is Cauchy whatever the correlation between the $p_j$, so no
correlation matrix has to be estimated. It is computed for the gene as a whole
and per index cell type. `combine = "brown"` is the alternative: Brown's
extension of Fisher's method, which consumes the coefficient correlation
matrix and combines *one-sided* p-values, separately for the up and down
directions. The sidedness differs by necessity — the tangent transform
diverges as a one-sided $p \to 1$, so under Cauchy a gene up in one niche and
down in another would cancel, whereas Brown's $-2\log p$ is bounded and safe
one-sided. Both populate the same slots, so everything downstream is
combiner-agnostic.

**Across bandwidths.** The gene-level and per-index combined p-values from
each bandwidth are combined by a second Cauchy combination, weighted by the
gene's relative log-likelihood across bandwidths, $w_\sigma \propto
\exp(\ell_{g,\sigma} - \max_\sigma \ell_{g,\sigma})$ with small weights
thresholded, so the bandwidth that fits a gene best carries the most weight.

**The hierarchical FDR.** A three-level nested Benjamini–Hochberg cascade:
gene level; then index cell type within gene; then (index, niche) within
those — the outer levels gated at half the nominal rate per direction and
merged, the innermost at the nominal rate. What survives is a row keyed by
(gene, index type, niche type) with the best bandwidth, its direction and its
three FDR values.

## Niche-only mode

With `condition = NULL` the design drops the condition terms *and* the bare
niche main effects,

```
~ 0 + <covariates> + CellType + CellType:(niches)
```

and the `CellType:niche` slopes — nuisance in the conditioned model — become
the tested quantity, cell-means coded. The main effects must go: a niche's
main effect equals the sum of its per-type interactions, so keeping it would
alias one interaction per niche away and leave that cell type untestable.
Everything downstream is shared. The random effects apply as before; note
that the tested slope is now a within-sample contrast on a spatially
autocorrelated covariate, which spiDE does not model, so this mode is mildly
anti-conservative and borderline calls deserve corresponding caution.

# Reading a fit

A fit with `random != "none"` carries these slots on each `SpiDEFit`
(`fits(res)[[1]]`, or `$`):

| slot | what it holds | how to read it |
|---|---|---|
| `@tau2` | the variance components, named `SampleInt`, `SampleCellTypeInt` (default `re.celltype = TRUE`) and `SampleSlope` (`random = "slope"`) | on the log-mean scale; index by name, never by position |
| `@df` | the reference df: a named per-coefficient vector (`satterthwaite`) or a scalar $S-2$ (`between`) | index the column you test; small values mean few patients and a wide-tailed $t$, which is expected |
| `@penalty` | the per-column ridge penalty used, $0$ on fixed columns and $1/\hat\tau^2_k$ on random ones | a very small penalty on a block means that source of between-patient variation converged large |
| `@polish` | the per-gene convergence diagnostics | see *What the stage reports* |

```{r diagnostics}
f <- fits(fs)[[1]]
f$tau2
colSums(f$polish[, c("restarted", "capped", "singular", "psi_bound")])
```

Three caveats govern use.

- **Sample-level covariates** (Age, Sex, Stage) are collinear with the
  per-sample random intercept and are rejected under `random != "none"`; the
  intercept subsumes them. Adjust for cell-level covariates only — and do
  adjust for per-cell library size, which the cell-type intercepts cannot
  carry.
- **Few samples.** The variance components and the reference df are only as
  good as the between-sample information; a warning is issued when samples
  are very few, and the trade-offs by sample size are in the calibration
  vignette.
- **The mixed fit is not "fixed with extras".** Even when $\hat\tau^2 \to 0$ it
  does not reproduce `random = "none"`: it also replaces the sub-design
  covariance and the NB dispersion scaling with the full penalised covariance
  and the Pearson dispersion, so its standard errors are the correctly scaled
  ones and `"none"`'s apparent extra power is partly spurious.

# The patient-level question

What the nested intercept removes from the niche slope is a real association
in the data. "Do patients whose tumour compartment is fibroblast-rich express
this gene differently *in* tumour cells?" is a legitimate question with one
observation per patient, and `compositionTest()` asks it there: per (sample,
index type), a pseudobulk log-CPM profile of the index cells and the mean
niche density around them; then, per (index, niche) pair, a `limma` moderated
$t$ across samples for the pooled association (`"niche"`) and for its
difference between conditions (`"condition:niche"`, the patient-level
counterpart of $\beta$), with patient-level covariates allowed — exactly the
ones `fitSpiDE()` rejects. On a cohort it typically reports that neighbour
types' own markers rise in the index cells' pseudobulk with the neighbours'
density, uniformly across conditions: composition, plausibly with a
segmentation-spillover component, and the reason the niche test keeps it out.
Report the two results side by side, and never one as the other.

# Choosing the estimator

`fitSpiDE(random = "intercept")` with the defaults is the estimator for
discovery: it fits all cells jointly, so thin cell types borrow strength, and
its hierarchical FDR is built for the full hypothesis space. `random = "none"`
is kept for reproducing older results only. `twoStageSpiDE()` estimates each
patient's niche slopes first and contrasts them across patients, removing
pseudo-replication by construction; it is the estimator for a pre-specified
hypothesis on well-populated index types, and it is documented in
`vignette("spiDE-twostage")`. The measured operating characteristics of every
choice named in this vignette are collected in `vignette("spiDE-calibration")`.

# Session info

```{r sessioninfo}
sessionInfo()
```
