---
title: "Calibration: reading lambda, and what the benchmarks measured"
author: "Dharmesh D. Bhuva"
package: spiDE
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Calibration: reading lambda, and what the benchmarks measured}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, collapse = TRUE, comment = "#>", message = FALSE,
  warning = FALSE, fig.width = 7, fig.height = 4.2, dpi = 150, fig.retina = 1
)
set.seed(1)
```

```{r libs}
library(spiDE)
library(ggplot2)
source(system.file("spide_palette.R", package = "spiDE"))
```

# Overview

This vignette has two halves. The first is a diagnostic: how to read the
genomic inflation factor $\lambda$ from your own fit, per cell type, and what
to do about it. The second is the evidence: every number this package quotes
about its defaults — null calibration, power, and what was measured on a real
cohort — collected in one place with pointers to the reports that carry the
full tables (https://bhuvalab.github.io/spiDE-research/). The model those
numbers are about is in `vignette("spiDE-model")`.

# Reading lambda

A `spiDE` fit tests thousands of genes against hundreds of
`celltype:condition:niche` coefficients, and those coefficients are **not
equally well estimated**. A tumour compartment with tens of thousands of cells
spread over every patient supports a very different quality of inference from
a dendritic-cell compartment with a few hundred cells concentrated in a
handful of patients — yet both are reported in the same table, against the
same FDR threshold. $\lambda$ detects that.

- $\lambda$ measures whether your test statistics follow the null
  distribution you assumed. It is a property of the **test**, not of the
  biology.
- In `spiDE` it varies *within a single fit*, across cell types, because cell
  types differ in how many patients contribute cells.
- $\lambda < 1$ costs **power, not validity**. A null result for a poorly
  calibrated cell type is weak evidence of absence.
- The right response is to restrict which coefficients you test, **not** to
  rescale the statistics and **not** to change the model.

## What lambda is

For a one-degree-of-freedom test, the squared statistic $t^2$ is a $\chi^2_1$
variate under the null. The genomic inflation factor compares the observed
median with the null's:

$$\lambda = \frac{\mathrm{median}(t^2)}{\mathrm{median}(\chi^2_1)}
          = \frac{\mathrm{median}(t^2)}{0.4549}$$

The median is the important choice: a minority of true positives, however
large, barely moves it, so $\lambda$ reads the bulk of the distribution, which
is overwhelmingly nulls (Devlin and Roeder, 1999).

| $\lambda$ | reading |
|---|---|
| $= 1$ | the observed null matches the theoretical one; p-values mean what they say |
| $> 1$ | statistics too large: confounding or anti-conservatism, **or** genuinely widespread signal |
| $< 1$ | statistics too small: the test is conservative |

```{r lambda-fn}
lambda <- function(t) stats::median(t^2, na.rm = TRUE) / stats::qchisq(0.5, 1)
```

## Lambda below 1 does not mean "no signal"

Under a pure null with a calibrated test, $\lambda = 1$ exactly. Absence of
biology gives 1, not less than 1:

```{r pure-null}
z <- matrix(rnorm(200 * 50), 200, 50)
round(lambda(z), 3)
```

So $\lambda = 0.68$ is never a statement that a cell type is quiescent. It
says the reference distribution is wrong for the statistic's sampling
behaviour — here, that the test is conservative. Wald tests are asymptotic:
when few patients contribute cells, the variance is biased upward and $|t|$
comes out systematically too small. Because $|t|$ scales with
$\sqrt\lambda$, a deflated test needs a proportionally larger effect to clear
the same bar:

```{r cost-table}
t_true <- stats::qnorm(1 - 0.001 / 2)          # |t| = 3.29 for p = 0.001
lam <- c(1.00, 0.90, 0.80, 0.70, 0.60, 0.50)
cost <- data.frame(lambda = lam,
                   t_observed = round(t_true * sqrt(lam), 2),
                   p_reported = signif(2 * stats::pnorm(-abs(t_true * sqrt(lam))), 2))
cost$times_worse <- round(cost$p_reported / 0.001, 1)
cost
```

A cell type at $\lambda = 0.5$ reports a genuine $p = 0.001$ effect as about
$p = 0.02$; after a genome-wide correction such a compartment is effectively
unable to produce a discovery.

## What drives it

The `condition` contrast is between patients, and the random intercept
absorbs patient-level shifts, so what buys precision is the number of
*patients* contributing usable cells of that type, not the raw cell count.
$\lambda$ can depart from 1 in either direction, and the direction says which
failure dominates:

- **Many cells over few patients** — pseudo-replication; statistics too
  large, $\lambda > 1$. This is what the mixed-effects correction fixes.
- **Few cells and few patients** — small-sample Wald conservatism;
  statistics too small, $\lambda < 1$.

Both resolve toward 1 as information accumulates. A third driver is not a
variance at all: a between-patient composition association loading onto the
niche slopes when the design has no (sample × cell type) intercept. It
inflates $\lambda$ in exactly the pattern a real programme shift would —
concentrated in the compartments with the most cells, largest for the
brightest genes — and `re.celltype = TRUE` (the default) removes it; refit
before interpreting an inflated compartment from an older fit. The measured
form is under *The real cohort* below. A small measured cost of the defaults
on a clean null also means a $\lambda$ a few percent above 1 is expected from
the method and is not evidence of a data problem.

## Diagnosing your own fit

Read $\lambda$ **per index cell type**, never as one number for the fit — a
global value is an abundance-weighted mixture that hides the heterogeneity.

```{r per-celltype, eval = FALSE}
lambda_by_celltype <- function(fit, type = "ResponseNiche") {
  cm <- as.data.frame(fit@coefmap)
  # `covtype` spans every design column, `t_stat` only the tested ones:
  # always match on column names
  cols <- intersect(colnames(fit@t_stat), cm$covariate[cm$type == type])
  idx  <- cm$index[match(cols, cm$covariate)]
  out  <- lapply(split(cols, idx), function(cc) {
    data.frame(n_col = length(cc), lambda = lambda(fit@t_stat[, cc, drop = FALSE]))
  })
  cbind(index = names(out), do.call(rbind, out))
}
# lambda_by_celltype(fits(res)[[1]])
```

Pair it with the number of patients contributing cells of each type, the
quantity that predicts it:

```{r patient-spread, eval = FALSE}
tab <- table(spe$cell_type, spe$sample_id)
data.frame(index = rownames(tab), n_cells = rowSums(tab),
           n_patients = rowSums(tab > 0), n_pat_ge10 = rowSums(tab >= 10))
```

## What to do about it

**Do not rescale by lambda.** Classical genomic control divides every
statistic by $\lambda$; it assumes one inflation factor for all tests, whereas
here $\lambda$ varies between cell types within a fit. A global correction
deflates the calibrated compartments and leaves the sparse ones
under-powered. The heterogeneity is the finding.

**Do not reach for a bigger model.** A random slope, a different penalty or a
refit will not move $\lambda$ when the cause is small-sample conservatism.
Two checks before blaming the model: a fixed-effects fit should be *worse*
calibrated than the mixed fit, not better; and the dispersion multiplier at
inference should have a median below 1, i.e. be shrinking standard errors.
Conservatism at small $n$ is expected behaviour for a Wald test on a sparse
negative binomial fit.

**Restrict what you test.** Stop testing coefficients that cannot support
inference, filtering on patients contributing cells rather than on total
cells. The under-powered tail stops consuming the multiple-testing budget, and
the output stops presenting uninformative nulls as evidence. Report the
threshold, and treat a null from an excluded compartment as weak evidence of
absence.

**When lambda exceeds 1.** Because it is a median, a handful of strong hits
cannot move it; a large fraction of genes must shift. Residual confounding
affects null genes indiscriminately; a genuine programme-level shift
concentrates in one compartment and one direction and reproduces in an
independent contrast. Where one cell type shows $\lambda > 1$ and the rest of
the fit sits at or below 1, a coordinated shift in that compartment is the
more parsimonious reading — and it argues for interpreting at the level of
the programme rather than gene by gene.

# What the benchmarks measured

Each paragraph is one measured fact with the number that matters, in the
order the defaults are introduced in the model vignette. The simulation study
uses a realistic simulator (Gamma gene means, a mean–variance dispersion
trend, library sizes of about 1,500–2,000 counts per cell, configurable
spatial layouts), 40 replicates per design point, sample sizes $S$ from 4 to
30 and 300 cells per sample unless stated. Full tables: the *Simulation study*
report on the benchmark site. All figures are for `random = "intercept"` with
`df.method = "satterthwaite"` unless stated.

## The random intercept

On a null with per-sample intercepts and no response effect,
`random = "none"` rejects the response contrast at about **0.71** against a
nominal 0.05, worse as cell counts become imbalanced across samples; the
random intercept holds about **0.04**, matching a calibrated pseudobulk
reference. That is why `"none"` is not the default and should not be used for
inference.

## The reference degrees of freedom

`df.method = "between"` tests every coefficient against the scalar $S - 2$
and is severely over-conservative for the niche slopes when patients are few:
null type-I about **0.001** at $S = 4$ against 0.05, with near-zero power.
`"satterthwaite"` holds type-I in **0.042–0.065** over the whole sampled range
and gains about **0.10** in recall, with a mild liberal drift at larger $S$
(worst 0.065). A Kenward–Roger correction of the covariance is the indicated
refinement.

## The small-sample regime

`fitSpiDE(random = "intercept")` is the most powerful calibrated estimator
once samples are plentiful — recall **0.66** at $S = 30$ against 0.45 for the
best two-stage arm — but it fails at small $S$ in a way the null table does
not show: realised FDP **0.35** at $S = 4$ against a nominal 0.05, recovering
to 0.04 by $S = 16$. Any sub-analysis that thins the patient count re-enters
this regime.

## The 0.99.17 defaults: nested intercept and per-gene convergence

The two defaults were adopted on real-cohort evidence (below). Measured as
switch arms on the simulation study, which places niche cells by the same
potential in every sample and therefore plants **no** composition effect, they
can only cost there, and they do:

| null type-I, mean over $S \ge 10$ | at $S = 4$ |
|---|---|
| design term alone: 0.058 | 0.036–0.049 |
| convergence only: 0.066 | 0.086–0.095 |
| nested block only: 0.068 | 0.037–0.043 |
| both (the defaults): 0.071 | 0.085–0.096 |

(Per-cell standard error about 0.0016.) The two switches act at different
sample sizes. Convergence's cost is a small-$S$ effect — the per-gene
dispersion gives up cross-gene moderation, which matters most when data are
few — and fades onto the design-term arm by $S = 24$. The nested block is
harmless at $S = 4$ and holds a constant $+0.010$ from $S = 16$, which points
at its reference df or the estimation of its variance component. The pair
gains recall (**0.402** against 0.361 at $S = 30$) with slightly better FDP
from $S = 10$ and a worse one at $S = 4$ (0.59 against 0.35). None of this is
a reason to revert: the defaults were chosen for a confound real cohorts
carry, because patients differ in composition. The two candidate refinements
— a moderated dispersion at the converged mean, and the nested block's
reference df — are what this measurement will judge.

## The `CellType:condition` term, and a design that looks more powerful

The design spiDE shipped before 0.99.7 had no `CellType:condition` column,
and on the simulation study it reports far more niche calls: recall **0.36**
against 0.17 at FDR 0.05 on the same dataset, concentrated in marker genes.
That is not a better test. One power dataset ($S = 24$, $\beta = 2$) was fit
under four package builds including one that differs from that design *only*
by the added column; the build makes no difference (slopes correlate at
1.000), the column is the whole gap. Without it the niche slope exceeds the
term-restored one by **0.25×** the flat-response coefficient, gene by gene
($r = 0.88$ on planted genes, 0.92 on null genes) — the mean of the planted
effect counted as niche dependence — which inflates the estimate by only
about 10%; the $t$ is 1.5–1.6× larger because the standard error is
**0.67–0.72×**, a regression through the origin. On flat-shift genes with no
niche effect the same channel produces fabricated niche calls; the recall
saturates at about 0.5 from $\beta = 2$; and on the real cohort the
term-restored design recovered **41% more** triplets. Two-sided Cauchy
combination gave more discoveries than one-sided Brown in every build, so
sidedness is not involved. The recall of a design is comparable across
designs only when they estimate the same quantity.

## The combiner

Cauchy (two-sided) matches or beats Brown's method on power while holding
type-I error under correlation without estimating the correlation matrix;
report: *Combining niche p-values*.

## The two-stage estimator

On the simulation study its `"ols"` stage-1 path is the best-calibrated arm
(null type-I **0.043–0.054**, flat in $S$), with raw power **0.451** at
$S = 30$ against 0.293 (`"nb"`) and 0.274 (`"spanorm"`, the most liberal at
0.055–0.071 and the least powerful); FDP 0.032 against 0.173 for `"spanorm"`
at $S = 16$. Pooling the dispersion per sample (`pool.psi = TRUE`) is
consistently better calibrated (0.067 against 0.073 paired over 30 datasets).
On the real cohort the picture reverses for thin index types: its
$\mathrm{sd}(t)$ tracks cells per (patient, index) subset at $r = -0.84$ —
Tumor at 388 cells per subset is calibrated (1.05), Mast at 38 cells is not
(1.65) — and the mixed GLM, which fits all cells jointly, is the better
calibrated estimator across all index types (1.04 against 1.36). Dropout
confounded with condition also fires there in B cell, DC and Monocyte.
Report: *Two-stage estimation*.

## The gene-set layer

`spiGSEA(test = "competitive")` is calibrated; the self-contained form is
not (it called 20.6 of 208 null sets per replicate, all false) and is kept
only to reproduce older scripts. Report: *Gene-set inference*.

## The real cohort

The sharpest test in this package's history, and the one to repeat before any
claim of niche-dependent DE: a **niche-shuffle null** on a 55-patient CosMx
cohort, permuting the rows of the niche matrix within (sample, cell type) so
that the true slope is zero by construction, in `free` (random rows) and
`block` (toroidal shift, preserving spatial smoothness) forms. Report: *The
niche-shuffle null*.

- **Under the design shipped before 0.99.17, the real data were
  indistinguishable from their null**: $\mathrm{sd}(t)$ 1.100 against a
  shuffle maximum of 1.101. The heavy tail was per-gene — standardising each
  gene by its own null spread took the $p < 10^{-6}$ excess from 59.9× to
  2.6× — and confined to the brightest genes in the most populous index
  types. Twelve variance-side explanations were refuted. The cause was the
  composition confound of the model vignette: with a gene's converged fit
  held fixed, the permutation $t$ had the expected spread but a per-column
  mean up to $\pm 3$ that equalled the real-data $t$ column by column
  ($r = 0.66$); the shuffle preserves each group's mean niche density and so
  preserves the confound, which is why real and null matched.
- **With the nested intercept and convergence** the `free` null is flat at
  **0.96–0.98** in every expression band with zero $|t| > 4.89$ exceedances
  of 101,508 where the old design produced 95; the `block` null keeps a
  0.99 → 1.21 gradient in the ~100 brightest genes, the spatially smooth
  component, so calibrate against `block`. The real cohort now exceeds its
  null — 97 exceedances against a replicated null range of 5–11, with 100
  random control genes inside the range — and gives about a hundred calls at
  a per-gene empirical FDP of 0.05.
- **The count is not a set of findings.** Splitting Plasma out of the B cell
  compartment keeps the count and loses every immunoglobulin call; five
  imaging covariates keep 84 of 104 and drop the keratin-in-immune calls;
  bandwidths 10, 50 and 70 share only four triplets. One triplet survives
  every perturbation. The patient-level composition test on the same cohort
  shows neighbour types' own markers rising in the index cells' pseudobulk
  with the neighbours' density, uniformly across conditions — which is what
  the old design's immunoglobulin-in-tumour calls were.
- **Raw counts with a global library-size slope** are the best-calibrated
  substrate measured: the expression gradient of the null went from 1.207 to
  1.149 and the median null $\mathrm{sd}(t)$ from 1.055 to 1.008. A
  back-transformed assay is not counts, and cell-type-specific size factors
  buy nothing.

# Session info

```{r session}
sessionInfo()
```
