---
title: "Two-stage estimation: benchmarked against the published simulation study"
author: "Dharmesh D. Bhuva"
package: spiDE
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Two-stage estimation benchmark}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4, dpi = 150)
```

```{r libs}
library(spiDE); library(ggplot2)
source(system.file("spide_palette.R", package = "spiDE"))
bench <- function(f) {
  p <- system.file("extdata", "benchmark", f, package = "spiDE")
  if (!nzchar(p) || !file.exists(p)) return(NULL)
  readRDS(p)
}
# One canonical table per scenario -- the two-stage arm is additional ROWS in the
# published tables, selected by `method`, not a parallel file that would carry a
# stale copy of the other arms.
NULLT <- bench("null_type1.rds")
if (!is.null(NULLT)) NULLT <- NULLT[NULLT$design == "celltype-response", ]
POW <- bench("power_fdr.rds")
if (!is.null(POW)) POW <- POW[POW$sweep == "samples" & POW$fdr == 0.05 &
                              POW$design == "celltype-response" & POW$layout == "gradient" &
                              (POW$df.method %in% c("satterthwaite", "twostage")), ]
```

# What is being compared

`condition` is assigned per **patient**, so patients are the experimental units.
[fitSpiDE()] treats cells as replicates of that contrast and corrects for it with
patient random effects (`random = "none" | "intercept" | "slope"`).
[twoStageSpiDE()] instead estimates each patient's niche slope and contrasts
those slopes — 55 numbers rather than 77,000.

This benchmarks the two-stage estimator on the **same simulator, the same design
points and the same measured quantity** as the published study: type-I from the
p-value of the single `CellTypeA:conditionResponder:B` coefficient per gene, and
TPR/FDP from discoveries restricted to that A:B triplet. 10,000 genes, 300 cells
per sample, sigma 50, 40 replicates per point — 600 replicates, 2.2 min each.

The comparison is **unpaired**: the published arms seed each job by hashing its
specification, this arm uses independent seeds. At 40 replicates x 10,000 genes
the type-I estimates are precise enough that this does not affect the reading,
but small differences should not be over-interpreted.

# Null calibration

```{r null-fig, fig.height = 3.6}
if (!is.null(NULLT)) {
  d <- NULLT[NULLT$df.method %in% c("satterthwaite", "twostage") | NULLT$method == "fixed", ]
  d$method <- factor(d$method, c("fixed", "intercept", "slope", "twostage"))
  print(ggplot(d, aes(S, type1, colour = method)) +
    geom_hline(yintercept = 0.05, linetype = 2, colour = "grey40") +
    geom_line(linewidth = 0.7) + geom_point(size = 1.9) +
    facet_wrap(~ layout) + scale_colour_spide("method") +
    labs(x = "samples (S)", y = "type-I error", colour = NULL,
         title = "Null calibration against a nominal 0.05",
         subtitle = "dashed line = nominal; above it is anti-conservative"))
}
```

```{r null-tab}
if (!is.null(NULLT)) {
  d <- NULLT[NULLT$df.method %in% c("satterthwaite", "twostage") | NULLT$method == "fixed",
             c("layout", "S", "method", "type1")]
  w <- reshape(d, idvar = c("layout", "S"), timevar = "method", direction = "wide")
  names(w) <- sub("^type1\\.", "", names(w))
  knitr::kable(w[order(w$layout, w$S), ], digits = 4, row.names = FALSE,
               caption = "Type-I error, nominal 0.05, 40 replicates per point.")
}
```

**`intercept` and `slope` exceed nominal from S >= 10** (up to 0.076 on the
clustered layout), **`fixed` is drastically conservative** (~0.004-0.009), and
**two-stage is conservative everywhere** (0.017 at S = 4 rising to 0.043 at
S = 30). The trend is what the estimator's construction predicts: the
patient-level contrast has little to work with at S = 4 and approaches nominal
as patients accumulate.

# Power, and where the methods cross over

```{r pow-fig, fig.height = 3.6}
if (!is.null(POW)) {
  p <- POW; p$method <- factor(p$method, c("intercept", "slope", "twostage"))
  long <- rbind(data.frame(S = p$S, method = p$method, value = p$tpr, metric = "TPR"),
                data.frame(S = p$S, method = p$method, value = p$fdp, metric = "FDP"))
  print(ggplot(long, aes(S, value, colour = method)) +
    geom_hline(data = data.frame(metric = "FDP", y = 0.05), aes(yintercept = y),
               linetype = 2, colour = "grey40") +
    geom_line(linewidth = 0.7) + geom_point(size = 1.9) +
    facet_wrap(~ metric, scales = "free_y") + scale_colour_spide("method") +
    labs(x = "samples (S)", y = NULL, colour = NULL,
         title = "Power and realized FDP, samples sweep, FDR 0.05"))
}
```

```{r pow-tab}
if (!is.null(POW)) {
  knitr::kable(POW[order(POW$S, POW$method), c("S","method","tpr","fdp","n_disc")],
               digits = 3, row.names = FALSE,
               caption = "Samples sweep, gradient layout, FDR 0.05.")
}
```

There is a **crossover at S ~ 24**:

* **Below it, two-stage is badly underpowered.** At S = 4 it makes no
  discoveries at all where `intercept` reaches TPR 0.040 — but note `intercept`
  buys that with **FDP 0.346**, so those discoveries are mostly false.
* **At S = 24 they are comparable** (TPR 0.265 vs 0.315) with two-stage at less
  than half the FDP (0.014 vs 0.034).
* **At S = 30 two-stage dominates outright**: TPR **0.432** against 0.361, with
  FDP **0.018** against 0.033. More power *and* better error control.

# Reading this for a real study

The crossover is the practical result. Two-stage needs patients, because that is
what it treats as replicates; below roughly 20 samples it should not be used.
Above it, the honest error structure stops costing power and starts paying.

The YTMA CosMx cohort has **55 patients**, comfortably past the crossover — and
on that cohort, with condition labels permuted so every triplet is null,
`fitSpiDE(random = "intercept")` returns ~576 false calls in every replicate
(realized FDR 1.00) where two-stage returns **zero**, in 3.7 minutes against 128.

Two limits worth carrying:

* **Multiplicity is untouched.** Two-stage fixes the error structure, not the
  size of the hypothesis space, and the hierarchical cascade does not help for
  sparse per-gene signal — combining a gene's mostly-null triplets dilutes a
  single true effect to no better than Bonferroni. Restrict `index`, `niche` and
  the gene set to a pre-specified hypothesis.
* **These are simulations.** The generative model is the package's own, so it
  cannot expose a mismatch shared by simulator and method.
